---
title: "MapBiomas land-cover → fire-vegetation class reclassification"
author: "MapBiomas Argentina Fire"
date: today
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
---

## Purpose

Define the mapping from MapBiomas land-cover class codes to the fire-vegetation
classes used by the col1 RF model:

| Code | Name | Notes |
|------|------|-------|
| 0 | unburnable | water, snow, urban, rocky outcrop |
| 1 | forest | native forest, flooded forest, silviculture |
| 2 | shrubland | arbustal |
| 3 | grassland / agri | grassland, wetland, mosaic uses |

The resulting mapping goes into `utils/constants.py` as
`MB_RECLASS_FROM` and `MB_RECLASS_TO`.

## Setup

```{r setup}
library(rgee)
library(tidyverse)
library(sf)

ee_Initialize()  # requires earthengine-api Python env; run ee_install() once
```

## MapBiomas class legend

```{r legend}
# Load class legend (update path as needed)
# legend <- read_csv("../data/mapbiomas_legend.csv")
#
# Placeholder — populate when the CSV is available.
# Expected columns: class_code, class_name_es, class_name_en
```

## Proposed reclassification (from col0)

```{r col0_mapping}
# Inherited from collection-00/utils/functions.js
col0_from <- c(3, 66, 6, 12, 11, 75, 63, 21,  9, 29, 25, 24, 33, 34, 27)
col0_to   <- c(1,  2, 1,  3,  3,  3,  3,  3,  1,  0,  0,  0,  0,  0,  0)

tibble(
  mb_class  = col0_from,
  fire_class = col0_to,
  fire_name  = recode(col0_to, `0`="unburnable", `1`="forest", `2`="shrubland", `3`="grassland_agri"),
)
```

## Sample N training points per fire-class and region

```{r sample_counts}
# TODO: after step 01 assets exist, query them here to validate class distribution.
#
# For each region, load training_observations_v1 (or the CSV download from it)
# and tabulate:
#   - N burned / unburned per fire_class
#   - N per region
#   - Temporal coverage (years with observations)
```

## Decisions

*Document any changes to the col0 mapping and their rationale here.*

<!--
Example:
- Class 75 (Apicum) reassigned from grassland → unburnable because ...
-->

## Final mapping for constants.py

```{r final_mapping}
# When satisfied, copy these vectors into utils/constants.py

mb_reclass_from <- col0_from   # update if needed
mb_reclass_to   <- col0_to

cat("MB_RECLASS_FROM =", deparse(as.integer(mb_reclass_from)), "\n")
cat("MB_RECLASS_TO   =", deparse(as.integer(mb_reclass_to)), "\n")
```
