---
title: "Burn probability time-series metrics"
format:
  html:
    self-contained: true
    code-fold: true
execute:
  warning: false
  message: false
---

## Overview

Per pixel per year, the fire-detection algorithm produces an intra-annual sequence of burn probabilities at Landsat observation dates. We explore summary metrics derived from that irregular time series to identify which derived quantities best discriminate a true burn signal from noise, slow vegetation stress (drought), or a partial disturbance (deforestation).

Four synthetic signals are created over 46 equally-spaced observations (DOY 1–365). From each probability vector we compute forward differences and rolling means at multiple window lengths, then visualise all metrics together.

## Setup

```{r}
library(tidyverse)

# Single source of truth for plot styling. Bump `base_size` to scale every font
# in every plot, or tweak individual elements here to update all plots at once.
theme_burn <- function(base_size = 14) {
  theme_minimal(base_size = base_size) +
    theme(
      strip.placement   = "outside",
      strip.text.y.left = element_text(angle = 0, hjust = 1,
                                       size   = base_size - 2,
                                       margin = margin(r = 6)),
      strip.text.x      = element_text(face   = "bold",
                                       size   = base_size - 1,
                                       margin = margin(b = 6)),
      panel.spacing     = unit(0.85, "lines"),
      panel.grid.minor  = element_blank(),
      plot.title        = element_text(face   = "bold",
                                       size   = base_size + 2,
                                       margin = margin(b = 6)),
      plot.subtitle     = element_text(size   = base_size - 2,
                                       colour = "grey45",
                                       margin = margin(b = 8)),
      axis.text         = element_text(size = base_size - 2),
      axis.title        = element_text(size = base_size - 1),
      legend.position   = "top",
      legend.box        = "horizontal",
      legend.title      = element_text(size = base_size - 1),
      legend.text       = element_text(size = base_size - 2)
    )
}

theme_set(theme_burn())
```

## Synthetic signals

```{r}
n   <- 46
doy <- seq(1, 365, length.out = n)

make_signal <- function(type, peak = 1.0) {
  p <- rep(0.0, n)
  switch(type,
    fire = {
      # abrupt jump to `peak` at obs 10; plateau 3 obs; fast linear decline to 0 by obs 25
      p[10:12] <- peak
      p[12:25] <- seq(peak, 0, length.out = 14)
    },
    noise = {
      # near-zero everywhere except a single cloud-contamination spike at obs 10
      p[10] <- peak
    },
    drought = {
      # slow linear increase from obs 3 to obs 22; slow linear decline to obs 44
      p[3:22]  <- seq(0, peak, length.out = 20)
      p[22:44] <- seq(peak, 0, length.out = 23)
    },
    deforestation = {
      # abrupt jump to `peak`; plateau 3 obs; slow decline to 0 by obs 40
      p[10:12] <- peak
      p[12:40] <- seq(peak, 0, length.out = 29)
    }
  )
  p
}

# Each row defines one signal. Multiple rows per `case` are allowed and will be
# overlaid in the same column of the facet grid, distinguished by colour.
signal_specs <- tibble::tribble(
  ~case,           ~level,         ~peak,
  "fire",          NA_character_,  1.00,
  "noise",         NA_character_,  0.90,
  "drought",       NA_character_,  1.00,
  "deforestation", "0.50",         0.50,
  "deforestation", "0.75",         0.75
)
```

## Derived metrics

Let $p_t$ denote the burn probability at observation $t \in \{1, \dots, n\}$ on date $d_t$. Each quantity below is assigned to the indicated date (always **forward** — the date of the earliest observation entering the window, except $\Delta p_t$ which is dated at the second operand).

$$
\begin{aligned}
\Delta p_t                 &= p_t - p_{t-1},                                       &\quad& d_t \\[2pt]
\Sigma\Delta p^{(2)}_t     &= \Delta p_t + \Delta p_{t+1} \;=\; p_{t+1} - p_{t-1}, &\quad& d_t \\[2pt]
\Sigma\Delta p^{(3)}_t     &= \Delta p_t + \Delta p_{t+1} + \Delta p_{t+2}
                              \;=\; p_{t+2} - p_{t-1},                             &\quad& d_t \\[2pt]
\bar p^{(2)}_t             &= \tfrac{1}{2}\,(p_t + p_{t+1}),                        &\quad& d_t \\[2pt]
\bar p^{(3)}_t             &= \tfrac{1}{3}\,(p_t + p_{t+1} + p_{t+2}),              &\quad& d_t \\[2pt]
\pi^{(3)}_t                &= p_t \, p_{t+1} \, p_{t+2},                            &\quad& d_t \\[2pt]
q^{(3)}_t                  &= p_t \, p_{t+1} \;+\; p_{t+1} \, p_{t+2},              &\quad& d_t \\[2pt]
J_t                        &= \Delta p_t \;\cdot\; \min\!\bigl(p_t,\, p_{t+1},\, p_{t+2}\bigr), &\quad& d_t
\end{aligned}
$$

The diff-sums telescope to $\Sigma\Delta p^{(k)}_t = p_{t+k-1} - p_{t-1}$, i.e. wider single-step differences — by construction blind to the values between the endpoints. The product $\pi^{(3)}$ and the persistence-gated jump $J$ are both zero whenever any of their factors is zero, so a single-obs spike cannot trigger them. Products are cheap if the upstream model already exposes log-probabilities (e.g. logistic regression): $\log \pi^{(3)}_t = \log p_t + \log p_{t+1} + \log p_{t+2}$.

```{r}
# For a probability vector p at dates d, compute:
#   pdiff      : p[t] - p[t-1], assigned to forward date t
#   pdiff_sum2 : rolling sum of 2 consecutive diffs, date of first diff
#   pdiff_sum3 : rolling sum of 3 consecutive diffs, date of first diff
#   mean2      : 2-obs forward rolling mean of p, date of first obs
#   mean3      : 3-obs forward rolling mean of p, date of first obs
#   prod3      : 3-obs forward rolling product of p, date of first obs
#                — spike-killer: any zero in the window forces the window to zero
#   q3         : sum of products of consecutive pairs in a 3-obs window,
#                p[t]·p[t+1] + p[t+1]·p[t+2], date of first obs.
#                — fires when the high value is flanked by another high value;
#                  in log-space this is logsumexp(log p[t]+log p[t+1], log p[t+1]+log p[t+2])
#   jump_x_min : (p[t] - p[t-1]) · min(p[t], p[t+1], p[t+2]), date = t
#                — combined: requires both an abrupt jump AND persistence after,
#                  so it suppresses noise (no persistence) and drought (no jump) at once
compute_metrics <- function(p, d) {
  n  <- length(p)
  pd <- diff(p)          # length n-1
  m  <- length(pd)
  pd_doy <- d[-1]        # forward dates for each diff

  list(
    prob       = tibble(doy = d,              value = p),
    pdiff      = tibble(doy = pd_doy,         value = pd),
    pdiff_sum2 = tibble(doy = pd_doy[-m],     value = pd[-m] + pd[-1]),
    pdiff_sum3 = tibble(doy = pd_doy[1:(m-2)],
                        value = pd[1:(m-2)] + pd[2:(m-1)] + pd[3:m]),
    mean2      = tibble(doy = d[-n],          value = (p[-n] + p[-1]) / 2),
    mean3      = tibble(doy = d[1:(n-2)],     value = (p[1:(n-2)] + p[2:(n-1)] + p[3:n]) / 3),
    prod3      = tibble(doy = d[1:(n-2)],     value = p[1:(n-2)] * p[2:(n-1)] * p[3:n]),
    q3         = tibble(doy = d[1:(n-2)],
                        value = p[1:(n-2)] * p[2:(n-1)] + p[2:(n-1)] * p[3:n]),
    jump_x_min = tibble(
      doy   = d[2:(n-2)],
      value = (p[2:(n-2)] - p[1:(n-3)]) *
              pmin(p[2:(n-2)], p[3:(n-1)], p[4:n])
    )
  )
}

var_levels  <- c("prob", "pdiff", "pdiff_sum2", "pdiff_sum3",
                 "mean2", "mean3", "prod3", "q3", "jump_x_min")
case_levels <- c("fire", "noise", "drought", "deforestation")

df <- purrr::pmap_dfr(signal_specs, function(case, level, peak) {
  p      <- make_signal(case, peak)
  mets   <- compute_metrics(p, doy)
  sig_id <- if (is.na(level)) case else paste(case, level, sep = "_")
  purrr::map_dfr(var_levels, function(vr) {
    mets[[vr]] |> mutate(
      signal_id = sig_id,
      case      = case,
      level     = level,
      variable  = vr
    )
  })
}) |>
  mutate(
    case      = factor(case,     levels = case_levels),
    variable  = factor(variable, levels = var_levels),
    color_grp = dplyr::coalesce(level, "single")
  )
```

## Plot

```{r}
#| fig-width: 13
#| fig-height: 19

# DOY of max pdiff per case — vertical reference in the prob panel only.
# Multiple signals per case have the same transition DOY in these synthetics,
# so we keep one vline per case.
vline_df <- df |>
  filter(variable == "pdiff") |>
  group_by(signal_id, case) |>
  slice_max(value, n = 1, with_ties = FALSE) |>
  ungroup() |>
  group_by(case) |>
  slice_head(n = 1) |>
  ungroup() |>
  select(case, doy_vline = doy) |>
  mutate(variable = factor("prob", levels = var_levels))

var_labels <- c(
  prob       = "Burn prob.",
  pdiff      = "Δp  (1-step)",
  pdiff_sum2 = "Δp sum  (2)",
  pdiff_sum3 = "Δp sum  (3)",
  mean2      = "Mean  (2-obs)",
  mean3      = "Mean  (3-obs)",
  prod3      = "Product  (3-obs)",
  q3         = "Σ pair-prod  (3-obs)",
  jump_x_min = "Δp · min₃"
)

case_labels <- c(
  fire          = "Fire",
  noise         = "Cloud noise",
  drought       = "Drought",
  deforestation = "Deforestation"
)

ggplot(df, aes(x = doy, y = value, group = signal_id, colour = color_grp)) +
  geom_hline(yintercept = 0, colour = "grey80", linewidth = 0.35) +
  geom_line(linewidth = 0.65) +
  geom_point(size = 0.85) +
  geom_vline(
    data        = vline_df,
    aes(xintercept = doy_vline),
    colour      = "tomato",
    linetype    = "dashed",
    linewidth   = 0.7,
    inherit.aes = FALSE
  ) +
  facet_grid(
    variable ~ case,
    scales   = "free_y",
    switch   = "y",
    labeller = labeller(variable = var_labels, case = case_labels)
  ) +
  scale_colour_manual(
    name   = "Deforestation peak",
    values = c(single = "grey25", `0.50` = "#1f78b4", `0.75` = "#e31a1c"),
    breaks = c("0.50", "0.75")
  ) +
  scale_x_continuous(
    breaks = c(1, 91, 182, 274, 365),
    labels = c("1", "91", "182", "274", "365")
  ) +
  labs(
    x        = "Day of year",
    y        = NULL,
    title    = "Intra-annual burn probability — raw signal and derived metrics",
    subtitle = "Dashed red line (prob panel): DOY of max Δp — sharpest single-step probability increase"
  )
```

## Irregular sampling: penalising long gaps

All metrics above treat $\Delta p_t = p_t - p_{t-1}$ as instantaneous — they ignore the elapsed time between observations $t-1$ and $t$. Landsat sampling is far from regular: clouds, scene failures, and the L8+L9 combined revisit (~8 d) produce gaps that can stretch to weeks or months. Under irregular sampling, a *slow* drought sampled across a long gap can produce a $\Delta p$ that visually mimics a *fast* fire transition.

Dividing by $\Delta t$ (a rate) overcorrects — it penalises *every* non-zero gap linearly, including the 8–30 day gaps that are still perfectly trustworthy. Instead, we apply a **soft gap-penalty weight** that is essentially 1 for short gaps and saturates to 0 for long ones. A stretched-exponential (Weibull survival) shape — slow near zero, then dropping fast — has exactly that behaviour:

$$
w(\tau) \;=\; \exp\!\bigl[-(\tau/\tau_0)^{k}\bigr], \qquad \tau_0 = 80\ \text{d},\ k = 4
$$

The exponent $k = 4$ keeps the curve flat near 0; with $k = 1$ it would be plain exponential and start dropping immediately. Numerically: $w(30\,\text{d}) \approx 0.98$, $w(60\,\text{d}) \approx 0.73$, $w(90\,\text{d}) \approx 0.20$, $w(120\,\text{d}) \approx 0.006$.

The gap-weighted persistence-gated jump is

$$
J^{w}_t \;=\; (p_t - p_{t-1}) \cdot w(\tau_t) \cdot \min\!\bigl(p_t,\, p_{t+1},\, p_{t+2}\bigr),
$$

with $\tau_t = d_t - d_{t-1}$ the gap (in days) immediately preceding observation $t$.

To expose the failure mode of the time-agnostic detector we add a **sparsely-sampled drought** scenario: same underlying drought signal as before, but only ~9 observations retained across the year (mean gap ≈ 45 d). The time-agnostic $J$ inflates there because consecutive *visible* obs span large $\Delta p$; the gap-weighted $J^{w}$ does not.

```{r}
#| label: gap-weight-shape
#| fig-width: 6.5
#| fig-height: 3

gap_weight <- function(tau_days, tau0 = 80, k = 4) {
  exp(-(tau_days / tau0)^k)
}

tibble(tau = seq(0, 180, by = 1), w = gap_weight(tau)) |>
  ggplot(aes(tau, w)) +
  geom_line(linewidth = 0.9, colour = "steelblue") +
  geom_vline(xintercept = c(30, 120), colour = "grey60", linetype = "dotted") +
  annotate("text", x = 30,  y = 0.05, label = "30 d",  hjust = -0.15, size = 3, colour = "grey40") +
  annotate("text", x = 120, y = 0.05, label = "120 d", hjust = -0.15, size = 3, colour = "grey40") +
  scale_y_continuous(limits = c(0, 1), expand = c(0, 0)) +
  scale_x_continuous(breaks = seq(0, 180, by = 30)) +
  labs(
    x        = "Gap τ  (days)",
    y        = "Weight w(τ)",
    title    = "Gap-penalty weight",
    subtitle = "w(τ) = exp(−(τ / 80)⁴)"
  )
```

```{r}
# Sparse drought: keep only ~9 obs over the year (gaps ~32–57 days).
sparse_keep <- c(1, 5, 10, 16, 22:46)

build_obs <- function(case, peak, keep = seq_len(n)) {
  p_full <- make_signal(case, peak)
  list(p = p_full[keep], d = doy[keep])
}

signals2 <- list(
  fire              = list(case = "fire",           level = NA_character_,
                           obs = build_obs("fire",          1.00)),
  noise             = list(case = "noise",          level = NA_character_,
                           obs = build_obs("noise",         0.90)),
  drought           = list(case = "drought",        level = NA_character_,
                           obs = build_obs("drought",       1.00)),
  drought_sparse    = list(case = "drought_sparse", level = NA_character_,
                           obs = build_obs("drought",       1.00, sparse_keep)),
  deforestation_050 = list(case = "deforestation",  level = "0.50",
                           obs = build_obs("deforestation", 0.50)),
  deforestation_075 = list(case = "deforestation",  level = "0.75",
                           obs = build_obs("deforestation", 0.75))
)

compute_jw <- function(p, d) {
  nn  <- length(p)
  pd  <- diff(p)              # length nn - 1
  tau <- diff(d)              # gap in days, length nn - 1
  w   <- gap_weight(tau)      # length nn - 1
  fwd <- d[-1]                # forward date for each diff / each gap

  idx       <- 1:(nn - 3)     # for jump: t = 2..(nn - 2)
  jx_doy    <- fwd[idx]
  persist3  <- pmin(p[2:(nn - 2)], p[3:(nn - 1)], p[4:nn])
  j_agn     <- pd[idx]            * persist3   # time-agnostic = jump_x_min
  j_w       <- pd[idx] * w[idx]   * persist3   # gap-weighted

  list(
    prob       = tibble(doy = d,        value = p),
    weight     = tibble(doy = fwd,      value = w),
    jump_x_min = tibble(doy = jx_doy,   value = j_agn),
    jump_w     = tibble(doy = jx_doy,   value = j_w)
  )
}

var2_levels  <- c("prob", "weight", "jump_x_min", "jump_w")
case2_levels <- c("fire", "noise", "drought", "drought_sparse", "deforestation")

df2 <- purrr::imap_dfr(signals2, function(sig, sig_id) {
  mets <- compute_jw(sig$obs$p, sig$obs$d)
  purrr::map_dfr(var2_levels, function(vr) {
    mets[[vr]] |> mutate(
      year_frac = (doy - 1) / 364,
      signal_id = sig_id,
      case      = sig$case,
      level     = sig$level,
      variable  = vr
    )
  })
}) |>
  mutate(
    case      = factor(case,     levels = case2_levels),
    variable  = factor(variable, levels = var2_levels),
    color_grp = dplyr::coalesce(level, "single")
  )
```

```{r}
#| fig-width: 15
#| fig-height: 12

case2_labels <- c(
  fire           = "Fire",
  noise          = "Cloud noise",
  drought        = "Drought",
  drought_sparse = "Drought (sparse)",
  deforestation  = "Deforestation"
)

var2_labels <- c(
  prob       = "Burn prob.",
  weight     = "Gap weight  w(τ)",
  jump_x_min = "Δp · min₃   (time-agnostic)",
  jump_w     = "Δp · w(τ) · min₃   (gap-weighted)"
)

ggplot(df2, aes(x = year_frac, y = value,
                group = signal_id, colour = color_grp)) +
  geom_hline(yintercept = 0, colour = "grey80", linewidth = 0.35) +
  geom_line(linewidth = 0.65) +
  geom_point(size = 1.1) +
  facet_grid(
    variable ~ case,
    scales   = "free_y",
    switch   = "y",
    labeller = labeller(variable = var2_labels, case = case2_labels)
  ) +
  scale_colour_manual(
    name   = "Deforestation peak",
    values = c(single = "grey25", `0.50` = "#1f78b4", `0.75` = "#e31a1c"),
    breaks = c("0.50", "0.75")
  ) +
  scale_x_continuous(breaks = seq(0, 1, by = 0.25)) +
  labs(
    x        = "Fractional year",
    y        = NULL,
    title    = "Gap-weighted jump-and-persistence under irregular sampling",
    subtitle = "Row 3 (Δp·min₃) inflates for the sparsely-sampled drought — big visible Δp across long gaps.\nRow 4 weights each diff by w(τ) and the spurious signal collapses, while fire and deforestation are preserved."
  )
```

## Per-observation features for SNIC seeds and candidates

Rather than collapse everything into one composite, we expose each per-obs quantity as a feature, so the downstream SNIC seed/candidate logic can weight them independently. The feature vector at observation $t$ (in a pixel with $N^{\text{obs}}_{\text{year}}$ valid annual observations) is:

- **Diff certainty** &nbsp; $\tau^{\text{pre}}_t = d_t - d_{t-1}$ &nbsp; (days, smaller = better; soft weight $w(\tau^{\text{pre}}_t)$).
- **Persistence span** &nbsp; $\tau^{\text{post}}_t = d_{t+2} - d_t$ &nbsp; (days; larger = stronger persistence *within* a cap).
- **Adaptive persistence** — fixed-$k = 3$ when sampling is dense, falling back to $k = 2$ when too sparse:
  $$
  \operatorname{persist}^{\text{adapt}}_t \;=\;
  \begin{cases}
    \min(p_t,\, p_{t+1},\, p_{t+2}) & \text{if } \tau^{\text{post}}_t \le \tau_{\max} \\
    \min(p_t,\, p_{t+1})            & \text{otherwise}
  \end{cases}
  \qquad \tau_{\max} = 60\,\text{d}
  $$
- **Annual obs count** &nbsp; $N^{\text{obs}}_{\text{year}}$ &nbsp; (pixel-level scalar; low values down-weight everything).
- **Composite** &nbsp; $J^{w,\text{adapt}}_t = (p_t - p_{t-1}) \cdot w(\tau^{\text{pre}}_t) \cdot \operatorname{persist}^{\text{adapt}}_t$.

**Seed** (strict): $J^{w,\text{adapt}}_t \ge 0.5$ AND $\tau^{\text{pre}}_t \le 30$ d AND $\tau^{\text{post}}_t \le 60$ d AND $N^{\text{obs}}_{\text{year}} \ge 35$.

**Candidate** (permissive): $J^{w,\text{adapt}}_t \ge 0.15$ AND $\tau^{\text{pre}}_t \le 60$ d AND $N^{\text{obs}}_{\text{year}} \ge 20$.

SNIC then grows segments from seeds; candidates attach if spatially consistent. Isolated candidates (e.g., a single drought pixel flagged across a long pre-jump gap) get rejected for lack of neighbour support.

```{r}
# Adaptive persistence: 3-obs min when the window's time span ≤ tau_max,
# otherwise fall back to 2-obs min (and to p[t] alone at the trailing edge).
persist_adapt_fn <- function(p, d, tau_max = 60) {
  nn       <- length(p)
  p3       <- rep(NA_real_, nn)
  p2       <- rep(NA_real_, nn)
  tau_post <- rep(NA_real_, nn)
  if (nn >= 3) {
    p3[1:(nn - 2)]       <- pmin(p[1:(nn - 2)], p[2:(nn - 1)], p[3:nn])
    tau_post[1:(nn - 2)] <- d[3:nn] - d[1:(nn - 2)]
  }
  if (nn >= 2) p2[1:(nn - 1)] <- pmin(p[1:(nn - 1)], p[2:nn])

  out <- p
  out <- ifelse(!is.na(p2), p2, out)
  out <- ifelse(!is.na(p3) & !is.na(tau_post) & tau_post <= tau_max, p3, out)
  out
}

compute_snic_features <- function(p, d,
                                   tau_max       = 60,
                                   theta_high    = 0.5,
                                   theta_low     = 0.15,
                                   tau_pre_seed  = 30,
                                   tau_pre_cand  = 60,
                                   tau_post_seed = 60,
                                   n_min_seed    = 35,
                                   n_min_cand    = 20) {
  nn         <- length(p)
  n_obs_year <- nn

  tau_pre  <- c(NA_real_, diff(d))
  tau_post <- if (nn >= 3) c(d[3:nn] - d[1:(nn - 2)], NA_real_, NA_real_)
              else rep(NA_real_, nn)
  w_pre    <- gap_weight(tau_pre)

  persist_adapt <- persist_adapt_fn(p, d, tau_max = tau_max)
  dp            <- c(NA_real_, diff(p))
  J_w_adapt     <- dp * w_pre * persist_adapt

  is_seed <- !is.na(J_w_adapt) & J_w_adapt >= theta_high &
             !is.na(tau_pre)   & tau_pre   <= tau_pre_seed &
             !is.na(tau_post)  & tau_post  <= tau_post_seed &
             n_obs_year >= n_min_seed

  is_cand <- !is.na(J_w_adapt) & J_w_adapt >= theta_low &
             !is.na(tau_pre)   & tau_pre   <= tau_pre_cand &
             n_obs_year >= n_min_cand

  tibble(
    doy           = d,
    p             = p,
    tau_pre       = tau_pre,
    tau_post      = tau_post,
    w_pre         = w_pre,
    persist_adapt = persist_adapt,
    dp            = dp,
    J_w_adapt     = J_w_adapt,
    n_obs_year    = n_obs_year,
    flag          = factor(
      dplyr::case_when(is_seed ~ "seed", is_cand ~ "candidate", TRUE ~ "none"),
      levels = c("none", "candidate", "seed")
    )
  )
}

df3 <- purrr::imap_dfr(signals2, function(sig, sig_id) {
  feats <- compute_snic_features(sig$obs$p, sig$obs$d)
  feats |> mutate(
    year_frac = (doy - 1) / 364,
    signal_id = sig_id,
    case      = sig$case,
    level     = sig$level
  )
}) |>
  mutate(
    case      = factor(case, levels = case2_levels),
    color_grp = dplyr::coalesce(level, "single")
  )
```

```{r}
#| fig-width: 15
#| fig-height: 14

var3_levels <- c("prob", "tau_pre", "tau_post", "persist_adapt", "J_w_adapt")

var3_labels <- c(
  prob          = "Burn prob.",
  tau_pre       = "τ_pre  (d)",
  tau_post      = "τ_post  (d)",
  persist_adapt = "Persist (adaptive)",
  J_w_adapt     = "J  (w, adapt)"
)

df3_long <- df3 |>
  select(year_frac, signal_id, case, color_grp, flag,
         prob = p, tau_pre, tau_post, persist_adapt, J_w_adapt) |>
  tidyr::pivot_longer(c(prob, tau_pre, tau_post, persist_adapt, J_w_adapt),
                      names_to = "variable", values_to = "value") |>
  mutate(variable = factor(variable, levels = var3_levels))

# Threshold reference lines per row
hlines_df <- tibble::tribble(
  ~variable,    ~yintercept, ~thr_type,
  "tau_pre",    30,          "seed",
  "tau_pre",    60,          "candidate",
  "tau_post",   60,          "seed",
  "J_w_adapt",  0.5,         "seed",
  "J_w_adapt",  0.15,        "candidate"
) |>
  mutate(variable = factor(variable, levels = var3_levels))

# Flag markers — only on the prob and J_w_adapt rows
flag_markers <- df3_long |>
  filter(variable %in% c("prob", "J_w_adapt"), flag != "none")

ggplot(df3_long, aes(x = year_frac, y = value,
                     group = signal_id, colour = color_grp)) +
  geom_hline(yintercept = 0, colour = "grey80", linewidth = 0.35) +
  geom_hline(
    data        = hlines_df,
    aes(yintercept = yintercept, linetype = thr_type),
    colour      = "grey55",
    linewidth   = 0.45,
    inherit.aes = FALSE
  ) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 0.85) +
  geom_point(
    data        = flag_markers,
    aes(x = year_frac, y = value, fill = flag),
    shape       = 21,
    colour      = "grey20",
    size        = 2.6,
    stroke      = 0.5,
    inherit.aes = FALSE
  ) +
  facet_grid(
    variable ~ case,
    scales   = "free_y",
    switch   = "y",
    labeller = labeller(variable = var3_labels, case = case2_labels)
  ) +
  scale_colour_manual(
    name   = "Deforestation peak",
    values = c(single = "grey25", `0.50` = "#1f78b4", `0.75` = "#e31a1c"),
    breaks = c("0.50", "0.75")
  ) +
  scale_fill_manual(
    name   = "SNIC flag",
    values = c(seed = "#e31a1c", candidate = "#ff9b1f")
  ) +
  scale_linetype_manual(
    name   = "Threshold",
    values = c(seed = "dashed", candidate = "dotted")
  ) +
  scale_x_continuous(breaks = seq(0, 1, by = 0.25)) +
  labs(
    x        = "Fractional year",
    y        = NULL,
    title    = "Per-observation features and SNIC seed/candidate flags",
    subtitle = "Red = seed, orange = candidate. Dashed/dotted greys mark the seed and candidate thresholds on τ_pre, τ_post, and J."
  )
```

## Special case: low-severity forest burns (Patagonia)

In Patagonian forests, **low-severity fires can kill the canopy on a delayed schedule** — the burn probability rises slowly over weeks to months rather than jumping abruptly. To the focal jumper $J^{w,\text{adapt}}_t$ this looks indistinguishable from drought, and the SNIC seed/candidate criteria above will not fire because $\Delta p_t$ stays small at every individual step.

The discriminating signature is the *combination* of: a real, sustained probability after the slow rise (unlike drought, which recovers); a pre-disturbance state that was clearly healthy forest; and the *absence* of any sharp-edged signal — both in the year of interest and in the previous year.

**Candidate criterion** — forest class, Patagonia only, candidates only (never seeds, because we have no reliable transition date):

$$
\begin{aligned}
&\max_t \,\operatorname{persist}^{\text{adapt}}_{Y,\,t}    &&\ge\, \theta_{\text{persist}}                && \text{(crown does die — unlike drought)} \\
\wedge\;&\,\text{NDVI}^{\text{wet}}_{Y-1}                   &&\ge\, \theta_{\text{NDVI}}                    && \text{(was a healthy forest)} \\
\wedge\;&\max_t \, J^{w,\text{adapt}}_{Y,\,t}               &&\le\, \theta_{J,\,\text{low}}                 && \text{(no abrupt jump this year — unlike fire)} \\
\wedge\;&\max_t \, J^{w,\text{adapt}}_{Y-1,\,t}             &&\le\, \theta_{J,\,\text{low}}                 && \text{(no abrupt jump last year either)} \\
\wedge\;&\,\text{land cover}_{Y-1}                          &&=\,   \text{forest}                           &&  \\
\wedge\;&\,\text{region}                                    &&=\,   \text{PAT}                              &&
\end{aligned}
$$

The two annual-max constraints are the key. They together separate this regime from the three nearest distractors:

| Distractor                                    | What breaks the criterion                                              |
|-----------------------------------------------|------------------------------------------------------------------------|
| Drought (slow rise, full recovery)            | $\max_t \operatorname{persist}^{\text{adapt}}_{Y,t}$ stays low — the prob returns to zero |
| Regular fire (sharp transition)               | $\max_t J^{w,\text{adapt}}_{Y,t}$ is high — fails the "low jumper this year" check        |
| Burned previous year, still looks dead        | $\max_t J^{w,\text{adapt}}_{Y-1,t}$ is high — fails the "low jumper last year" check     |
| Non-forest land cover with similar dynamics   | Fails the land-cover gate                                              |

These pixels enter SNIC strictly as **candidates** for the forest fire class, never as seeds. They depend on spatial coherence with neighbouring sharp-edge fire seeds to be retained — a delayed-mortality patch typically borders the higher-severity core, so the candidate ring attaches to the seed-anchored segment.

The criterion is intentionally narrow: it would generate many false positives if applied outside forest land cover or outside Patagonia, where slow-mortality dynamics have different causes (e.g., agriculture rotation, shrub senescence).

