---
title: "Land cover remap v2 — fire-class definitions for LR fitting"
subtitle: "Discussion document: how should we remap MapBiomas land cover into fittable fire-classes?"
author: "MapBiomas Argentina Fire"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: true
    df-print: kable
execute:
  warning: false
  message: false
---

## Why this notebook

The first remap notebook (`land_cover_remap.qmd`) validated two competing remap
*proposals* read live from the Google Sheet. Since then the remap has been frozen into a
canonical, code-readable table — `config/veg_fire_remap.csv` — which both the Python
pipeline and the R fitting code now read. This notebook is the v2 follow-up: it evaluates
**the current canonical remap as it actually stands**, using the full training
observations (`training_observations_*_v1.csv`, all five regions), so the team can confirm
the fire-classes are fittable **before** fitting all the models.

The motivating finding was the CV feasibility report
(`scripts/cv_feasibility_report.py` → `models/cv_feasibility_v1.csv`): grouped
leave-fires-out K-fold CV is only meaningful when a class has enough distinct fires
carrying positive (burned) observations. An earlier pass flagged one class as unfittable
(`agriculture_cuyo`, only **2** fires with positives) plus 183 CUYO observations whose
MapBiomas class was **not in the remap at all**. The remap has since been revised — CUYO's
cultivated cover is merged with the Patagonian agriculture class into a single cross-region
class `agriculture_cuyo-pat` (which also absorbs the previously-unmapped class 19) — and
this notebook confirms those gaps are now closed.

This notebook reproduces and extends that report with two tables:

- **Table 1 — raw class × region.** One row per `(mb_class_raw, region)`: the MapBiomas
  source class, the `veg_fire` class it is proposed to map to, observation counts,
  burned/unburned split, number of fires carrying positives, per-fire robustness, and the
  long-run area fraction the class occupies in its region.
- **Table 2 — proposed remap summary.** One row per `veg_fire` class, aggregated across
  its regions, with the same columns as `cv_feasibility_v1.csv` (obs, pos, %pos, fires+,
  K, pos/fold, ash%neg) — the basis for the merge / split / drop decision.

Everything is computed from the same single pass over the raw observations, so the two
tables are mutually consistent. Fire-level CV statistics key on a **region-unique fire id**
(`region_fireid`, e.g. `BA_fire_01`) because bare `fire_id`s repeat across regions and
would otherwise conflate physically distinct fires — the pipeline
(`scripts/cv_feasibility_report.py` and `workflow/02-model_fitting.R`) uses the same key.
Table 2 is then cross-checked against the canonical `cv_feasibility_v1.csv` as a regression
guard; it should match exactly.

> **CV vocabulary.** *fires+* = number of distinct fires that carry at least one burned
> observation in the class. *K* = `min(10, fires+)` — the realized number of CV folds
> (we hold out whole fires, not random rows). *pos/fold* = positives ÷ K, the average
> number of burned observations per held-out fold; below ~50 a fold is too thin to
> estimate burn probability reliably. *ash%neg* = share of a class's *negatives* that come
> from **pure-negative fires** (fires with zero burned observations anywhere — ash /
> drought false-positive controls such as PAT `fire_46`/`fire_47`). Those are kept as
> point-distributed hard negatives, **not** treated as held-out folds.

## Setup — load remap and observations

```{r setup}
library(data.table)
library(tidyverse)
library(knitr)
library(glue)

REGIONS <- c("BA", "CHACO", "PAMPA", "CUYO", "PAT")
VERSION <- 1L
TARGET_K         <- 10L   # cap on number of leave-fires-out folds
MIN_POS_PER_FOLD <- 50L   # advisory floor for a usable fold

# ── Canonical remap (config CSV — the single source of truth) ────────────────
remap <- fread(here::here("..", "config", "veg_fire_remap.csv"))
remap[, mb_class_raw := as.integer(mb_class_raw)]

# ── Training observations — read only the columns we need (files are ~1 GB) ──
# region is taken from the filename, matching how cv_feasibility_report.py keys
# the remap join — fire_ids repeat across regions, so the file's region is the
# authoritative key, not any region column inside the CSV.
obs <- rbindlist(lapply(REGIONS, function(r) {
  d <- fread(
    here::here("..", "data", glue("training_observations_{r}_v{VERSION}.csv")),
    select = c("fire_id", "burned", "mb_class_raw")
  )
  d[, region := r]
  d[]
}))
obs[, `:=`(burned = as.integer(burned), mb_class_raw = as.integer(mb_class_raw))]

# fire_ids repeat across regions (fire_01 exists in every region), so they are
# not globally unique. Build a region-unique fire key, region_fireid (e.g.
# "BA_fire_01"). Every fire-level CV statistic below keys on this — NOT on the
# bare fire_id, which would conflate physically distinct fires across regions.
obs[, fire_uid := paste0(region, "_", fire_id)]

# ── Pure-negative fires: zero burned observations anywhere (ash / drought) ───
fire_npos      <- obs[, .(npos = sum(burned)), by = fire_uid]
pure_neg_fires <- fire_npos[npos == 0L, fire_uid]

# ── Join the remap on (region, mb_class_raw); unmatched rows are "unmapped" ──
rk <- remap[, .(region, mb_class_raw, arg_name, local_class,
                veg_fire, veg_fire_name, fittable)]
obs_m <- rk[obs, on = c("region", "mb_class_raw")]   # keep every observation

cat(
  "Observations read:", format(nrow(obs), big.mark = ","), "\n",
  "Regions:", paste(REGIONS, collapse = ", "), "\n",
  "Pure-negative fires:", paste(sort(pure_neg_fires), collapse = ", "), "\n",
  "Unmapped observations (no veg_fire):",
  format(obs_m[is.na(veg_fire_name), .N], big.mark = ","), "\n"
)
```

## Unmapped classes — present in the data but absent from the remap

Any `(region, mb_class_raw)` that appears in the training observations but has no row in
`config/veg_fire_remap.csv`. These observations are silently dropped at fit time, so any
class with non-trivial counts here is a remap gap to close.

```{r unmapped}
unmapped <- obs_m[is.na(veg_fire_name),
                  .(n_obs = .N, burned = sum(burned)),
                  by = .(region, mb_class_raw)][order(region, -n_obs)]

# Attach the MapBiomas country-level name for context (from any region's remap row).
mb_names <- unique(remap[, .(mb_class_raw, arg_name)])[
  , .(arg_name = first(arg_name)), by = mb_class_raw]
unmapped <- mb_names[unmapped, on = "mb_class_raw"]
setcolorder(unmapped, c("region", "mb_class_raw", "arg_name", "n_obs", "burned"))

if (nrow(unmapped) == 0) {
  cat("No unmapped observations — every (region, mb_class) is in the remap.\n")
} else {
  kable(unmapped,
        col.names = c("region", "MB id", "MB class (country)", "n obs", "burned"),
        format.args = list(big.mark = ","))
}
```

## Area fraction per class (long-run, 1999–2023)

Mean mapped area per region-class from the `areas_regiones` sheet, expressed as a fraction
of the region's total mapped area. Used to weight how much of the landscape a fire-class
actually covers (a class can be tiny in observations but large on the ground, or vice
versa). Read live from the working Google Sheet; if unreachable, area columns fall back to
`NA` and the rest of the notebook is unaffected.

```{r area}
sheet_url <- "https://docs.google.com/spreadsheets/d/17ZShb8D0JaJw4nLvBDzt19xF6Fdg8lGHYs4Jogh0X1A"

area_by_class <- tryCatch({
  library(googlesheets4)
  gs4_deauth()
  areas_raw <- read_sheet(sheet_url, sheet = "areas_regiones")

  year_cols <- names(areas_raw)[
    suppressWarnings(!is.na(as.integer(names(areas_raw)))) &
    suppressWarnings(as.integer(names(areas_raw)) >= 1999) &
    suppressWarnings(as.integer(names(areas_raw)) <= 2023)
  ]

  areas <- areas_raw |>
    mutate(id_local = suppressWarnings(as.integer(unlist(id_local)))) |>
    filter(!is.na(id_local)) |>
    mutate(mean_area = rowMeans(pick(all_of(year_cols)), na.rm = TRUE)) |>
    select(region, id_local, mean_area)

  # Crosswalk MB class → region-local ids by pulling "ID=NN" out of local_class.
  crosswalk <- as_tibble(remap) |>
    filter(!is.na(local_class), local_class != "") |>
    select(region, mb_class_raw, local_class) |>
    mutate(id_local = str_extract_all(local_class, "(?<=ID=)\\s*\\d+") |>
                      map(~ as.integer(str_trim(.x)))) |>
    unnest(id_local) |>
    distinct(region, mb_class_raw, id_local)

  crosswalk |>
    left_join(areas, by = c("region", "id_local"),
              relationship = "many-to-many") |>
    group_by(region, mb_class_raw) |>
    summarise(mean_area = sum(mean_area, na.rm = TRUE), .groups = "drop") |>
    group_by(region) |>
    mutate(area_frac = mean_area / sum(mean_area)) |>
    ungroup() |>
    select(region, mb_class_raw, area_frac) |>
    as.data.table()
}, error = function(e) {
  message("Area sheet read failed (", conditionMessage(e),
          ") — area columns will be NA.")
  data.table(region = character(), mb_class_raw = integer(),
             area_frac = double())
})

cat("Area crosswalk rows:", nrow(area_by_class), "\n")
```

## Table 1 — raw class × region

One row per `(mb_class_raw, region)` mapped class, grouped by the proposed `veg_fire`
class and ordered within each region by `veg_fire` code. `burned (%)` is positives over
total observations for that raw class; `fires+` and `med pos/fire` describe how those
positives spread across fires (a class whose burned obs sit in one or two fires is fragile
even if its total is large); `area %` is the long-run area share within the region.

```{r table1, results='asis'}
# Per-fire robustness for each raw class: fires carrying positives, and the
# median number of positives per such fire.
rob <- obs_m[burned == 1L, .(npos = .N), by = .(region, mb_class_raw, fire_uid)][
  , .(fires_pos = .N, med_pos_per_fire = as.double(median(npos))),
  by = .(region, mb_class_raw)]

t1 <- obs_m[!is.na(veg_fire_name), .(
  arg_name      = first(arg_name),
  local_class   = first(local_class),
  veg_fire      = first(veg_fire),
  veg_fire_name = first(veg_fire_name),
  fittable      = first(fittable),
  burned        = sum(burned),
  unburned      = sum(burned == 0L),
  n_obs         = .N
), by = .(region, mb_class_raw)]

t1 <- rob[t1, on = c("region", "mb_class_raw")]
t1 <- area_by_class[t1, on = c("region", "mb_class_raw")]
t1[is.na(fires_pos),       fires_pos := 0L]
t1[is.na(med_pos_per_fire), med_pos_per_fire := 0]
t1[, burned_pct := paste0(format(burned, big.mark = ","),
                          " (", round(100 * burned / n_obs, 1), "%)")]
t1[, area_pct := ifelse(is.na(area_frac), NA_real_, round(100 * area_frac, 2))]

for (rg in REGIONS) {
  cat("\n\n### Region:", rg, "\n\n")
  d <- t1[region == rg][order(veg_fire, mb_class_raw)]
  if (nrow(d) == 0) { cat("_No mapped observations._\n"); next }
  print(kable(
    d[, .(mb_class_raw, arg_name, local_class,
          veg_fire_name, veg_fire, fittable,
          burned_pct, unburned, n_obs, fires_pos, med_pos_per_fire, area_pct)],
    format = "markdown",
    col.names = c("MB id", "MB class (country)", "MB class (region)",
                  "→ veg_fire", "code", "fit?",
                  "burned (%)", "unburned", "n obs",
                  "fires+", "med pos/fire", "area %"),
    format.args = list(big.mark = ",")
  ))
  cat("\n")
}
```

## Table 2 — proposed remap summary (one row per veg_fire class)

Aggregated across each class's regions. Columns mirror `cv_feasibility_v1.csv`. Sorted
fittable-first, then by observation count. `K` is the realized fold count
(`min(10, fires+)`); `pos/fold` is positives ÷ K. Flags mark the classes that cannot be
cross-validated cleanly.

```{r table2}
# Regions a class spans per the config (matches cv_feasibility's `regions`).
class_regions <- remap[!is.na(veg_fire_name) & veg_fire_name != "",
  .(regions = paste(sort(unique(region)), collapse = "+")), by = veg_fire_name]

t2 <- obs_m[!is.na(veg_fire_name), .(
  veg_fire    = first(veg_fire),
  fittable    = first(fittable),
  n_obs       = .N,
  n_pos       = sum(burned),
  fires_pos   = uniqueN(fire_uid[burned == 1L])
), by = veg_fire_name]

ash <- obs_m[!is.na(veg_fire_name) & burned == 0L, .(
  n_neg   = .N,
  ash_neg = sum(fire_uid %in% pure_neg_fires)
), by = veg_fire_name]

t2 <- ash[t2, on = "veg_fire_name"]
t2 <- class_regions[t2, on = "veg_fire_name"]
t2[, `:=`(
  pct_pos = round(100 * n_pos / n_obs, 1),
  K       = fifelse(fittable, pmin(TARGET_K, fires_pos), 0L),
  ash_pct = fifelse(n_neg > 0L, round(100 * ash_neg / n_neg, 1), 0)
)]
t2[, pos_per_fold := fifelse(K > 0L, round(n_pos / K, 1), 0)]

# Low K (<10 fires) is accepted as fine — only genuinely broken cases are flagged:
# K≤2 (CV impossible) and thin folds (<50 positives/fold).
t2[, flag := fcase(
  !fittable,                              "non-fittable",
  K <= 2L,                                "✗ CV impossible (K≤2)",
  pos_per_fold < MIN_POS_PER_FOLD,        "⚠ thin folds (<50/fold)",
  default = "ok"
)]

setorder(t2, -fittable, -n_obs)

kable(
  t2[, .(veg_fire_name, veg_fire, regions, fittable,
         n_obs, n_pos, pct_pos, fires_pos, K, pos_per_fold, ash_pct, flag)],
  format = "markdown",
  col.names = c("veg_fire", "code", "regions", "fit?",
                "obs", "pos", "%pos", "fires+", "K", "pos/fold", "ash%neg", "flag"),
  format.args = list(big.mark = ",")
)
```

### Cross-check against the canonical `cv_feasibility_v1.csv`

Table 2 is recomputed here from the raw observations and compared to the pipeline's own
report. Both now key fire-level statistics on the region-unique `region_fireid`, so this is
a regression guard: every column should match exactly. Any row below is a discrepancy —
most likely a stale `cv_feasibility_v1.csv` (re-run `cv_feasibility_report.py` after a data
change to refresh it).

> Before the region-key fix the pipeline keyed on the bare `fire_id`, which conflated fires
> sharing an id across regions: `shrubland_cuyo-pampa` reported `fires+` 29 instead of 38,
> and the four PAMPA pure-negative fires were missed, understating `ash%neg` for the Pampa
> classes (e.g. `agriculture_pampa` 0.0 → 18.1).

```{r verify}
cvf <- fread(here::here("..", "models", "cv_feasibility_v1.csv"))

cmp <- merge(
  t2[, .(veg_fire_name, obs_nb = n_obs, pos_nb = n_pos, fpos_nb = fires_pos,
         K_nb = K, posfold_nb = pos_per_fold, ash_nb = ash_pct)],
  cvf[, .(veg_fire_name, obs_cv = n_obs, pos_cv = n_pos, fpos_cv = n_fires_with_pos,
          K_cv = realized_K, posfold_cv = mean_pos_per_fold,
          ash_cv = pct_neg_from_pure_neg_fires)],
  by = "veg_fire_name", all = TRUE
)

diffs <- cmp[is.na(obs_nb) | is.na(obs_cv) |
             obs_nb != obs_cv | pos_nb != pos_cv | fpos_nb != fpos_cv |
             K_nb != K_cv | abs(posfold_nb - posfold_cv) > 0.2 |
             abs(ash_nb - ash_cv) > 0.2]

if (nrow(diffs) == 0) {
  cat("✔ Table 2 matches cv_feasibility_v1.csv exactly for all", nrow(cmp),
      "classes (obs, pos, fires+, K, pos/fold, ash%neg).\n")
} else {
  cat("⚠", nrow(diffs), "class(es) differ from cv_feasibility_v1.csv ",
      "(re-run cv_feasibility_report.py to refresh it):\n\n")
  print(kable(diffs, format = "markdown", format.args = list(big.mark = ",")))
}
```

## Discussion — the remap as it now stands

```{r discussion-figs}
n_fittable <- t2[fittable == TRUE, .N]
n_unmapped_obs <- if (nrow(unmapped) == 0) 0L else sum(unmapped$n_obs)
agcp <- t2[veg_fire_name == "agriculture_cuyo-pat"]
```

The remap was revised after the first CV-feasibility pass, and the two blocking problems it
had flagged are now resolved. With **`r n_fittable`** fittable classes, the numbers below are
pulled live from Table 2 / the unmapped table.

**1. `agriculture_cuyo` (formerly K=2, unfittable) is gone.** CUYO's cultivated cover is now
merged with the Patagonian agriculture class into a single cross-region class,
**`agriculture_cuyo-pat`** (`r format(agcp$n_obs, big.mark=",")` obs, `r format(agcp$n_pos,
big.mark=",")` pos, `r agcp$fires_pos` fires with positives → K=`r agcp$K`). Pooling the two
regions' positive-bearing fires lifts it well clear of the two-fold limit that made the
standalone CUYO class impossible to cross-validate.

**2. CUYO MapBiomas class 19 ("Cultivos temporarios") is now mapped** (also to
`agriculture_cuyo-pat`), closing the silent-drop gap — across all regions
`r format(n_unmapped_obs, big.mark=",")` observations now fall outside the remap (see the
unmapped table above).

**3. Low-K classes are accepted as fittable.** A handful of classes carry their positives in
fewer than ten fires, so their grouped CV rests on fewer held-out folds. This is treated as
fine: the classes are fit normally and no remap change is made on account of K alone. Only
genuinely broken cases (K≤2, or very thin folds) are flagged in Table 2.

**4. Ash-heavy negative classes.** Classes with a high `ash%neg` (notably `grassland_pat`,
`shrubland_pat`, and `agriculture_pampa`) draw a large share of their negatives from the
pure-negative ash/drought fires (`r paste(sort(pure_neg_fires), collapse=", ")`). Four of
those are PAMPA fires that are pure-negative within Pampa but share bare ids with burning
fires elsewhere — they only surface once fires are keyed region-uniquely (`region_fireid`).
That is by design: those points are the hard negatives the model most needs. But it means
the positive:negative balance in those classes is partly driven by the false-positive
controls, not by pre-fire windows. Keep them; just be aware the fold composition differs
from the other classes.

**5. Everything else is healthy.** The large fittable classes (`grassland_pampa`,
`forest-cerr_chaco`, `shrubland_cuyo-pampa`, `forest_ba`, …) all reach K=10 with thousands
of positives per fold and need no remap change.

### Status

The remap is ready to fit: the `agriculture_cuyo`/class-19 gaps are closed and every
fittable class cross-validates. The one fitting consequence is that the old standalone
`agriculture_pat` model is superseded — its output must be removed and the new
`agriculture_cuyo-pat` class fitted in its place.
