---
title: "Land cover remap for burn probability modelling"
author: "MapBiomas Argentina Fire"
date: today
---

## Motivation

Two decisions need to be made before fitting the Random Forest model:

1. **Vegetation type reclassification** — validate that the proposed MapBiomas → fire-class
   mapping produces groups large enough to train separate classifiers. If a class has too few
   burned observations it should be merged or dropped.

2. **Unburned downsampling** — the pre-fire window is long and unburned points also contribute
   unburned observations, so unburned likely outnumbers burned by a large margin. Assess
   whether downsampling is warranted.

Analysis is done **by `region_fire`**, with BA and CHACO merged into a single `CHACO-BA` group
(they share the same proposed remap). Fires `fire_46` and `fire_47` (PAT false-positive
controls — ash / drought) are excluded from class-balance counts.

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(googlesheets4)
library(knitr)

# ── Remap table from Google Sheets ──────────────────────────────────────────
# Sheet: https://docs.google.com/spreadsheets/d/17ZShb8D0JaJw4nLvBDzt19xF6Fdg8lGHYs4Jogh0X1A/edit?gid=1376068841
gs4_deauth()   # sheet is readable without auth
remap_raw <- read_sheet(
  "https://docs.google.com/spreadsheets/d/17ZShb8D0JaJw4nLvBDzt19xF6Fdg8lGHYs4Jogh0X1A",
  sheet = "remap_by_region"
)

remap <- remap_raw |>
  mutate(mb_class_raw = suppressWarnings(as.integer(unlist(id)))) |>
  filter(!is.na(mb_class_raw)) |>
  select(mb_class_raw, arg_name, region_name, region_fire,
         veg_fire_name_1, veg_fire_name_2) |>
  group_by(mb_class_raw, region_fire) |>
  summarise(
    arg_name        = first(arg_name),
    region_name     = first(na.omit(region_name)),
    veg_fire_name_1 = first(veg_fire_name_1),
    veg_fire_name_2 = first(veg_fire_name_2),
    .groups         = "drop"
  )

# ── region_name_ambig: region-level class name, with CHACO/BA shown separately
# when they differ (CHACO-BA rows in remap_raw come from two distinct regions).
region_name_ambig_tbl <- remap_raw |>
  mutate(mb_class_raw = suppressWarnings(as.integer(unlist(id)))) |>
  filter(!is.na(mb_class_raw), !is.na(region_name)) |>
  select(mb_class_raw, region, region_fire, region_name) |>
  distinct() |>
  mutate(rn_label = paste0(region, ": ", region_name)) |>
  group_by(mb_class_raw, region_fire) |>
  summarise(
    region_name_ambig = if (n_distinct(region_name) == 1) {
      first(region_name)
    } else if (n_distinct(region) == 1) {
      # Multiple names within one region (e.g. CUYO class 63): no region prefix needed.
      paste(sort(region_name), collapse = "; ")
    } else {
      # Multiple regions merged into one region_fire (CHACO-BA): prefix with region.
      paste(sort(rn_label), collapse = "; ")
    },
    .groups = "drop"
  )

remap <- remap |>
  left_join(region_name_ambig_tbl, by = c("mb_class_raw", "region_fire"))

# Class 75 (turberas) is absent from the sheet (which uses 73); add it as
# grassland for all region_fires so training observations are not lost.
remap <- bind_rows(
  remap,
  expand_grid(region_fire = c("CUYO", "PAT", "PAMPA", "CHACO-BA"), mb_class_raw = 75L) |>
    mutate(arg_name = "Turberas", region_name = NA_character_,
           veg_fire_name_1 = "grassland", veg_fire_name_2 = "grassland",
           region_name_ambig = NA_character_)
)

# Reference for remap tables: all 6 fire classes in display order.
all_veg_classes <- c("forest", "shrubland", "grassland", "agriculture",
                     "non-burnable", "non-observed")

# ── Area distribution from Google Sheets ─────────────────────────────────────
areas_raw <- read_sheet(
  "https://docs.google.com/spreadsheets/d/17ZShb8D0JaJw4nLvBDzt19xF6Fdg8lGHYs4Jogh0X1A",
  sheet = "areas_regiones"
)

year_cols <- names(areas_raw)[
  suppressWarnings(!is.na(as.integer(names(areas_raw)))) &
  suppressWarnings(as.integer(names(areas_raw)) >= 1999) &
  suppressWarnings(as.integer(names(areas_raw)) <= 2023)
]
areas <- areas_raw |>
  mutate(id_local = suppressWarnings(as.integer(unlist(id_local)))) |>
  filter(!is.na(id_local)) |>
  mutate(mean_area = rowMeans(pick(all_of(year_cols)), na.rm = TRUE)) |>
  select(region, id_local, mean_area) |>
  mutate(region_fire = case_when(
    region %in% c("BA", "CHACO") ~ "CHACO-BA",
    .default = region
  ))

crosswalk <- remap_raw |>
  mutate(mb_class_raw = suppressWarnings(as.integer(unlist(id)))) |>
  filter(!is.na(mb_class_raw), !is.na(region_name)) |>
  select(mb_class_raw, region_fire, region_name) |>
  mutate(id_local = str_extract_all(region_name, "(?<=ID=)\\s*\\d+") |>
                    map(~ as.integer(str_trim(.x)))) |>
  unnest(id_local) |>
  select(mb_class_raw, region_fire, id_local) |>
  distinct()

area_by_rf <- crosswalk |>
  left_join(areas, by = c("region_fire", "id_local"), relationship = "many-to-many") |>
  group_by(region_fire, mb_class_raw) |>
  summarise(mean_area = sum(mean_area, na.rm = TRUE), .groups = "drop") |>
  group_by(region_fire) |>
  mutate(area_frac = mean_area / sum(mean_area)) |>
  ungroup() |>
  select(region_fire, mb_class_raw, area_frac)

# ── Helper: build one remap summary table (per proposal column) ──────────────
# Returns: veg_class | burned (%) | unburned | obs frac (%) | area frac (%) |
#          n fires | med obs/fire
make_remap_tbl <- function(obs_r, veg_col, area_r, remap_r, total_obs) {
  counts <- obs_r |>
    count(.data[[veg_col]], burned_label, name = "n") |>
    pivot_wider(names_from = burned_label, values_from = n, values_fill = 0L) |>
    rename(veg_class = all_of(veg_col))

  robustness <- obs_r |>
    filter(burned == 1L) |>
    count(.data[[veg_col]], fire_id, name = "n_burned") |>
    group_by(veg_class = .data[[veg_col]]) |>
    summarise(
      n_fires      = n_distinct(fire_id),
      med_obs_fire = median(n_burned),
      .groups = "drop"
    )

  area_agg <- area_r |>
    left_join(remap_r |> select(mb_class_raw, veg_class = all_of(veg_col)),
              by = "mb_class_raw") |>
    group_by(veg_class) |>
    summarise(area_frac = sum(area_frac, na.rm = TRUE), .groups = "drop")

  tibble(veg_class = all_veg_classes) |>
    left_join(counts,     by = "veg_class") |>
    left_join(area_agg,   by = "veg_class") |>
    left_join(robustness, by = "veg_class") |>
    mutate(
      burned   = replace_na(burned,   0L),
      unburned = replace_na(unburned, 0L),
      total    = burned + unburned,
      burned_pct    = if_else(total > 0L,
                       paste0(burned, " (", round(burned / total * 100, 1), "%)"),
                       as.character(burned)),
      obs_frac_pct  = round(total / total_obs * 100, 2),
      area_frac_pct = round(replace_na(area_frac, 0) * 100, 2),
      n_fires       = replace_na(n_fires, 0L)
    ) |>
    select(veg_class, burned_pct, unburned, obs_frac_pct, area_frac_pct,
           n_fires, med_obs_fire)
}

# ── Training observations — all regions ─────────────────────────────────────
regions <- c("PAT", "BA", "CHACO", "CUYO", "PAMPA")
# Version of the training_observations export to read (PAT re-runs more often).
obs_versions <- c(PAT = 5, BA = 3, CHACO = 3, CUYO = 3, PAMPA = 3)

obs_all <- map_dfr(regions, \(r) {
  read_csv(
    here::here("..", "data", glue::glue("training_observations_{r}_v{obs_versions[r]}.csv")),
    col_types = cols(
      fire_id      = col_character(),
      burned       = col_integer(),
      mb_class_raw = col_integer(),
      focal_year   = col_integer(),
      date         = col_date(),
      region       = col_character(),
      .default     = col_double()
    ),
    show_col_types = FALSE
  )
}) |>
  mutate(
    region_fire  = case_when(
      region %in% c("BA", "CHACO") ~ "CHACO-BA",
      .default = region
    ),
    burned_label = if_else(burned == 1L, "burned", "unburned")
  ) |>
  filter(!fire_id %in% c("fire_46", "fire_47"))

cat(
  "Total rows (excl. fire_46/47):", nrow(obs_all), "\n",
  "Regions:", paste(sort(unique(obs_all$region)), collapse = ", "), "\n",
  "Region-fires:", paste(sort(unique(obs_all$region_fire)), collapse = ", "), "\n"
)
```

```{r join_remap, cache=FALSE}
obs_remapped <- obs_all |>
  left_join(remap, by = c("mb_class_raw", "region_fire"))
```

---

## Land cover remaps by region

*For each region: raw MapBiomas class breakdown, then the proposed remap.*

```{r region_tables, results='asis'}
region_fires <- c("PAT", "CUYO", "PAMPA", "CHACO-BA")

walk(region_fires, \(rf) {
  cat("\n\n### Region:", rf, "\n\n")

  obs_r     <- obs_remapped |> filter(region_fire == rf)
  area_r    <- area_by_rf   |> filter(region_fire == rf)
  remap_r   <- remap        |> filter(region_fire == rf)
  total_obs <- nrow(obs_all |> filter(region_fire == rf))

  # Compute both remap tables first; if the output is identical (e.g. because
  # differing MB classes are absent from training data) treat as one proposal.
  remap_tbl_1 <- make_remap_tbl(obs_r, "veg_fire_name_1", area_r, remap_r, total_obs)
  remap_tbl_2 <- make_remap_tbl(obs_r, "veg_fire_name_2", area_r, remap_r, total_obs)
  same_remap  <- identical(remap_tbl_1, remap_tbl_2)

  remap_col_names <- c("veg class", "burned (%)", "unburned",
                       "obs frac (%)", "area frac (%)", "n fires", "med obs/fire")

  # ── 1a. Raw MapBiomas classes ────────────────────────────────────────────
  cat("#### Raw MapBiomas classes\n\n")

  classes_r <- remap |>
    filter(region_fire == rf, !is.na(region_name)) |>
    select(mb_class_raw, arg_name, region_name_ambig, veg_fire_name_1, veg_fire_name_2)

  obs_counts_wide <- obs_all |>
    filter(region_fire == rf) |>
    count(mb_class_raw, burned_label) |>
    pivot_wider(names_from = burned_label, values_from = n, values_fill = 0L)

  raw_tbl <- classes_r |>
    left_join(obs_counts_wide, by = "mb_class_raw") |>
    left_join(select(area_r, mb_class_raw, area_frac), by = "mb_class_raw") |>
    mutate(
      burned   = replace_na(burned,   0L),
      unburned = replace_na(unburned, 0L),
      total    = burned + unburned,
      burned_pct    = if_else(total > 0L,
                       paste0(burned, " (", round(burned / total * 100, 1), "%)"),
                       as.character(burned)),
      obs_frac_pct  = round(total / total_obs * 100, 2),
      area_frac_pct = round(replace_na(area_frac, 0) * 100, 2)
    ) |>
    arrange(match(veg_fire_name_1, all_veg_classes), mb_class_raw)

  if (same_remap) {
    print(kable(
      raw_tbl |> select(mb_class_raw, arg_name, region_name_ambig,
                        veg_fire_name_1,
                        burned_pct, unburned, obs_frac_pct, area_frac_pct),
      format = "markdown",
      col.names = c("MB id", "MB class (country)", "MB class (region)",
                    "remap prop.",
                    "burned (%)", "unburned", "obs frac (%)", "area frac (%)")
    ))
  } else {
    raw_tbl_display <- raw_tbl |>
      select(mb_class_raw, arg_name, region_name_ambig,
             veg_fire_name_1, veg_fire_name_2,
             burned_pct, unburned, obs_frac_pct, area_frac_pct) |>
      mutate(
        differs         = !is.na(veg_fire_name_1) & !is.na(veg_fire_name_2) &
                          veg_fire_name_1 != veg_fire_name_2,
        veg_fire_name_1 = if_else(differs, paste0("**", veg_fire_name_1, "**"), veg_fire_name_1),
        veg_fire_name_2 = if_else(differs, paste0("**", veg_fire_name_2, "**"), veg_fire_name_2)
      ) |>
      select(-differs)
    print(kable(
      raw_tbl_display,
      format = "markdown",
      col.names = c("MB id", "MB class (country)", "MB class (region)",
                    "remap prop. 1", "remap prop. 2",
                    "burned (%)", "unburned", "obs frac (%)", "area frac (%)")
    ))
  }
  cat("\n")

  # ── 1b/c. Remap table(s) ────────────────────────────────────────────────
  if (same_remap) {
    cat("#### Remap proposal\n\n")
    print(kable(remap_tbl_1, format = "markdown", col.names = remap_col_names))
    cat("\n")
  } else {
    cat("#### Remap proposal 1\n\n")
    print(kable(remap_tbl_1, format = "markdown", col.names = remap_col_names))
    cat("\n")
    cat("#### Remap proposal 2\n\n")
    print(kable(remap_tbl_2, format = "markdown", col.names = remap_col_names))
    cat("\n")
  }
  cat("\n---\n")
})
```

