---
title: "Observation-level burn-probability model — design"
subtitle: "From the canonical-team term set to the reduced 129-term elastic-net design (MapBiomas Argentina Fuego, Collection 1, step 02)"
author: "Iván Barberá"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    theme: cosmo
    code-fold: true
    df-print: kable
    fig-align: center
fontsize: 11pt
execute:
  warning: false
  message: false
  cache: false
---

## What this notebook is

This is the single, end-to-end record of how the Collection-1 observation-level
burn-probability model was designed. The story runs in one direction:

1. **Term design (the greedy start).** We first wrote down the *maximal* design we
   were willing to evaluate in Google Earth Engine — the **canonical-team set, 427
   terms**: 21 focal spectral features, 40 previous-year MapBiomas-mosaic bands, and
   six interaction blocks built on top of them. The strategy was deliberately greedy:
   give the elastic net as much flexibility as the GEE compute budget allows and let
   regularization sort it out.
2. **The convergence problem.** `glmnet`'s elastic-net path was *slow to converge* on
   that design. The diagnosis below shows the cause was **numerical, not statistical**:
   a heavily collinear design matrix makes coordinate descent crawl in the
   ill-conditioned low-$\lambda$ tail.
3. **The reduction protocol.** We then reduced the design from **427 → 129 terms** with
   a specific, reproducible protocol — pairwise-correlation pruning, an exact-linear-
   combination argument (which is how MIRBI and the tasseled-cap trio leave), a
   multivariate VIF screen, and a final fitting-config fix. Everything in Part III below
   **runs on the full 5.72M-observation training set at render time**.
4. **The final design + fitting config**, matching `workflow/02-model_fitting.R`
   term-for-term (that script is the source of truth).

> **This notebook renders on the full data.** The correlation, separability, eigenvalue,
> and VIF chunks read all five region tables (`collection-01/data/training_observations_*_v1.csv`,
> ~3.8 GB, ~5.72M obs) and compute on the complete set. It is therefore a heavy render
> (several minutes, several GB of RAM). The big matrices are `rm()`-ed and `gc()`-ed
> between sections to keep the peak bounded.

---

# Part I — Term design (the greedy start)

## Scope and motivation

MapBiomas Argentina Fuego Collection 1 produces annual burned-area maps for all of
Argentina over 1999–2025 using Landsat C2 SR. The pipeline relies on a per-observation
(pixel-date) burn-probability classifier evaluated for **every Landsat scene** intersecting
the territory and the time window. That prediction step is the cost driver: the model is
applied many billions of times, so the structure must be **cheap to evaluate** in GEE.

Collection 0 (Patagonia pilot) used a logistic regression with 40 terms per vegetation type,
four fire-relevant indices (NBR, NBR2, MIRBI, NDVI), and a hand-picked interaction set that
included custom annual min/max summaries and $\Delta$ values. Collection 1 keeps logistic
regression — for the same compute reasons — but rebalances the design:

- **More focal flexibility**: many spectral features per observation instead of 4.
- **Previous-year context from MapBiomas**: replace custom-computed annual summaries with
  the official MapBiomas Argentina annual mosaic of year $y-1$, which provides
  phenologically informed composites at no extra preprocessing cost.
- **No deltas**: with rich pairwise interactions, monotonic effects and the inherent
  non-linearity of the logistic link absorb most of what $\Delta$ terms did in Collection 0.
  A 2-way interaction in $\eta$ already implies a roughly 3-way effect on the predicted
  probability.
- **Flexibility over interpretability**: this model is for prediction, not inference.
  Collinearity is acceptable *in principle* and addressed by an elastic-net fit (tuning both
  $\lambda$ and $\alpha$). The catch — that "acceptable" collinearity still has to be
  *fittable* in finite time — is the entire subject of Part II.
- **Stratification**: one model per **region × vegetation type** (using the MapBiomas LULC
  of year $y-1$, reclassified into a few coarse fire-classes), as in Collection 0.

The original design constraint was roughly **200–300 terms per linear predictor** to keep the
GEE compute graph small. Below ~200 we lose flexibility; much above ~300 the graph grows
enough to slow full-territory prediction without obvious gains. The reduction in Part III lands
the final model comfortably inside that envelope, at **129 terms**.

## Why the MapBiomas annual mosaic (instead of custom summaries)

In Collection 0 we computed our own per-pixel annual minimum and maximum of each fire index
with a robust order-statistic rule. That was viable for one region × four indices but does not
scale: doing the same over all of Argentina, for two years per focal year ($y-1$ in addition to
$y$), with many more features, would multiply preprocessing cost considerably.

The MapBiomas Argentina annual mosaic already provides, for every year, a coherent set of
pixel-level composites:

- median, dry-season median, wet-season median, minimum, and within-year standard deviation of
  the six optical bands;
- the same statistics for a broad set of vegetation and fire-relevant indices (NDVI, NDWI,
  EVI2, SAVI, GCVI, NDFI, SEFI, WEFI, …);
- SMA endmember fractions (GV, GVS, NPV, soil, shade, cloud);
- topographic information.

Using this mosaic instead of recomputing custom summaries removes a heavy preprocessing step,
ensures consistency with the LULC product used for stratification, offers seasonal composites
(wet, dry) we never had in Collection 0, and provides physically interpretable derived layers
(NPV as a fuel-load proxy, NDFI as a fire/disturbance signal). The trade-off: we cannot freely
define the summaries; we choose among what the mosaic exposes.

## Which previous-year summaries (median, wet, sd — and the dry debate)

A previous-year summary is useful only insofar as it describes the **stable spectral state** of
the pixel — the baseline against which the focal observation is compared. Summaries that move
sharply when the pixel itself was disturbed in $y-1$ are confounded by exactly the kind of
signal we want to detect later.

| Subscript | Meaning |
|:---|:---|
| $X_t$ | Value of feature $X$ at the focal observation (pixel-date) |
| $X_{\text{med}}$ | MapBiomas mosaic **median** of $X$ in year $y(t) - 1$ |
| $X_{\text{wet}}$ | MapBiomas mosaic **wet-season median** of $X$ in year $y(t) - 1$ |
| $X_{\text{dry}}$ | MapBiomas mosaic **dry-season median** of $X$ in year $y(t) - 1$ |
| $X_{\text{sd}}$ | MapBiomas mosaic **within-year standard deviation** of $X$ in year $y(t) - 1$ |

The canonical-team design carried **all four** summaries ($\text{med}, \text{wet}, \text{dry},
\text{sd}$). The reasoning for and against each:

- $\text{med}$ — robust central tendency. Insensitive to a single fire event unless the fire
  affected a large fraction of the year's images. **Keep.**
- $\text{wet}$ — median of wet-season images. The wet season precedes the fire season, so a
  fire in $y-1$ generally occurs *after* the wet composite is laid down: the most fire-robust
  baseline in the mosaic. **Keep.**
- $\text{sd}$ — within-year variability. Contextualizes change: pixels with naturally high
  $\text{sd}$ should be judged against a more permissive burn threshold. **Keep.**
- $\text{dry}$ — the dry season *is* the fire season, so a $y-1$ fire contaminates the dry
  composite in the *same direction* as the future fire we want to detect (dry NDVI drops, dry
  SWIR rises). For pixels that did **not** burn in $y-1$ (the majority) it still carries peak-
  senescence / end-of-season fuel signal, so the canonical-team set kept it and let the net
  decide. In the reduction (Part III) it is dropped from the **interaction** blocks (kept only
  as a main effect) because it is partly redundant with $\text{med}$ and risks bias on
  previously-burned pixels.

We never used $\text{amp}$ (max − min, inflated by either fire or normal phenology) or bare
$\text{min}$/$\text{max}$ (which only supported $\Delta$ features we dropped).

## Spectral feature equations

The base focal feature set is **17 features** (six optical bands + 11 indices); the
canonical-team proposal extends this to **21** by adding AFRI, kNDVI, EVI2, NIRv.

| Group | Feature | Definition |
|:---|:---|:---|
| Optical | BLUE, GREEN, RED, NIR, SWIR1, SWIR2 | Surface-reflectance bands |
| Fire | $\text{NBR}$ | $(\text{NIR} - \text{SWIR2}) / (\text{NIR} + \text{SWIR2})$ |
| Fire | $\text{NBR2}$ | $(\text{SWIR1} - \text{SWIR2}) / (\text{SWIR1} + \text{SWIR2})$ |
| Fire | $\text{MIRBI}$ | $10\,\text{SWIR2} - 9.8\,\text{SWIR1} + 2$ |
| Fire | $\text{NDVI}$ | $(\text{NIR} - \text{RED}) / (\text{NIR} + \text{RED})$ |
| Tasseled-cap | $\text{TCB}$ | $0.3029\,\text{BLUE} + 0.2786\,\text{GREEN} + 0.4733\,\text{RED} + 0.5599\,\text{NIR} + 0.5080\,\text{SWIR1} + 0.1872\,\text{SWIR2}$ |
| Tasseled-cap | $\text{TCG}$ | $-0.2941\,\text{BLUE} - 0.2430\,\text{GREEN} - 0.5424\,\text{RED} + 0.7276\,\text{NIR} + 0.0713\,\text{SWIR1} - 0.1608\,\text{SWIR2}$ |
| Tasseled-cap | $\text{TCW}$ | $0.1511\,\text{BLUE} + 0.1973\,\text{GREEN} + 0.3283\,\text{RED} + 0.3407\,\text{NIR} - 0.7117\,\text{SWIR1} - 0.4559\,\text{SWIR2}$ |
| Aux. moisture | $\text{NDMI}$ | $(\text{NIR} - \text{SWIR1}) / (\text{NIR} + \text{SWIR1})$ |
| Aux. snow | $\text{NDSI}$ | $(\text{GREEN} - \text{SWIR1}) / (\text{GREEN} + \text{SWIR1})$ |
| Aux. sparse veg | $\text{SAVI}$ | $[(\text{NIR} - \text{RED}) / (\text{NIR} + \text{RED} + L)]\,(1 + L)$, $L = 0.5$ |
| Aux. open water | $\text{NDWI}$ | $(\text{GREEN} - \text{NIR}) / (\text{GREEN} + \text{NIR})$ |
| Aux. aerosol-free | $\text{AFRI}$ | $(\text{NIR} - 0.5\,\text{SWIR2}) / (\text{NIR} + 0.5\,\text{SWIR2})$ |
| Aux. veg. kernel | $\text{kNDVI}$ | $\tanh(\text{NDVI}^2)$ |
| Aux. enhanced veg | $\text{EVI2}$ | $2.5\,(\text{NIR} - \text{RED}) / (\text{NIR} + 2.4\,\text{RED} + 1)$ |
| Aux. NIR veg | $\text{NIRv}$ | $\text{NIR} \times \text{NDVI}$ |

Notes: tasseled-cap coefficients are Baig et al. (2014) for Landsat 8 OLI. MIRBI is in the raw
Trigg & Flasse (2001) orientation; it *increases* after fire (opposite sign of NBR, NBR2, NDVI).
NDWI here is McFeeters (1996), distinct from NDMI (Gao 1996). AFRI is AFRI-2100 (Karnieli et al.
2001, Eq. 11a); kNDVI uses the RBF-kernel form (Camps-Valls et al. 2021); EVI2 is the blue-band-
free EVI (Jiang et al. 2008); NIRv is Badgley et al. (2017).

**Note two facts that drive the whole reduction in Part III:**

- $\text{MIRBI} = 10\,\text{SWIR2} - 9.8\,\text{SWIR1} + 2$ is an **exact linear function** of
  two raw bands we keep.
- $\text{TCB}, \text{TCG}, \text{TCW}$ are **exact linear functions** of the six raw bands.
- $\text{NBR}, \text{NBR2}, \text{NDMI}$ are **ratios** of pairs of raw bands — *near*-functions
  of them (nonlinear, so not exact, but strongly collinear).

Previous-year variables beyond the six optical bands, taken directly from the mosaic: NDVI and
NDWI (same formulas as focal), NPV (non-photosynthetic vegetation fraction from SMA — fine-fuel
proxy), and NDFI (Normalized Difference Fraction Index, Souza et al. 2005 — disturbance signal).

## The canonical-team design — 427 terms

The linear predictor is
$$\Pr(y_{c,t} = 1) = \sigma\!\Big(\alpha_{k[c]} + \sum_{j} \beta_{j,k[c]}\, x_{j,c,t}\Big),$$
with $k[c]$ the vegetation cluster of pixel $c$ in year $y-1$ and $\sigma$ the logistic function.
The canonical-team set organizes the terms into six blocks:

| Block | Description | Count |
|:---|:---|---:|
| 1 | Focal raw (21 features) | 21 |
| 2 | Prev-year raw (10 vars × 4 summaries) | 40 |
| 3 | Focal × focal pairwise, $\binom{21}{2}$ | 210 |
| 4 | Same-band prev × focal (8 vars × 4 summaries) | 32 |
| 5 | Prev context × focal fire **indices** ($4\times4\times4 - 4$) | 60 |
| 6 | Prev context × focal fire **bands** ($4\times4\times4$) | 64 |
| | **Total** | **427** |

The modeling claim behind blocks 5–6 is that **the meaning of a focal fire-index value depends
on the previous-year vegetation context**: a low $\text{NBR}_t$ in a high-NPV, low-NDFI pixel
(heavy dead biomass, recently disturbed) is far more likely to be fire than the same
$\text{NBR}_t$ in a low-NPV, high-NDFI pixel. Block 3 (the 210 focal pairs) carries most of the
raw flexibility. This was the greedy maximum: maximal expressiveness within a ~15-node GEE graph
(see §[GEE]). The problem was getting it to *fit*.

---

# Part II — The convergence problem

`02-model_fitting.R`'s elastic-net search was **slow to converge** on the 427-term design —
some classes took many hours per $\alpha$, occasionally not finishing. Two hypotheses:

- **Statistical**: not enough signal / too many obs, intrinsically hard problem.
- **Numerical**: the design matrix is so collinear that `glmnet`'s coordinate descent crawls in
  the ill-conditioned, low-$\lambda$ tail (where the active set is large), and a coarse
  $\lambda$ grid gives poor warm starts.

The leading suspect was numerical. The canonical-team design expands 21 focal features + 40
mosaic bands into 427 terms via pairwise products and curated crosses. **If the raw inputs are
already strongly correlated, every interaction block inherits and amplifies that redundancy** —
products of collinear factors are even more collinear than the factors. Coordinate descent on a
near-singular Gram matrix converges geometrically slowly.

Part III tests that hypothesis directly (it is correct) and acts on it in two ways: **remove the
*removable* redundancy** (exact linear combinations and duplicated axes add columns but no
information, so they only hurt conditioning) and **keep the *irreducible* redundancy** (bands +
their ratio-indices are intentionally near-collinear; that residual is what the elastic-net
**ridge** component, $\alpha<1$, is designed to absorb). The leftover knob — `glmnet`'s
convergence tolerance — turns out to be the single biggest speed lever (§[Fitting]).

---

# Part III — The collinearity-reduction protocol

Everything below executes on the full data. The protocol, in order:

1. **Focal mains**: pairwise correlation + univariate separability → keep 12 → drop MIRBI and
   the tasseled-cap trio as **exact linear combinations** → **11 focal mains**.
2. **Prev-year mains**: pairwise correlation → drop the duplicated visible bands → **32**.
3. **Focal × focal interactions**: prune on **centered-factor** products → **22**.
4. **Prev × focal interactions**: keep the curated B4–B6 blocks, restricted to median + sd →
   **64**.
5. **Multivariate VIF** + the **intentional near-rank-deficiency** (eigenvalues of the
   bands + ratio-indices block), the redundancy we keep on purpose.

## Setup

```{r setup}
suppressPackageStartupMessages({
  library(data.table)
  library(corrplot)   # clustered correlation heatmaps
})

# Resolve repo root whether the working dir is repo root or collection-01/notebooks.
find_root <- function() {
  for (p in c(".", "..", "../..", "../../..")) {
    if (file.exists(file.path(p, "collection-01", "config", "veg_fire_remap.csv")))
      return(normalizePath(p))
  }
  stop("Could not locate repo root (collection-01/config/veg_fire_remap.csv not found).")
}
ROOT     <- find_root()
VERSION  <- "1"
REGIONS  <- c("BA", "CHACO", "PAMPA", "CUYO", "PAT")
obs_csv  <- function(reg) file.path(ROOT, "collection-01", "data",
              sprintf("training_observations_%s_v%s.csv", reg, VERSION))

# The full 21-focal / 40-prev canonical-team raw set — the STARTING point we reduce from.
FOCAL21 <- c("BLUE","GREEN","RED","NIR","SWIR1","SWIR2","NBR","NBR2","MIRBI","NDVI",
             "TCB","TCG","TCW","NDMI","NDSI","SAVI","NDWI","AFRI","kNDVI","EVI2","NIRv") # 21
PREV_VARS10 <- c("blue","green","red","nir","swir1","swir2","ndvi","ndwi","npv","ndfi") # 10
SUMM4       <- c("median", "median_wet", "median_dry", "stdDev")                         # 4
PREV40 <- as.vector(t(outer(PREV_VARS10, SUMM4,
            function(v, s) sprintf("mb_mos_%s_%s", v, s))))                              # 40
stopifnot(length(FOCAL21) == 21, length(PREV40) == 40)

COR_HI <- 0.90        # |r| threshold for "redundant" pair reporting / dropping
```

```{r load-data}
# Read only the 61 raw columns (+ region) from all five tables; row-bind. The single slow step.
keep_cols <- c("region", FOCAL21, PREV40)
dt <- rbindlist(lapply(REGIONS, function(reg) {
  message("  reading ", basename(obs_csv(reg)), " ...")
  fread(obs_csv(reg), select = keep_cols)
}), use.names = TRUE)

cat(sprintf("Loaded %s obs x %d raw vars (%d focal + %d prev), regions: %s\n",
            format(nrow(dt), big.mark = ","), length(FOCAL21) + length(PREV40),
            length(FOCAL21), length(PREV40), paste(REGIONS, collapse = ", ")))
```

```{r na-report}
# Non-finite audit per group — high-NA columns distort any correlation.
na_summary <- function(cols) {
  s <- dt[, lapply(.SD, function(x) sum(!is.finite(x))), .SDcols = cols]
  data.table(var = cols, n_nonfinite = as.integer(unlist(s)),
             pct = round(100 * as.integer(unlist(s)) / nrow(dt), 3))[order(-n_nonfinite)]
}
cat("Focal group — non-finite counts (top 8):\n");        print(head(na_summary(FOCAL21), 8))
cat("\nPrevious-year group — non-finite counts (top 8):\n"); print(head(na_summary(PREV40), 8))
```

### Correlation helpers (big-data friendly)

`cor_fast` standardizes columns once and forms `crossprod / (n-1)` — a single BLAS matmul, far
faster than per-pair `cor()` at millions of rows. Rows with any non-finite value in the group are
dropped (complete-case *within the group*). `cor_spearman` ranks each column first (robust to
the indices' non-linearity). `find_redundant` is the iterative pairwise pruner from
`caret::findCorrelation`, reimplemented.

```{r cor-helpers}
cor_fast <- function(M) {
  ok <- is.finite(rowSums(M)); M <- M[ok, , drop = FALSE]
  attr_n <- nrow(M); M <- scale(M)
  C <- crossprod(M) / (attr_n - 1)
  C[!is.finite(C)] <- NA
  attr(C, "n_used") <- attr_n
  C
}
cor_spearman <- function(M) {
  R <- apply(M, 2, function(x) frank(x, ties.method = "average", na.last = "keep"))
  cor_fast(R)
}
high_pairs <- function(C, thresh = 0) {
  ut <- which(upper.tri(C), arr.ind = TRUE)
  d  <- data.table(var1 = rownames(C)[ut[, 1]], var2 = colnames(C)[ut[, 2]], r = C[ut])
  d  <- d[is.finite(r)][abs(r) >= thresh][order(-abs(r))]
  d[, r := round(r, 3)][]
}
# Display-only sanitizer for corrplot: clamp tiny floating-point overshoot back into [-1,1]
# and zero out any non-finite cell (e.g. a near-zero-variance mosaic column). Does NOT touch
# the matrices used for the numeric tables above.
for_plot <- function(C) { C <- round(C, 6); C[!is.finite(C)] <- 0; pmax(pmin(C, 1), -1) }
# Greedy redundancy pruning. protect: never dropped. score: higher = keep (drops the
# lower-scoring member of the worst pair); if NULL, drops the higher-mean-|r| (globally
# more-redundant) member.
find_redundant <- function(C, thresh, protect = character(0), score = NULL) {
  A <- abs(C); diag(A) <- NA; drop <- character(0)
  repeat {
    keep <- setdiff(rownames(A), drop)
    sub  <- A[keep, keep, drop = FALSE]
    if (!any(sub > thresh, na.rm = TRUE)) break
    mx   <- which(sub == max(sub, na.rm = TRUE), arr.ind = TRUE)[1, ]
    cand <- keep[c(mx[1], mx[2])]; free <- setdiff(cand, protect)
    if (length(free) == 0L) { A[cand[1], cand[2]] <- A[cand[2], cand[1]] <- NA; next }
    if (length(free) == 1L) { drop <- c(drop, free); next }
    loser <- if (!is.null(score)) free[which.min(score[free])]
             else free[which.max(c(mean(sub[free[1], ], na.rm = TRUE),
                                   mean(sub[free[2], ], na.rm = TRUE)))]
    drop <- c(drop, loser)
  }
  drop
}
```

## Step 1 — Focal mains (21 → 11)

```{r focal-matrix}
Mf  <- as.matrix(dt[, ..FOCAL21])
Cf_p <- cor_fast(Mf); Cf_s <- cor_spearman(Mf)
cat(sprintf("Focal Pearson computed on %s complete-case rows.\n",
            format(attr(Cf_p, "n_used"), big.mark = ",")))
```

```{r focal-heatmap, fig.width=9, fig.height=9}
corrplot(for_plot(Cf_p), method = "color", type = "upper", order = "hclust",
         addCoef.col = "grey25", number.cex = 0.55, tl.cex = 0.8, tl.col = "black",
         title = "Focal features (21) — Pearson, hclust order", mar = c(0, 0, 1.5, 0))
```

```{r focal-high-pairs}
cat("Focal — Pearson |r| >= ", COR_HI, ":\n", sep = ""); print(high_pairs(Cf_p, COR_HI))
cat("\nFocal — Spearman |r| >= ", COR_HI, " (rank-based):\n", sep = ""); print(high_pairs(Cf_s, COR_HI))
```

### Univariate separability (burned vs unburned)

How well does each index *alone* separate burned from unburned? Headline is the **AUC**
(Mann–Whitney, threshold-free): 0.5 = none, 1 = perfect; `sep` $= |\text{AUC}-0.5|\cdot 2$. This
needs the `burned` label (absent from `dt`), so it does a fresh lightweight full read of
`burned` + the 21 focal columns.

```{r separability}
sep_cols <- c("burned", FOCAL21)
sepdt <- rbindlist(lapply(REGIONS, function(reg) {
  message("  reading burned+focal from ", basename(obs_csv(reg)), " ...")
  fread(obs_csv(reg), select = sep_cols)
}), use.names = TRUE)

auc1 <- function(x, y) {
  ok <- is.finite(x); x <- x[ok]; y <- y[ok]
  n1 <- as.numeric(sum(y == 1L)); n0 <- as.numeric(sum(y == 0L))
  if (n1 == 0 || n0 == 0) return(NA_real_)
  r <- frank(x, ties.method = "average")
  (sum(r[y == 1L]) - n1 * (n1 + 1) / 2) / (n1 * n0)
}
yb  <- as.integer(sepdt$burned)
sep <- rbindlist(lapply(FOCAL21, function(f) {
  x  <- sepdt[[f]]; ok <- is.finite(x)
  a  <- auc1(x, yb)
  mb <- mean(x[ok & yb == 1L]); mu <- mean(x[ok & yb == 0L])
  vb <- var(x[ok & yb == 1L]);  vu <- var(x[ok & yb == 0L])
  n1 <- sum(ok & yb == 1L);     n0 <- sum(ok & yb == 0L)
  sp <- sqrt(((n1 - 1) * vb + (n0 - 1) * vu) / (n1 + n0 - 2))
  data.table(var = f, auc = a, sep = abs(a - 0.5) * 2,
             cohens_d = (mb - mu) / sp, mean_burned = mb, mean_unburned = mu)
}))
sep[, c("auc","sep","cohens_d","mean_burned","mean_unburned") :=
      .(round(auc,4), round(sep,4), round(cohens_d,3), round(mean_burned,4), round(mean_unburned,4))]
setorder(sep, -sep); print(sep)
rm(sepdt); invisible(gc())
```

### Automatic pruner (protect raw optical + NBR/NBR2; tiebreak by separability)

```{r focal-decision}
KEEP_FOCAL <- c("NBR", "NBR2", "BLUE", "GREEN", "RED", "NIR", "SWIR1", "SWIR2")  # always keep
sep_score  <- setNames(sep$sep, sep$var)
drop_focal <- find_redundant(Cf_p, COR_HI, protect = KEEP_FOCAL, score = sep_score)
A <- abs(Cf_p); diag(A) <- NA
focal_summary <- data.table(
  var = FOCAL21, separability = round(sep_score[FOCAL21], 4),
  mean_abscor = round(rowMeans(A, na.rm = TRUE), 3),
  max_abscor  = round(apply(A, 1, max, na.rm = TRUE), 3))
focal_summary[, status := fifelse(var %in% KEEP_FOCAL, "protected",
                          fifelse(var %in% drop_focal, "DROP", "keep"))]
setorder(focal_summary, status, -separability); print(focal_summary)
cat("\nPairwise pruner DROP:", paste(drop_focal, collapse = ", "), "\n")
```

### The exact-linear-combination cut (MIRBI + tasseled-cap)

The pairwise pruner is blind to **multivariate** redundancy: it compares two columns at a time,
so it cannot see that a variable is an exact linear combination of *several* protected ones. Two
groups of focal features are exactly that, and they leave the design even though their pairwise
$|r|$ never trips the 0.9 rule:

- **MIRBI** $= 10\,\text{SWIR2} - 9.8\,\text{SWIR1} + 2$ — exactly linear in two bands we keep.
- **TCB, TCG, TCW** — exact linear combinations of the six raw optical bands (Baig 2014).

With the raw bands protected, a linear model (and the band×· product terms) already **spans**
these indices: regressing any of them on the bands gives $R^2 \approx 1$. They add columns and
worsen conditioning but contribute no new direction. We verify the exactness two ways — an
$R^2$ from the regression, and the **smallest eigenvalue of the relevant correlation submatrix**,
which is the cleanest fingerprint of rank-deficiency (an exact linear dependence forces an
**exactly zero** eigenvalue).

```{r exact-combo}
r2_on <- function(y, xs) {                      # R^2 of y regressed on columns xs (complete cases)
  M <- as.matrix(dt[, c(y, xs), with = FALSE]); M <- M[is.finite(rowSums(M)), ]
  summary(lm(M[, 1] ~ M[, -1]))$r.squared
}
smallest_eig <- function(vars) {                # min eigenvalue of the correlation matrix
  C <- cor_fast(as.matrix(dt[, ..vars])); min(eigen(C, symmetric = TRUE)$values)
}
exact <- data.table(
  index   = c("MIRBI", "TCB", "TCG", "TCW"),
  spanned_by = c("SWIR1+SWIR2", "6 optical", "6 optical", "6 optical"),
  R2      = c(r2_on("MIRBI", c("SWIR1","SWIR2")),
              r2_on("TCB", c("BLUE","GREEN","RED","NIR","SWIR1","SWIR2")),
              r2_on("TCG", c("BLUE","GREEN","RED","NIR","SWIR1","SWIR2")),
              r2_on("TCW", c("BLUE","GREEN","RED","NIR","SWIR1","SWIR2"))))
exact[, R2 := round(R2, 6)]; print(exact)
cat(sprintf("\nmin eigenvalue of cor(SWIR1, SWIR2, MIRBI)      = %.3e  (exact dependence -> ~0)\n",
            smallest_eig(c("SWIR1","SWIR2","MIRBI"))))
cat(sprintf("min eigenvalue of cor(6 optical bands + TCB,TCG,TCW) = %.3e  (3 exact deps -> ~0)\n",
            smallest_eig(c("BLUE","GREEN","RED","NIR","SWIR1","SWIR2","TCB","TCG","TCW"))))
```

$R^2 = 1$ (to floating-point) and a correlation eigenvalue at machine-epsilon confirm exact
linear dependence: **drop MIRBI, TCB, TCG, TCW**. The ash/brightness signal TCB was prized for is
not lost — TCB *is* a fixed weighted sum of the six protected bands, so the model retains full
access to that contrast through the bands and their products.

### Focal decision — the reduced 11

Combining the pruner with two manual rules (protect one greenness + one moisture index; prefer
cheap normalized-difference indices over redundant ones) and the exact-combo cut:

**KEEP (11):** `BLUE GREEN RED NIR SWIR1 SWIR2` (raw optical) · `NBR NBR2` (burn) ·
**`NDVI`** (greenness) · **`NDMI`** (moisture) · `NDSI` (the one genuinely independent extra
axis, max $|r|\approx0.54$).

**DROP (10):** `MIRBI TCB TCG TCW` (exact linear combos) · `AFRI kNDVI SAVI EVI2 NIRv` (greenness/
burn cluster, redundant with NDVI/NBR) · `NDWI` (McFeeters NDWI $\approx -$NDVI, $r=-0.977$: a
greenness axis in disguise, not independent moisture).

Rationale highlights:

- **Greenness = NDVI, not kNDVI**: $r(\text{NDVI},\text{kNDVI})=0.986$, equal separability; NDVI
  is canonical/cheap and kNDVI's nonlinearity is recoverable from product terms.
- **Moisture = NDMI, not NDWI/TCW**: NDMI is cheaper than TCW, separates better, and
  $r(\text{NDMI},\text{TCW})=0.86$ so it covers TCW. We accept one strongly-collinear pair,
  $r(\text{NDMI},\text{NBR})=0.966$ — the useful moisture/ash signal is NDMI's component
  *orthogonal* to NBR (ash test: NDMI sep 0.73 vs NBR 0.45), and there is no cheap NIR–SWIR
  moisture index decorrelated from NBR (they share NIR). **This deliberately-kept pair is part of
  the irreducible near-rank-deficiency analysed in §[VIF].**

```{r focal-final}
FOCAL_KEEP <- c("BLUE","GREEN","RED","NIR","SWIR1","SWIR2","NBR","NBR2","NDVI","NDMI","NDSI") # 11
stopifnot(length(FOCAL_KEEP) == 11)
rm(Mf, Cf_p, Cf_s, A); invisible(gc())
cat("Focal mains kept (11):", paste(FOCAL_KEEP, collapse = ", "), "\n")
```

## Step 2 — Previous-year mains (40 → 32)

```{r prev-matrix}
Mp  <- as.matrix(dt[, ..PREV40]); Cp_p <- cor_fast(Mp)
cat(sprintf("Prev-year Pearson computed on %s complete-case rows.\n",
            format(attr(Cp_p, "n_used"), big.mark = ",")))
```

```{r prev-heatmap, fig.width=12, fig.height=12}
corrplot(for_plot(Cp_p), method = "color", type = "upper", order = "hclust", tl.cex = 0.6, tl.col = "black",
         title = "Previous-year mosaic (40) — Pearson, hclust order", mar = c(0, 0, 1.5, 0))
```

```{r prev-high-pairs}
cat("Prev-year — Pearson |r| >= ", COR_HI, ":\n", sep = ""); print(high_pairs(Cp_p, COR_HI))
```

The prev-year set is far **less collinear than focal** — essentially all $>0.9$ redundancy is one
structure: the **visible trio** (blue ≈ green ≈ red) are near-duplicates within *every* summary
($r\approx0.93$–$0.99$). The median/wet/dry/sd summaries of every other variable are distinct
enough to keep.

**Decision — keep `green` as the single visible representative (all 4 summaries) and drop `blue`
and `red` entirely (8 columns):** 8 base vars (`green, nir, swir1, swir2, ndvi, ndwi, npv, ndfi`)
× 4 summaries = **32 main terms**. These are cheap, only mildly collinear main effects; we keep
them generously. (Which prev bands enter the expensive *interaction* blocks is decided
separately, in Step 4.)

```{r prev-final}
PREV_DROP_VARS <- c("blue", "red")
PREV_MAIN <- PREV40[!grepl(sprintf("mb_mos_(%s)_", paste(PREV_DROP_VARS, collapse = "|")), PREV40)]
stopifnot(length(PREV_MAIN) == 32)
rm(Mp, Cp_p); invisible(gc())
cat("Prev mains kept (32):\n"); print(PREV_MAIN)
```

## Step 3 — Focal × focal interactions (210 → 22)

With 11 focal mains the pairwise block is $\binom{11}{2}=55$ candidates (down from
$\binom{21}{2}=210$). We keep an interaction only if it adds structure beyond the mains and the
other kept interactions — pruning by correlation with the **11 mains protected**.

**Products are built from CENTERED factors**, $(X-\bar X)(Y-\bar Y)$, not raw $X\!\cdot\!Y$ — the
only honest way to judge interaction redundancy, since a raw product is *mechanically* correlated
with its parents. Centering the *factors* genuinely changes the column:
$(X-a)(Y-b) = XY - aY - bX + ab$.

> **Is this just reparameterization?** Only without a penalty. In an unpenalized model that also
> holds $X$, $Y$ and an intercept, raw and centered products span the same space → identical
> predictions. **Under glmnet's penalty it is a different model**: the penalty is on the
> coefficients, which differ between parameterizations for the same surface, so the penalized
> optimum (selection + shrinkage) differs — in the better-conditioned direction. For
> **deployment** the distinction vanishes: once fitted, fold the centering constants into the
> intercept + main slopes, and GEE deploys raw band products (no means vector).

```{r focal-interactions}
OPT_K <- c("BLUE","GREEN","RED","NIR","SWIR1","SWIR2")
IDX_K <- c("NBR","NBR2","NDVI","NDMI","NDSI")
Mk <- as.matrix(dt[, ..FOCAL_KEEP]); Mk <- Mk[is.finite(rowSums(Mk)), ]
Mc <- sweep(Mk, 2, colMeans(Mk), "-")                    # CENTER factors before multiplying
pr <- combn(FOCAL_KEEP, 2)                               # 55 distinct pairs
P  <- vapply(seq_len(ncol(pr)), function(k) Mc[, pr[1, k]] * Mc[, pr[2, k]], numeric(nrow(Mc)))
colnames(P) <- apply(pr, 2, paste, collapse = ":")
Cint     <- cor_fast(cbind(Mk[, FOCAL_KEEP], P))         # raw mains + centered products
drop_int <- find_redundant(Cint, COR_HI, protect = FOCAL_KEEP)
keep_int_pw <- setdiff(colnames(P), drop_int)
blk <- function(t) { ab <- strsplit(t, ":")[[1]]
  if (all(ab %in% OPT_K)) "opt:opt" else if (all(ab %in% IDX_K)) "idx:idx" else "opt:idx" }
itab <- data.table(term = colnames(P), block = vapply(colnames(P), blk, ""),
                   status = fifelse(colnames(P) %in% keep_int_pw, "keep", "DROP"))
cat(sprintf("Pairwise prune: kept %d / 55  (|r| > %.2f, 11 mains protected, centered factors)\n",
            length(keep_int_pw), COR_HI))
print(itab[, .(keep = sum(status == "keep"), total = .N), by = block][order(block)])
rm(Mk, Mc, P, Cint); invisible(gc())
```

The pairwise prune keeps **25** of 55. A final **VIF-on-mains** screen (§[VIF]) then looks for the
ones still reproducible from the *main span* (a multivariate redundancy pairwise correlation
misses). At a strict $R^2_{\text{on mains}}>0.9$ that screen flags 6 (printed in §[VIF]); but none
of them exceed $R^2=0.95$, so the production design makes the **soft call at ~0.95** and drops only
the three clearest (`SWIR1:NBR2, SWIR2:NBR, NIR:NDVI`), keeping the three borderline ones
(`NIR:NBR, SWIR1:NDMI, BLUE:RED`). That yields the **22** focal interactions frozen in
`02-model_fitting.R` (the authoritative set — a stricter 0.9 cut would trim 3 more, to 19):

```{r focal-int-frozen}
FOCAL_INT <- c("BLUE:NBR","BLUE:NBR2","BLUE:NDMI","BLUE:RED","BLUE:SWIR1","GREEN:NIR",
               "NBR:NBR2","NBR:NDMI","NBR2:NDSI","NBR2:NDVI","NDMI:NDSI","NDVI:NDMI",
               "NDVI:NDSI","NIR:NBR","NIR:NDSI","NIR:SWIR1","NIR:SWIR2","SWIR1:NDMI",
               "SWIR1:NDSI","SWIR1:NDVI","SWIR1:SWIR2","SWIR2:NDVI")                  # 22
stopifnot(length(FOCAL_INT) == 22)
cat("Frozen focal interactions (22), used in fitting:\n"); print(FOCAL_INT)
```

This is **22 vs the old 210** — the single biggest term-count cut.

## Step 4 — Prev × focal interactions (curated B4–B6, median + sd → 64)

For prev × focal we keep the **curated block structure** rather than crossing everything: the
blocks encode a *relevance* prior — a band interacts with its own previous-year level (same-band),
and previous-year vegetation/fraction state modulates the fire signal — which a pure correlation
prune would discard (it removes *redundant* terms, not *useless* ones). Two reductions vs the
canonical-team blocks: each block uses the **median + sd** summaries only (dropping
$\text{wet}\approx\text{med}$ and the partly-redundant $\text{dry}$ from the interactions), and
block membership is aligned to the reduced main sets (11 focal, 32 prev).

| block | prev side (× med + sd) | × focal side | count |
|---|---|---|---:|
| **B4 same-band** | green, nir, swir1, swir2, ndvi | each × its own focal band | 5 × 2 = **10** |
| **B5 idx × fire-idx** | ndvi, ndwi, npv, ndfi | × {NBR, NBR2, NDVI}; skip ndvi×NDVI | 4×3×2 − 2 = **22** |
| **B6 idx × fire-band** | ndvi, ndwi, npv, ndfi | × {RED, NIR, SWIR1, SWIR2} | 4×4×2 = **32** |
| **total** | | | **64** |

These blocks are **not** correlation-pruned among themselves — they are already a curated, sparse
set. Their multivariate collinearity against the mains is checked next.

## Step 5 — Multivariate collinearity (VIF) and the redundancy we keep on purpose {#sec-vif}

Pairwise correlation can't see when a term is a linear combination of *several* others; VIF can:
$\text{VIF}_j = 1/(1 - R^2_j)$, with $R^2_j$ from regressing term $j$ on all others ($>5$
moderate, $>10$ serious). It is **n-invariant** (a function of the correlation matrix only).

**The caveat that dominates here, and your question about the bands + ratio-indices.** We
deliberately keep the 6 raw optical bands *and* indices that are (near-)functions of them: NBR,
NBR2, NDMI are **ratios** of NIR/SWIR pairs. That makes the protected main set **intentionally
near-rank-deficient** — `cor(X)` is numerically near-singular, and naïve `diag(solve(cor(X)))`
returns garbage (VIFs in the millions). That collinearity is **not** a bug and is **not** removed
by deleting terms; it is exactly what the **elastic-net ridge component ($\alpha<1$) absorbs**.

### How you *see* near-rank-deficiency: eigenvalues of the correlation matrix

The clean diagnostic is the eigenvalue spectrum of the correlation matrix of the suspect block.
For a set of $p$ standardized variables, $\sum\lambda_i = p$. An **exact** linear dependence
forces a $\lambda_i = 0$; a **near** dependence forces a $\lambda_i \approx 0$. The condition
number $\kappa = \lambda_{\max}/\lambda_{\min}$ measures how close to singular the block is
($\kappa \gtrsim 30$ is the usual flag; $\kappa \gtrsim 1000$ is severe).

Take the block you asked about — **{NBR, NBR2, NIR, SWIR1, SWIR2}**. NBR and NBR2 are ratios
built from {NIR, SWIR1, SWIR2}, so five variables live in essentially **three** underlying
dimensions: we expect the two smallest eigenvalues to collapse toward zero and $\kappa$ to be
large.

```{r eig-near-rank}
demo <- c("NBR","NBR2","NIR","SWIR1","SWIR2")
Cd   <- cor_fast(as.matrix(dt[, ..demo]))
ed   <- eigen(Cd, symmetric = TRUE)$values
print(data.table(component = seq_along(ed),
                 eigenvalue = round(ed, 5),
                 pct_var = round(100 * ed / sum(ed), 2),
                 cum_pct = round(100 * cumsum(ed) / sum(ed), 2)))
cat(sprintf("\nsum of eigenvalues = %.2f (= p = %d)\n", sum(ed), length(ed)))
cat(sprintf("lambda_max = %.4f,  lambda_min = %.3e\n", max(ed), min(ed)))
cat(sprintf("condition number  kappa = lambda_max / lambda_min = %.1f\n", max(ed) / min(ed)))
```

Read the printed spectrum (full-data values): the first two components carry **~95%** of the
variance, and the smallest eigenvalue collapses to $\lambda_{\min}\approx0.013$ — about **240×
below** $\lambda_{\max}\approx3.27$, giving $\kappa\approx243$. That is well past the
$\kappa\gtrsim30$ collinearity flag: five variables effectively live in ~two dominant directions.
This is *near*-rank-deficiency (ill-conditioning), **not** the exact singularity of
§[exact-combo]: there MIRBI / the tasseled-caps drove a correlation eigenvalue to *machine
epsilon* ($\approx10^{-13}$, an exactly-zero direction) and were removed; here the ratio-indices
are *non-linear* functions of the bands, so the small eigenvalue is small-but-finite — a block we
**keep** and condition with ridge rather than delete.

**Why keep a near-singular main block at all?** Because (a) the redundancy is *not exact* — the
ratio-indices carry a small but real component orthogonal to the bands (the ash/moisture signal in
NDMI⊥NBR is the worked example in Step 1), and (b) dropping a band or a ratio to "fix" the
condition number would throw that orthogonal component away. The elastic-net **ridge** term
($\alpha<1$) shrinks along the near-null directions instead of deleting them: it keeps the model
identifiable and the fit well-posed without discarding signal. This is precisely why the final
$\alpha$ grid (§[Fitting]) **excludes pure lasso** ($\alpha=1$) — some ridge is needed to
condition this block.

### VIF used only for the question it can answer: do interactions add anything beyond the mains?

Because the mains are near-singular by design, we use VIF only via a **pseudo-inverse** to ask:
what is the $R^2$ of each interaction regressed on the **main span**? (Well-defined even when the
mains are collinear.) Mains are never dropped; only interactions are screened.

```{r design-vif}
PREV_SUMM2 <- c("median","stdDev")
SAMEBAND_K <- c("green","nir","swir1","swir2","ndvi")
CONTEXT_K  <- c("ndvi","ndwi","npv","ndfi")
FIRE_IDX   <- c("NBR","NBR2","NDVI"); FIRE_BAND <- c("RED","NIR","SWIR1","SWIR2")

need <- unique(c(FOCAL_KEEP, PREV_MAIN))
D  <- as.matrix(dt[, ..need]); D <- D[is.finite(rowSums(D)), ]
Dc <- sweep(D, 2, colMeans(D), "-")                              # centered factors

fi  <- strsplit(keep_int_pw, ":")                               # pairwise-kept focal interactions
Xfi <- vapply(fi, function(ab) Dc[, ab[1]] * Dc[, ab[2]], numeric(nrow(Dc)))
colnames(Xfi) <- keep_int_pw

specs <- list()                                                  # prev x focal blocks (median+sd)
for (v in SAMEBAND_K) for (s in PREV_SUMM2)
  specs <- c(specs, list(c(sprintf("mb_mos_%s_%s", v, s), toupper(v))))
for (v in CONTEXT_K)  for (s in PREV_SUMM2) for (f in FIRE_IDX)
  if (!(v == "ndvi" && f == "NDVI")) specs <- c(specs, list(c(sprintf("mb_mos_%s_%s", v, s), f)))
for (v in CONTEXT_K)  for (s in PREV_SUMM2) for (f in FIRE_BAND)
  specs <- c(specs, list(c(sprintf("mb_mos_%s_%s", v, s), f)))
Xpf <- vapply(specs, function(z) Dc[, z[1]] * Dc[, z[2]], numeric(nrow(Dc)))
colnames(Xpf) <- vapply(specs, function(z) paste(z[1], z[2], sep = ":"), "")

X <- cbind(D[, c(FOCAL_KEEP, PREV_MAIN)], Xfi, Xpf)
mains <- c(FOCAL_KEEP, PREV_MAIN); ints <- setdiff(colnames(X), mains)
C  <- cor(X); eA <- eigen(C, symmetric = TRUE)
rk <- function(M){ e <- eigen(M, symmetric = TRUE)$values; sum(e > max(e) * 1e-8) }
cat(sprintf("Assembled design: %d terms (%d main + %d interaction)\n",
            ncol(X), length(mains), length(ints)))
cat(sprintf("MAIN block rank: %d/%d  |  FULL rank: %d/%d  |  recip-cond %.1e\n",
            rk(C[mains, mains]), length(mains),
            sum(eA$values > max(eA$values)*1e-8), ncol(X), min(eA$values)/max(eA$values)))

ev   <- eigen(C[mains, mains], symmetric = TRUE); tol <- max(ev$values) * 1e-8
pinv <- ev$vectors %*% ((ifelse(ev$values > tol, 1/ev$values, 0)) * t(ev$vectors))
R2   <- vapply(ints, function(j){ cc <- C[j, mains]
          min(max(as.numeric(t(cc) %*% pinv %*% cc), 0), 1) }, 0.0)
vift <- data.table(term = ints, R2_on_mains = round(R2, 3),
                   vif = round(1/(1 - pmin(R2, 1 - 1e-12)), 1))[order(-R2_on_mains)]
print(vift[1:12])
drop_vif <- vift[R2_on_mains > 0.9, term]
cat(sprintf("\nVIF drop (R^2_on_mains > 0.9 / VIF > 10): %d interactions: %s\n",
            length(drop_vif), paste(drop_vif, collapse = ", ")))
rm(D, Dc, Xfi, Xpf, X, C, eA); invisible(gc())
```

**Reading it (full-data values).** The MAIN block is now **full rank, 43/43** — dropping the
exact-combo indices (MIRBI, TCB/TCG/TCW) *restored* rank that the canonical-team mains lacked — yet
it stays **ill-conditioned**, the κ≈243 bands + ratio-indices block above. So the residual here is
*conditioning*, not rank-deficiency, and it is left to the ridge term, not removed. The full
assembled design printed as **132 terms** (43 mains + 89 interactions = the 25 pairwise-kept focal
+ 64 prev×focal) is rank **131/132** — one *interaction* is linearly dependent (recip-cond
$\sim10^{-16}$), which is exactly what the VIF screen catches. At $R^2_{\text{on mains}}>0.9$ it
flags **6** focal opt×idx pairs (`NIR:NBR, SWIR1:NBR2, SWIR1:NDMI, SWIR2:NBR, BLUE:RED, NIR:NDVI`)
that survived the pairwise prune but are multivariately ≥90% reproducible from the mains. None
exceed $R^2=0.95$; production drops the three clearest and keeps three borderline (see Step 3),
taking the focal interactions to **22** and the design to its final **129** (43 mains + 22 focal +
64 prev×focal).

---

# Part IV — The final design and fitting config {#sec-fitting}

## Final design — before vs after

| component | canonical-team (427) | reduced (129) |
|---|--:|--:|
| focal mains | 21 | **11** |
| prev-year mains | 40 | **32** |
| focal × focal | 210 | **22** |
| prev × focal (B4–B6) | 156 | **64** |
| **total terms (+ intercept)** | **427** | **129** |

A **~70% cut** (427 → 129), concentrated in the interaction blocks — exactly where coordinate
descent was crawling. The block structure used in fitting (`workflow/02-model_fitting.R`):

```{r blocks}
BLOCKS <- c(focal = 11, prev = 32, pairs = 22, sameband = 10, cross_idx = 22, cross_band = 32)
stopifnot(sum(BLOCKS) == 129)
print(as.data.table(as.list(BLOCKS))[, total := sum(BLOCKS)][])
```

## Fitting config — the convergence fix was numerical

The reduction removes the *removable* collinearity; the leftover (intentional) collinearity is
handled at fit time. The decisive findings, all in `02-model_fitting.R`:

- **`thresh = 1e-4`** (relaxed from glmnet's `1e-7` default) — **the single biggest speed lever.**
  The default tolerance forces coordinate descent to grind out spurious precision in the
  ill-conditioned low-$\lambda$ tail; `1e-4` is far below the noise floor of a CV-selected model
  and converges in a fraction of the iterations. A per-class adaptive fallback loosens further
  (×5, up to `1e-2`) under a wall-clock budget, so no class can stall.
- **`nlambda = 50`, `lambda.min.ratio = 1e-4`** — a finer $\lambda$ path gives better warm starts
  (each fit initialized from the previous $\lambda$); the CV optimum sits near the path floor, so
  we go deep.
- **$\alpha$ grid $\{0.25, 0.5, 0.75\}$** — pure ridge ($\alpha=0$) and pure lasso ($\alpha=1$)
  are both dropped: ridge was never best in CV and is slowest, and lasso lacks the ridge component
  that **conditions the intentionally near-singular main block** (§[VIF]). The interior keeps that
  component. Selected at `lambda.min` (data-rich, near-MLE wanted).
- **Centered-factor interactions, folded back at export** — products are fit from mean-centered
  factors (better-conditioned under the penalty); after fitting, the centering is folded into the
  intercept + main slopes so the **exported** coefficients are on the raw-product scale and GEE
  deploys raw band multiplies (no means vector). Verified numerically (predictions identical) on
  every fit.
- **CV design** — grouped K-fold ($K=10$, group = region-unique fire id, so whole fires are held
  out and within-fire temporal autocorrelation cannot leak), folds packed by stratified greedy
  balancing of obs + positives, pure-negative fires point-distributed across folds, tuned on
  binomial deviance. Out-of-fold $p_i$ saved per observation.

With the reduced, conditioned design and the relaxed tolerance, fits that previously ran for hours
per $\alpha$ complete in minutes — confirming the original problem was **numerical
(ill-conditioning), not statistical**.

## GEE computational considerations {#sec-gee}

The model is evaluated for every Landsat observation in Argentina, 1999–2025. The implementation
keeps the GEE compute graph at roughly **15 nodes regardless of term count** by expressing each
interaction *block* as one `.multiply()` between two parallel multiband images, rather than
looping per term (which would chain O(n) server-side nodes).

```js
// ---- build the two source images with distinct band names ----
var focal = computeFocal(landsat).rename(focalNames);   // 11 bands, e.g. "NBR_t"
var prev  = mbMosaic.select(prevBandsRaw).rename(prevNames); // 32 bands, e.g. "NDVI_med"
var X0    = focal.addBands(prev);                        // 43 mains

// ---- each interaction block in ONE multiply (parallel band lists) ----
var pairs = X0.select(pairs_a).multiply(X0.select(pairs_b)).rename(pairs_names);     // 22
var same  = X0.select(same_a ).multiply(X0.select(same_b )).rename(same_names);      // 10
var cidx  = X0.select(cidx_a ).multiply(X0.select(cidx_b )).rename(cidx_names);      // 22
var cband = X0.select(cband_a).multiply(X0.select(cband_b)).rename(cband_names);     // 32

// ---- assemble 129-band predictor, weighted sum + sigmoid ----
var X    = X0.addBands(pairs).addBands(same).addBands(cidx).addBands(cband);
var coef = ee.Image.constant(coefValues).rename(coefNames);   // 129 floats, name-matched
var eta  = X.multiply(coef).reduce(ee.Reducer.sum()).add(alpha);
var p    = eta.multiply(-1).exp().add(1).pow(-1);              // sigmoid
```

The graph size does not grow with the term count: a 500-term version would have the same node
count as a 50-term one. Band names must be globally unique (the `_t` / `_med` / `_sd` suffix
convention handles it; `addBands` throws on a clash). SMA-derived layers (NPV, NDFI) are *consumed*
as ordinary mosaic bands — never recomputed per scene — so they cost zero per-pixel array ops on
our side. Tasseled-cap, had we kept it, would be one cheap linear `.expression()`; dropping it
nonetheless removes columns that only hurt the *fit*.

**Numerical precision at prediction.** GEE uses float32 (~7 significant digits, relative). For
$|\eta|>10$ the sigmoid saturates and precision is irrelevant; near $\eta=0$ the contributions are
by definition not all large, so float32 summation error in $p$ is $\sim10^{-5}$ — negligible
against downstream smoothing/region-growing. No prediction-time rescaling required. (Optional
insurance, if ever needed: a single 32-band centering of the prev mosaic before products, with the
means stored in the coefficient JSON; one node, reused across all focal scenes via GEE's
common-subexpression dedup.)

## Summary of design choices

- **Logistic regression**: cheap at prediction, trivial in GEE, sufficient when richly
  parameterized.
- **11 focal mains + 22 focal pairs**: capture non-monotonic fire signatures and disambiguate fire
  from drought, ash, deforestation, defoliation and agriculture — the dominant Collection-0
  confusions — after dropping exact-linear-combo indices (MIRBI, tasseled-cap) that added columns
  but no information.
- **32 prev-year mains + 64 prev × focal interactions**: turn the focal classifier into a
  *context-aware* one without per-pixel temporal-change features.
- **Reduction 427 → 129**: removes the *removable* collinearity (exact dependences, duplicated
  axes) that made coordinate descent crawl; the *irreducible* near-rank-deficiency (bands + their
  ratio-indices) is kept and handled by the elastic-net ridge component.
- **Elastic net $\{0.25,0.5,0.75\}$, `lambda.min`, `thresh=1e-4`**: ridge conditions the kept
  near-singular block; the relaxed tolerance is what actually made fitting fast.
- **No deltas, no annual min/max, no $\text{dry}$/$\text{amp}$ in interactions**: absorbed by
  interactions and the logistic non-linearity, or confounded by the very signal we detect.
```

