---
title: "Model fit diagnostics"
subtitle: "Per veg_fire class — elastic-net tuning, coefficients, calibration, OOF predictions"
author: "Iván Barberá"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    embed-resources: true
    theme: cosmo
    fig-align: center
    code-fold: true
---

```{r setup, message=FALSE, warning=FALSE}
library(data.table)
library(tidyverse)
library(scales)
library(knitr)
library(patchwork)

models_dir <- "../models"                        # notebook runs from collection-01/notebooks/
theme_set(theme_minimal(base_size = 12))
block_lvls <- c("focal", "prev", "pairs", "sameband", "cross_idx", "cross_band")

# By-fire time-series panels (end of each class's "By fire" section, below).
source("../scripts/ts_plot_functions.R")
ts_cache <- readRDS(file.path(models_dir, "ts_plot_cache_v1.rds"))

# Discover every fitted class from its per-class cv_metrics sidecar, ordered by code.
metric_files <- Sys.glob(file.path(models_dir, "class_*_cv_metrics.csv"))
codes <- sort(as.integer(sub(".*class_0*(\\d+)_cv_metrics\\.csv$", "\\1", metric_files)))
stopifnot(length(codes) > 0)
```

This report renders one top-level section per **fitted veg_fire class** (discovered
from `models/class_*_cv_metrics.csv`), each with the same subsections: summary,
elastic-net tuning, coefficients, calibration, OOF probability distribution, and a
by-fire breakdown. The per-class plotting code lives in the child template
`_model_fit_diagnostics_child.qmd`. Currently fitted: **`r length(codes)`** class(es).

```{r render-classes, results='asis', message=FALSE, warning=FALSE}
out <- character()
for (CLASS_CODE in codes) {
  tag <- sprintf("class_%02d", CLASS_CODE)
  rd  <- function(suffix) fread(file.path(models_dir, sprintf("%s_%s", tag, suffix)))
  metrics <- rd("cv_metrics.csv")
  tuning  <- rd("tuning.csv")
  coefs   <- rd("coefficients.csv")
  oof     <- rd("oof_predictions.csv")
  cls     <- metrics$veg_fire_name[1]
  out <- c(out, knitr::knit_child("_model_fit_diagnostics_child.qmd",
                                  envir = environment(), quiet = TRUE))
}
cat(out, sep = "\n")
```
