---
title: "Object size and model uncertainty — where to cut the minimum fire size"
subtitle: "Step 06 · all fire-years · collection-01"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: true
    fig-width: 9
    fig-height: 5
---

```{r setup}
#| message: false
#| warning: false
knitr::opts_chunk$set(message = FALSE, warning = FALSE, dev = "png", dpi = 120)
# Every step-06 path is written from the repo ROOT. knitr restores the working directory after
# each chunk, so a setwd() in a later chunk would silently apply to that chunk only (and an
# unresolvable POLY_DIR yields an EMPTY table, not an error) — root.dir is the only thing that
# holds for the whole render.
knitr::opts_knit$set(root.dir = normalizePath(if (dir.exists("collection-01")) "." else "../.."))
library(data.table)
library(ggplot2)
library(patchwork)
library(knitr)
theme_set(theme_bw(base_size = 11) +
          theme(panel.grid.minor = element_blank(),
                strip.background = element_rect(fill = "grey92", colour = NA)))
```

```{r paths}
source("collection-01/scripts/objects_data_functions.R")
PRED_DIR <- "collection-01/data/objects-predictions"
VARIANT  <- "grouped"          # the deployed model (docs/06 "Predictor variant")
PX_HA    <- 0.09               # 30 m Landsat pixel
```

## Why this notebook

Two decisions hang on the size distribution:

1. **The collection's minimum mapped fire size.** Every burned-area product declares one. Ours
   is not yet chosen, and the honest basis for choosing it is not a convention borrowed from
   another product — it is *where our own model stops being able to tell fire from non-fire*.
2. **Where the mapping effort and the error live.** 61 % of objects sit in one band, and the
   thresholds already differ by size, so any statement about accuracy is meaningless without
   knowing how objects and area distribute across sizes.

Section 1–4 need only the step-05 metrics. Sections 5–7 need a `predict all` run.

```{r load-sizes}
years <- object_years()
if (!length(years)) stop("no step-05 raster_metrics CSVs under ", POLY_DIR,
                         " (wd = ", getwd(), ") — nothing to summarise")
sizes <- rbindlist(lapply(years, function(fy) {
  d <- fread(file.path(POLY_DIR, sprintf("objects_%d_raster_metrics.csv", fy)),
             select = c("oid", "n_pixels", "area_ha"))
  d[, fire_year := fy][]
}))
# set(), not `:=`: the column and the helper share a name, and inside data.table's j-scope the
# argument promise fails to resolve under knitr's eval frame
set(sizes, j = "size_class", value = size_class(sizes$area_ha))
cat(sprintf("%s objects, fire-years %d-%d, %.2f Mha total\n",
            format(nrow(sizes), big.mark = ","), min(years), max(years),
            sum(sizes$area_ha) / 1e6))
```

## 1 · How the objects and the area split by size

`size_class` uses the **display** breaks (0.5, 1, 50, 300, 1000 ha) — six classes. These are not
the four threshold bands: `<0.5` / `0.5-1` exist only because the minimum-size decision lives
inside them, and `300-1000` / `>=1000` because we want to see whether the very large objects
behave apart (the threshold work found they do not, statistically).

```{r size-table}
tab <- sizes[, .(n = .N, area_kha = sum(area_ha) / 1e3), by = size_class][order(size_class)]
tab[, `:=`(pct_n = 100 * n / sum(n), pct_area = 100 * area_kha / sum(area_kha))]
tab[, `:=`(cum_pct_n = cumsum(pct_n), cum_pct_area = cumsum(pct_area))]
kable(tab, digits = c(0, 0, 1, 2, 3, 2, 3), format.args = list(big.mark = ","),
      col.names = c("size class", "objects", "area (kha)", "% objects", "% area",
                    "cum % objects", "cum % area"),
      caption = "Objects and area by display size class, all fire-years pooled.")
```

The asymmetry between the two "%" columns is the whole story of a minimum-size cut: it buys a
large reduction in object count for a small loss of area. The `cum %` columns read directly as
"if I cut here, I discard this many objects and this much area".

```{r size-plot}
#| fig-height: 4
p1 <- ggplot(tab, aes(size_class, n)) +
  geom_col(fill = "grey35") + scale_y_log10(labels = scales::comma) +
  labs(x = NULL, y = "objects (log)", title = "Object count") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
p2 <- ggplot(tab, aes(size_class, pct_area)) +
  geom_col(fill = "firebrick") +
  labs(x = NULL, y = "% of total area", title = "Share of burned area") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
p1 + p2
```

## 2 · The pixel scale — what "<1 ha" actually means

Areas are **not** on a 0.09 ha grid. The objects carry lat/lon pixel coordinates — ~30 m *at the
equator* — and area is measured on the ellipsoid, so one pixel is `900 * cos(lat)` m²: 831 m² in
Formosa, 517 m² in southern Santa Cruz. Harmless, but it means a sub-hectare class is a pixel-count
**range**, not a fixed count, and the same 15-pixel object is `>=1 ha` in the north and `<1 ha` in
Patagonia.

```{r px-scale}
sizes[, px_ha := area_ha / n_pixels]
pxq <- sizes[, .(min = min(px_ha), q05 = quantile(px_ha, .05), median = median(px_ha),
                 q95 = quantile(px_ha, .95), max = max(px_ha))]
kable(round(pxq * 1e4, 1), col.names = c("min", "q05", "median", "q95", "max"),
      caption = "Per-pixel ground area (m²) across the country.")
cat(sprintf("0.5 ha = %.0f-%.0f pixels   |   1 ha = %.0f-%.0f pixels\n",
            0.5 / max(sizes$px_ha), 0.5 / min(sizes$px_ha),
            1 / max(sizes$px_ha), 1 / min(sizes$px_ha)))
```

What *does* bound the sub-hectare decision is that these objects have almost no measurable
geometry: convexity, circularity and `shape_index` on a 4-pixel blob are close to meaningless, so
the model is working almost entirely from spectral/context predictors there.

```{r quantum}
q <- sizes[n_pixels <= 24, .(objects = .N, area_ha_range = sprintf("%.2f-%.2f", min(area_ha),
                                                                   max(area_ha))),
           by = n_pixels][order(n_pixels)]
kable(q, format.args = list(big.mark = ","),
      col.names = c("pixels", "objects", "area range (ha)"),
      caption = "The smallest objects by pixel count, with the latitude-driven area spread.")
```

Note the shape of that count column: it **dips** from 3796 one-pixel objects to 778 at five
pixels, then climbs monotonically (6 px → 3081, 20 px → 12474). A segmentation floor would
produce a hard cut, not a dip-then-rise, so something is producing isolated 1-2 pixel objects that
the 3-5 pixel range does not get. The step-05 1-pixel dilation connectivity hack (docs/05) is the
obvious suspect. Worth a look in QGIS — it is a small share of area either way.

```{r quantum-note}
#| results: asis
sub <- sizes[area_ha < 1]
cat(sprintf(
  "\nSub-hectare objects: **%s** (%.2f %% of objects) holding **%.1f kha** (%.3f %% of area). ",
  format(nrow(sub), big.mark = ","), 100 * nrow(sub) / nrow(sizes),
  sum(sub$area_ha) / 1e3, 100 * sum(sub$area_ha) / sum(sizes$area_ha)))
cat(sprintf("Of these, %s are a **single pixel** (%.1f %% of the sub-hectare set).\n",
            format(sum(sub$n_pixels == 1L), big.mark = ","),
            100 * mean(sub$n_pixels == 1L)))
```

## 3 · The expanded distribution

```{r expanded}
#| fig-height: 4.2
brk <- SIZE_BREAKS_VIEW[is.finite(SIZE_BREAKS_VIEW) & SIZE_BREAKS_VIEW > 0]
ggplot(sizes, aes(area_ha)) +
  geom_histogram(bins = 90, fill = "grey40") +
  geom_vline(xintercept = brk, colour = "firebrick", linetype = 2) +
  scale_x_log10(breaks = c(.09, .5, 1, 10, 50, 300, 1e3, 1e4, 1e5),
                labels = c("0.09", "0.5", "1", "10", "50", "300", "1k", "10k", "100k")) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "object area (ha, log scale)", y = "objects",
       title = "Size distribution of every step-05 object",
       subtitle = "dashed = display class breaks; 0.09 ha is the one-pixel floor")
```

The distribution is the classic fire-size power law with a hard left wall at the pixel floor.
Read the left wall carefully: a monotone rise into the smallest bin is what a **detection-limited**
distribution looks like — the smallest classes are dominated by whatever the segmentation
produces at its resolution limit, real fire or not, which is exactly why the model's confidence
there (§6) is the thing to look at.

## 4 · Is the small-object share stable across years?

If the sub-hectare fraction jumped around by year, a single global minimum size would be the
wrong instrument. If it is stable, one cut is defensible.

```{r by-year}
#| fig-height: 4.2
yr <- sizes[, .(n = .N, area_kha = sum(area_ha) / 1e3,
                pct_n_lt1    = 100 * mean(area_ha < 1),
                pct_area_lt1 = 100 * sum(area_ha[area_ha < 1]) / sum(area_ha),
                pct_n_lt05   = 100 * mean(area_ha < 0.5)), by = fire_year][order(fire_year)]
py1 <- ggplot(yr, aes(fire_year)) +
  geom_col(aes(y = n), fill = "grey40") + scale_y_continuous(labels = scales::comma) +
  labs(x = NULL, y = "objects", title = "Objects per fire-year")
py2 <- ggplot(yr, aes(fire_year)) +
  geom_line(aes(y = pct_n_lt1, colour = "objects < 1 ha")) +
  geom_line(aes(y = pct_n_lt05, colour = "objects < 0.5 ha")) +
  geom_line(aes(y = pct_area_lt1, colour = "area < 1 ha")) +
  scale_colour_manual(NULL, values = c("objects < 1 ha" = "grey20",
                                       "objects < 0.5 ha" = "steelblue",
                                       "area < 1 ha" = "firebrick")) +
  labs(x = "fire-year", y = "% of the year", title = "Sub-hectare share by year") +
  theme(legend.position = "bottom")
py1 / py2
```

```{r by-year-table}
kable(yr, digits = c(0, 0, 1, 2, 3, 2), format.args = list(big.mark = ","),
      col.names = c("fire-year", "objects", "area (kha)", "% obj <1 ha", "% area <1 ha",
                    "% obj <0.5 ha"),
      caption = "Per-year totals and sub-hectare shares.")
```

## 5 · The labelled data by size and class

Where the labels are is where the model has evidence. This is the constraint on any
size-specific claim, including the thresholds.

```{r labels}
tag <- clean_tagged(verbose = FALSE)
set(tag, j = "size_class", value = size_class(tag$area_ha))
lt <- dcast(tag[, .N, by = .(size_class, class)], size_class ~ class,
            value.var = "N", fill = 0L, drop = FALSE)
setnames(lt, c("0", "1"), c("non-fire", "fire"), skip_absent = TRUE)
lt[, total := `non-fire` + fire][, prevalence := round(fire / total, 3)]
kable(lt, format.args = list(big.mark = ","),
      caption = "Labelled objects by size class and class (the fitting set, 5255 objects).")
```

```{r labels-plot}
#| fig-height: 4
ggplot(tag, aes(area_ha, fill = factor(class, 0:1, c("non-fire", "fire")))) +
  geom_histogram(bins = 60, position = "identity", alpha = .55) +
  geom_vline(xintercept = brk, colour = "grey30", linetype = 2) +
  scale_x_log10(breaks = c(.09, .5, 1, 10, 50, 300, 1e3, 1e4, 1e5),
                labels = c("0.09", "0.5", "1", "10", "50", "300", "1k", "10k", "100k")) +
  scale_fill_manual(NULL, values = c("non-fire" = "steelblue", "fire" = "firebrick")) +
  labs(x = "object area (ha, log scale)", y = "labelled objects",
       title = "Size distribution of the labelled set, by class") +
  theme(legend.position = "bottom")
```

```{r labels-vs-all}
cmp <- merge(sizes[, .(all_pct = 100 * .N / nrow(sizes)), by = size_class],
             tag[, .(lab_pct = 100 * .N / nrow(tag)), by = size_class],
             by = "size_class", all.x = TRUE)[order(size_class)]
cmp[is.na(lab_pct), lab_pct := 0]
kable(cmp, digits = 2, col.names = c("size class", "% of all objects", "% of labels"),
      caption = "Labels are NOT a size-representative sample of the objects.")
```

## 6 · Model probability and uncertainty

```{r load-preds}
pf <- list.files(PRED_DIR, pattern = sprintf("^objects_[0-9]{4}_pred_%s\\.csv$", VARIANT),
                 full.names = TRUE)
pred_years <- sort(as.integer(sub("^objects_([0-9]{4}).*$", "\\1", basename(pf))))
HAS_PREDS  <- length(pred_years) > 0L
COMPLETE   <- setequal(pred_years, years)
```

```{r preds-status}
#| results: asis
if (!HAS_PREDS) {
  cat("::: {.callout-warning}\n## No predictions yet\n",
      "Run `Rscript collection-01/workflow/06-object_model.R predict all` (~37 min, tmux),",
      "then re-render. Sections 1-5 above are complete and do not depend on it.\n:::\n")
} else if (!COMPLETE) {
  cat(sprintf("::: {.callout-warning}\n## Partial predictions\n%d of %d fire-years scored (%s).%s\n:::\n",
      length(pred_years), length(years), paste(range(pred_years), collapse = "-"),
      " Everything below covers only those years."))
} else {
  cat(sprintf("All %d fire-years scored.\n", length(pred_years)))
}
```

```{r prob-join}
#| eval: !expr HAS_PREDS
pr <- rbindlist(lapply(pf, fread))
d  <- merge(sizes, pr[, .(oid, p_mean, p_sd, p_q05, p_q95, p_width)], by = "oid")
d  <- cbind(d, apply_thresholds(d$area_ha, d$p_mean))
# "undecided": the posterior interval straddles the cut that applies to this object, i.e. the
# model cannot place it on either side at the 5-95 % level. This is the uncertainty measure that
# speaks to a minimum-size decision — unlike p_width, it is relative to the decision being made.
d[, undecided := p_q05 < p_thresh & p_q95 > p_thresh]
cat(sprintf("%s scored objects joined (%s unscored: NA predictors)\n",
            format(sum(!is.na(d$p_mean)), big.mark = ","),
            format(sum(is.na(d$p_mean)), big.mark = ",")))
```

### 6.1 By size class

```{r prob-table}
#| eval: !expr HAS_PREDS
ps <- d[!is.na(p_mean), .(
  n            = .N,
  area_kha     = sum(area_ha) / 1e3,
  p_thresh     = first(p_thresh),
  med_p        = round(median(p_mean), 3),
  mean_width   = round(mean(p_width), 3),
  med_width    = round(median(p_width), 3),
  pct_fire     = round(100 * mean(fire == 1L), 1),
  pct_area_fire= round(100 * sum(area_ha[fire == 1L]) / sum(area_ha), 1),
  pct_undecided= round(100 * mean(undecided), 1)), by = size_class][order(size_class)]
kable(ps, format.args = list(big.mark = ","),
      col.names = c("size class", "objects", "area (kha)", "cut", "median p", "mean width",
                    "median width", "% called fire", "% area called fire", "% undecided"),
      caption = "Model output by size class. 'cut' is the deployed threshold for that class; '% undecided' = the p_q05-p_q95 interval straddles it.")
```

```{r prob-dists}
#| eval: !expr HAS_PREDS
#| fig-height: 6
dd <- d[!is.na(p_mean)]
g1 <- ggplot(dd, aes(p_mean)) +
  geom_histogram(bins = 60, fill = "grey40") +
  geom_vline(aes(xintercept = p_thresh), colour = "firebrick", linetype = 2) +
  facet_wrap(~size_class, scales = "free_y", nrow = 2) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "p_mean", y = "objects", title = "Posterior mean probability of fire",
       subtitle = "dashed = the deployed cut for that class")
g2 <- ggplot(dd, aes(p_width)) +
  geom_histogram(bins = 60, fill = "steelblue4") +
  facet_wrap(~size_class, scales = "free_y", nrow = 2) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "p_width (q95 - q05)", y = "objects", title = "Posterior width — how unsure the model is")
g1 / g2
```

### 6.2 Global distributions

```{r prob-global}
#| eval: !expr HAS_PREDS
#| fig-height: 3.6
gg1 <- ggplot(dd, aes(p_mean)) + geom_histogram(bins = 80, fill = "grey40") +
  scale_y_continuous(labels = scales::comma) + labs(x = "p_mean", y = "objects", title = "p_mean, all objects")
gg2 <- ggplot(dd, aes(p_width)) + geom_histogram(bins = 80, fill = "steelblue4") +
  scale_y_continuous(labels = scales::comma) + labs(x = "p_width", y = NULL, title = "p_width, all objects")
gg1 + gg2
```

```{r prob-vs-size}
#| eval: !expr HAS_PREDS
#| fig-height: 4.2
ggplot(dd[sample(.N, min(.N, 300000L))], aes(area_ha, p_width)) +
  geom_bin2d(bins = 80) +
  scale_x_log10(breaks = c(.09, .5, 1, 10, 50, 300, 1e3, 1e4, 1e5),
                labels = c("0.09", "0.5", "1", "10", "50", "300", "1k", "10k", "100k")) +
  scale_fill_viridis_c(trans = "log10", labels = scales::comma) +
  labs(x = "area (ha, log)", y = "p_width", title = "Uncertainty vs object size",
       subtitle = "300 k-object subsample")
```

### 6.3 The minimum-size decision

```{r min-size}
#| eval: !expr HAS_PREDS
cuts <- c(0, 0.09, 0.18, 0.5, 1, 2, 5, 10)
ms <- rbindlist(lapply(cuts, function(k) {
  keep <- dd[area_ha >= k]
  drop <- dd[area_ha < k]
  data.table(min_size_ha = k,
             objects_kept   = nrow(keep),
             pct_obj_dropped  = round(100 * nrow(drop) / nrow(dd), 2),
             pct_area_dropped = round(100 * sum(drop$area_ha) / sum(dd$area_ha), 3),
             pct_fire_area_dropped = round(
               100 * sum(drop$area_ha[drop$fire == 1L]) / sum(dd$area_ha[dd$fire == 1L]), 3),
             mean_width_dropped = round(mean(drop$p_width), 3),
             pct_undecided_dropped = round(100 * mean(drop$undecided), 1),
             mean_width_kept = round(mean(keep$p_width), 3),
             pct_undecided_kept = round(100 * mean(keep$undecided), 1))
}))
kable(ms, format.args = list(big.mark = ","),
      caption = "What each candidate minimum size discards, and how uncertain the discarded vs retained objects are.")
```

Read this table as a trade: a cut is *cheap* when `pct_area_dropped` is small, and *justified*
when `mean_width_dropped` / `pct_undecided_dropped` are clearly worse than the retained set —
i.e. we are discarding precisely the objects the model cannot classify.

## 7 · Notes

- Everything here is the **grouped** (22-predictor) model, the deployed one.
- `% called fire` figures inherit the threshold-level caveat from docs/06: the labels are not a
  random sample of objects, so the *level* of the small-object cuts is uncalibrated and these
  percentages are an upper bound on plausible fire rates, not an estimate of them.
- Objects with NA predictors (all-dieback, docs/05 §3) carry NA probabilities and are excluded
  from §6 counts but present in §1-4.
