---
title: "Objects: size, model probability, and the classification cuts"
subtitle: "Step 06 · all fire-years · collection-01"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: true
    fig-width: 9
    fig-height: 5
---

```{r setup}
#| message: false
#| warning: false
knitr::opts_chunk$set(message = FALSE, warning = FALSE, dev = "png", dpi = 120)
# Every step-06 path is written from the repo ROOT. knitr restores the working directory after
# each chunk, so a setwd() in a later chunk would silently apply to that chunk only (and an
# unresolvable POLY_DIR yields an EMPTY table, not an error) — root.dir is the only thing that
# holds for the whole render.
knitr::opts_knit$set(root.dir = normalizePath(if (dir.exists("collection-01")) "." else "../.."))
library(data.table)
library(ggplot2)
library(patchwork)
library(knitr)
theme_set(theme_bw(base_size = 11) +
          theme(panel.grid.minor = element_blank(),
                strip.background = element_rect(fill = "grey92", colour = NA)))
```

```{r paths}
source("collection-01/scripts/objects_data_functions.R")
PRED_DIR <- OBJ_PRED_DIR
ANA_DIR  <- OBJ_ANA_DIR
```

## Why this notebook

Four things that have to be read together, because each one decides how to read the others:

1. **The collection's minimum mapped fire size.** Every burned-area product declares one. Ours
   is not yet chosen, and the honest basis for choosing it is not a convention borrowed from
   another product — it is *where our own model stops being able to tell fire from non-fire*.
2. **Where the mapping effort and the error live.** 61 % of objects sit in one band, and the
   thresholds differ by size, so any statement about accuracy is meaningless without
   knowing how objects and area distribute across sizes.
3. **The classification cuts themselves** (§7) — the per-size-band Youden thresholds that turn
   `p_mean` into the `fire` call, with the sensitivity/specificity they buy and how well the
   labels actually identify them.
4. **The per-year sanity check** (§8). This is the diagnostic that caught the `fire_year` /
   `year_calendar` label-prevalence leak (docs/06 §4), and it
   is the first thing to look at after any refit.

Sections 1–4 need only the step-05 metrics. Sections 5–8 need a `predict all` run; §7 also needs
`scripts/objects_threshold.R`.

```{r load-sizes}
years <- object_years()
if (!length(years)) stop("no step-05 raster_metrics CSVs under ", POLY_DIR,
                         " (wd = ", getwd(), ") — nothing to summarise")
sizes <- rbindlist(lapply(years, function(fy) {
  d <- fread(file.path(POLY_DIR, sprintf("objects_%d_raster_metrics.csv", fy)),
             select = c("oid", "n_pixels", "area_ha"))
  d[, fire_year := fy][]
}))
# set(), not `:=`: the column and the helper share a name, and inside data.table's j-scope the
# argument promise fails to resolve under knitr's eval frame
set(sizes, j = "size_class", value = size_class(sizes$area_ha))
cat(sprintf("%s objects, fire-years %d-%d, %.2f Mha total\n",
            format(nrow(sizes), big.mark = ","), min(years), max(years),
            sum(sizes$area_ha) / 1e6))
```

## 1 · How the objects and the area split by size

`size_class` uses the **display** breaks (0.5, 1, 50, 300, 1000 ha) — six classes. These are not
the four threshold bands: `<0.5` / `0.5-1` exist only because the minimum-size decision lives
inside them, and `300-1000` / `>=1000` because we want to see whether the very large objects
behave apart (the threshold work found they do not, statistically).

```{r size-table}
tab <- sizes[, .(n = .N, area_kha = sum(area_ha) / 1e3), by = size_class][order(size_class)]
tab[, `:=`(pct_n = 100 * n / sum(n), pct_area = 100 * area_kha / sum(area_kha))]
tab[, `:=`(cum_pct_n = cumsum(pct_n), cum_pct_area = cumsum(pct_area))]
kable(tab, digits = c(0, 0, 1, 2, 3, 2, 3), format.args = list(big.mark = ","),
      col.names = c("size class", "objects", "area (kha)", "% objects", "% area",
                    "cum % objects", "cum % area"),
      caption = "Objects and area by display size class, all fire-years pooled.")
```

The asymmetry between the two "%" columns is the whole story of a minimum-size cut: it buys a
large reduction in object count for a small loss of area. The `cum %` columns read directly as
"if I cut here, I discard this many objects and this much area".

```{r size-plot}
#| fig-height: 4
p1 <- ggplot(tab, aes(size_class, n)) +
  geom_col(fill = "grey35") + scale_y_log10(labels = scales::comma) +
  labs(x = NULL, y = "objects (log)", title = "Object count") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
p2 <- ggplot(tab, aes(size_class, pct_area)) +
  geom_col(fill = "firebrick") +
  labs(x = NULL, y = "% of total area", title = "Share of burned area") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
p1 + p2
```

## 2 · The pixel scale — what "<1 ha" actually means

Areas are **not** on a 0.09 ha grid. The objects carry lat/lon pixel coordinates — ~30 m *at the
equator* — and area is measured on the ellipsoid, so one pixel is `900 * cos(lat)` m²: 831 m² in
Formosa, 517 m² in southern Santa Cruz. Harmless, but it means a sub-hectare class is a pixel-count
**range**, not a fixed count, and the same 15-pixel object is `>=1 ha` in the north and `<1 ha` in
Patagonia.

```{r px-scale}
sizes[, px_ha := area_ha / n_pixels]
pxq <- sizes[, .(min = min(px_ha), q05 = quantile(px_ha, .05), median = median(px_ha),
                 q95 = quantile(px_ha, .95), max = max(px_ha))]
kable(round(pxq * 1e4, 1), col.names = c("min", "q05", "median", "q95", "max"),
      caption = "Per-pixel ground area (m²) across the country.")
cat(sprintf("0.5 ha = %.0f-%.0f pixels   |   1 ha = %.0f-%.0f pixels\n",
            0.5 / max(sizes$px_ha), 0.5 / min(sizes$px_ha),
            1 / max(sizes$px_ha), 1 / min(sizes$px_ha)))
```

What *does* bound the sub-hectare decision is that these objects have almost no measurable
geometry: convexity, circularity and `shape_index` on a 4-pixel blob are close to meaningless, so
the model is working almost entirely from spectral/context predictors there.

```{r quantum}
q <- sizes[n_pixels <= 24, .(objects = .N, area_ha_range = sprintf("%.2f-%.2f", min(area_ha),
                                                                   max(area_ha))),
           by = n_pixels][order(n_pixels)]
kable(q, format.args = list(big.mark = ","),
      col.names = c("pixels", "objects", "area range (ha)"),
      caption = "The smallest objects by pixel count, with the latitude-driven area spread.")
```

Note the shape of that count column: it **dips** from 3796 one-pixel objects to 778 at five
pixels, then climbs monotonically (6 px → 3081, 20 px → 12474). A segmentation floor would
produce a hard cut, not a dip-then-rise, so something is producing isolated 1-2 pixel objects that
the 3-5 pixel range does not get. The step-05 1-pixel dilation connectivity hack (docs/05) is the
obvious suspect. Worth a look in QGIS — it is a small share of area either way.

```{r quantum-note}
#| results: asis
sub <- sizes[area_ha < 1]
cat(sprintf(
  "\nSub-hectare objects: **%s** (%.2f %% of objects) holding **%.1f kha** (%.3f %% of area). ",
  format(nrow(sub), big.mark = ","), 100 * nrow(sub) / nrow(sizes),
  sum(sub$area_ha) / 1e3, 100 * sum(sub$area_ha) / sum(sizes$area_ha)))
cat(sprintf("Of these, %s are a **single pixel** (%.1f %% of the sub-hectare set).\n",
            format(sum(sub$n_pixels == 1L), big.mark = ","),
            100 * mean(sub$n_pixels == 1L)))
```

## 3 · The expanded distribution

```{r expanded}
#| fig-height: 4.2
brk <- SIZE_BREAKS_VIEW[is.finite(SIZE_BREAKS_VIEW) & SIZE_BREAKS_VIEW > 0]
ggplot(sizes, aes(area_ha)) +
  geom_histogram(bins = 90, fill = "grey40") +
  geom_vline(xintercept = brk, colour = "firebrick", linetype = 2) +
  scale_x_log10(breaks = c(.09, .5, 1, 10, 50, 300, 1e3, 1e4, 1e5),
                labels = c("0.09", "0.5", "1", "10", "50", "300", "1k", "10k", "100k")) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "object area (ha, log scale)", y = "objects",
       title = "Size distribution of every step-05 object",
       subtitle = "dashed = display class breaks; 0.09 ha is the one-pixel floor")
```

The distribution is the classic fire-size power law with a hard left wall at the pixel floor.
Read the left wall carefully: a monotone rise into the smallest bin is what a **detection-limited**
distribution looks like — the smallest classes are dominated by whatever the segmentation
produces at its resolution limit, real fire or not, which is exactly why the model's confidence
there (§6) is the thing to look at.

## 4 · Is the small-object share stable across years?

If the sub-hectare fraction jumped around by year, a single global minimum size would be the
wrong instrument. If it is stable, one cut is defensible.

```{r by-year}
#| fig-height: 4.2
yr <- sizes[, .(n = .N, area_kha = sum(area_ha) / 1e3,
                pct_n_lt1    = 100 * mean(area_ha < 1),
                pct_area_lt1 = 100 * sum(area_ha[area_ha < 1]) / sum(area_ha),
                pct_n_lt05   = 100 * mean(area_ha < 0.5)), by = fire_year][order(fire_year)]
py1 <- ggplot(yr, aes(fire_year)) +
  geom_col(aes(y = n), fill = "grey40") + scale_y_continuous(labels = scales::comma) +
  labs(x = NULL, y = "objects", title = "Objects per fire-year")
py2 <- ggplot(yr, aes(fire_year)) +
  geom_line(aes(y = pct_n_lt1, colour = "objects < 1 ha")) +
  geom_line(aes(y = pct_n_lt05, colour = "objects < 0.5 ha")) +
  geom_line(aes(y = pct_area_lt1, colour = "area < 1 ha")) +
  scale_colour_manual(NULL, values = c("objects < 1 ha" = "grey20",
                                       "objects < 0.5 ha" = "steelblue",
                                       "area < 1 ha" = "firebrick")) +
  labs(x = "fire-year", y = "% of the year", title = "Sub-hectare share by year") +
  theme(legend.position = "bottom")
py1 / py2
```

```{r by-year-table}
kable(copy(yr)[, fire_year := as.character(fire_year)],
      digits = c(0, 0, 1, 2, 3, 2), format.args = list(big.mark = ","),
      col.names = c("fire-year", "objects", "area (kha)", "% obj <1 ha", "% area <1 ha",
                    "% obj <0.5 ha"),
      caption = "Per-year totals and sub-hectare shares.")
```

## 5 · The labelled data by size and class

Where the labels are is where the model has evidence. This is the constraint on any
size-specific claim, including the thresholds.

```{r labels}
tag <- clean_tagged(verbose = FALSE)
set(tag, j = "size_class", value = size_class(tag$area_ha))
lt <- dcast(tag[, .N, by = .(size_class, class)], size_class ~ class,
            value.var = "N", fill = 0L, drop = FALSE)
setnames(lt, c("0", "1"), c("non-fire", "fire"), skip_absent = TRUE)
lt[, total := `non-fire` + fire][, prevalence := round(fire / total, 3)]
kable(lt, format.args = list(big.mark = ","),
      caption = "Labelled objects by size class and class (the fitting set, 5255 objects).")
```

```{r labels-plot}
#| fig-height: 4
ggplot(tag, aes(area_ha, fill = factor(class, 0:1, c("non-fire", "fire")))) +
  geom_histogram(bins = 60, position = "identity", alpha = .55) +
  geom_vline(xintercept = brk, colour = "grey30", linetype = 2) +
  scale_x_log10(breaks = c(.09, .5, 1, 10, 50, 300, 1e3, 1e4, 1e5),
                labels = c("0.09", "0.5", "1", "10", "50", "300", "1k", "10k", "100k")) +
  scale_fill_manual(NULL, values = c("non-fire" = "steelblue", "fire" = "firebrick")) +
  labs(x = "object area (ha, log scale)", y = "labelled objects",
       title = "Size distribution of the labelled set, by class") +
  theme(legend.position = "bottom")
```

```{r labels-vs-all}
cmp <- merge(sizes[, .(all_pct = 100 * .N / nrow(sizes)), by = size_class],
             tag[, .(lab_pct = 100 * .N / nrow(tag)), by = size_class],
             by = "size_class", all.x = TRUE)[order(size_class)]
cmp[is.na(lab_pct), lab_pct := 0]
kable(cmp, digits = 2, col.names = c("size class", "% of all objects", "% of labels"),
      caption = "Labels are NOT a size-representative sample of the objects.")
```

## 6 · Model probability and uncertainty

```{r load-preds}
pf <- list.files(PRED_DIR, pattern = "^objects_[0-9]{4}_pred\\.csv$", full.names = TRUE)
pred_years <- sort(as.integer(sub("^objects_([0-9]{4}).*$", "\\1", basename(pf))))
HAS_PREDS  <- length(pred_years) > 0L
COMPLETE   <- setequal(pred_years, years)
```

```{r preds-status}
#| results: asis
if (!HAS_PREDS) {
  cat("::: {.callout-warning}\n## No predictions yet\n",
      "Run `Rscript collection-01/workflow/06-object_model.R predict all` (~37 min, tmux),",
      "then re-render. Sections 1-5 above are complete and do not depend on it.\n:::\n")
} else if (!COMPLETE) {
  cat(sprintf("::: {.callout-warning}\n## Partial predictions\n%d of %d fire-years scored (%s).%s\n:::\n",
      length(pred_years), length(years), paste(range(pred_years), collapse = "-"),
      " Everything below covers only those years."))
} else {
  cat(sprintf("All %d fire-years scored.\n", length(pred_years)))
}
```

```{r prob-join}
#| eval: !expr HAS_PREDS
pr <- rbindlist(lapply(pf, fread))
d  <- merge(sizes, pr[, .(oid, p_mean, p_sd, p_q05, p_q95, p_width)], by = "oid")
d  <- cbind(d, apply_thresholds(d$area_ha, d$p_mean))
# "undecided": the posterior interval straddles the cut that applies to this object, i.e. the
# model cannot place it on either side at the 5-95 % level. This is the uncertainty measure that
# speaks to a minimum-size decision — unlike p_width, it is relative to the decision being made.
d[, undecided := p_q05 < p_thresh & p_q95 > p_thresh]
cat(sprintf("%s scored objects joined (%s unscored: NA predictors)\n",
            format(sum(!is.na(d$p_mean)), big.mark = ","),
            format(sum(is.na(d$p_mean)), big.mark = ",")))
```

### 6.1 By size class

```{r prob-table}
#| eval: !expr HAS_PREDS
ps <- d[!is.na(p_mean), .(
  n            = .N,
  area_kha     = sum(area_ha) / 1e3,
  p_thresh     = first(p_thresh),
  med_p        = round(median(p_mean), 3),
  mean_width   = round(mean(p_width), 3),
  med_width    = round(median(p_width), 3),
  pct_fire     = round(100 * mean(fire == 1L), 1),
  pct_area_fire= round(100 * sum(area_ha[fire == 1L]) / sum(area_ha), 1),
  pct_undecided= round(100 * mean(undecided), 1)), by = size_class][order(size_class)]
kable(ps, format.args = list(big.mark = ","),
      col.names = c("size class", "objects", "area (kha)", "cut", "median p", "mean width",
                    "median width", "% called fire", "% area called fire", "% undecided"),
      caption = "Model output by size class. 'cut' is the deployed threshold for that class; '% undecided' = the p_q05-p_q95 interval straddles it.")
```

```{r prob-dists}
#| eval: !expr HAS_PREDS
#| fig-height: 6
dd <- d[!is.na(p_mean)]
g1 <- ggplot(dd, aes(p_mean)) +
  geom_histogram(bins = 60, fill = "grey40") +
  geom_vline(aes(xintercept = p_thresh), colour = "firebrick", linetype = 2) +
  facet_wrap(~size_class, scales = "free_y", nrow = 2) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "p_mean", y = "objects", title = "Posterior mean probability of fire",
       subtitle = "dashed = the deployed cut for that class")
g2 <- ggplot(dd, aes(p_width)) +
  geom_histogram(bins = 60, fill = "steelblue4") +
  facet_wrap(~size_class, scales = "free_y", nrow = 2) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "p_width (q95 - q05)", y = "objects", title = "Posterior width — how unsure the model is")
g1 / g2
```

### 6.2 Global distributions

```{r prob-global}
#| eval: !expr HAS_PREDS
#| fig-height: 3.6
gg1 <- ggplot(dd, aes(p_mean)) + geom_histogram(bins = 80, fill = "grey40") +
  scale_y_continuous(labels = scales::comma) + labs(x = "p_mean", y = "objects", title = "p_mean, all objects")
gg2 <- ggplot(dd, aes(p_width)) + geom_histogram(bins = 80, fill = "steelblue4") +
  scale_y_continuous(labels = scales::comma) + labs(x = "p_width", y = NULL, title = "p_width, all objects")
gg1 + gg2
```

```{r prob-vs-size}
#| eval: !expr HAS_PREDS
#| fig-height: 4.2
ggplot(dd[sample(.N, min(.N, 300000L))], aes(area_ha, p_width)) +
  geom_bin2d(bins = 80) +
  scale_x_log10(breaks = c(.09, .5, 1, 10, 50, 300, 1e3, 1e4, 1e5),
                labels = c("0.09", "0.5", "1", "10", "50", "300", "1k", "10k", "100k")) +
  scale_fill_viridis_c(trans = "log10", labels = scales::comma) +
  labs(x = "area (ha, log)", y = "p_width", title = "Uncertainty vs object size",
       subtitle = "300 k-object subsample")
```

### 6.3 The minimum-size decision

```{r min-size}
#| eval: !expr HAS_PREDS
cuts <- c(0, 0.09, 0.18, 0.5, 1, 2, 5, 10)
ms <- rbindlist(lapply(cuts, function(k) {
  keep <- dd[area_ha >= k]
  drop <- dd[area_ha < k]
  data.table(min_size_ha = k,
             objects_kept   = nrow(keep),
             pct_obj_dropped  = round(100 * nrow(drop) / nrow(dd), 2),
             pct_area_dropped = round(100 * sum(drop$area_ha) / sum(dd$area_ha), 3),
             pct_fire_area_dropped = round(
               100 * sum(drop$area_ha[drop$fire == 1L]) / sum(dd$area_ha[dd$fire == 1L]), 3),
             mean_width_dropped = round(mean(drop$p_width), 3),
             pct_undecided_dropped = round(100 * mean(drop$undecided), 1),
             mean_width_kept = round(mean(keep$p_width), 3),
             pct_undecided_kept = round(100 * mean(keep$undecided), 1))
}))
kable(ms, format.args = list(big.mark = ","),
      caption = "What each candidate minimum size discards, and how uncertain the discarded vs retained objects are.")
```

Read this table as a trade: a cut is *cheap* when `pct_area_dropped` is small, and *justified*
when `mean_width_dropped` / `pct_undecided_dropped` are clearly worse than the retained set —
i.e. we are discarding precisely the objects the model cannot classify.

## 7 · The classification cuts

`p_mean` becomes a `fire` call through a **per-size-band threshold**, chosen by Youden's J on
**out-of-fold** predictions (`scripts/objects_threshold.R` on `oof_grid_5.csv`). Never in-sample: a
cut picked on probabilities the model has already seen the answers for is not a cut, it is a fit.

Why Youden and not F1 or accuracy: **J = sens + spec − 1 is the only criterion here that does not
move with prevalence**, and our labelled set is not a random sample of objects (§5, and see §8) — so
every prevalence-dependent criterion is biased by how the points were collected rather than by the
model.

```{r th-load}
th_chosen <- file.path(ANA_DIR, "threshold_chosen_grid_5.csv")
th_at05   <- file.path(ANA_DIR, "threshold_at_0.5_grid_5.csv")
th_sweep  <- file.path(ANA_DIR, "threshold_sweep_grid_5.csv")
HAS_TH    <- file.exists(th_chosen)
```

```{r th-status}
#| results: asis
if (!HAS_TH)
  cat("::: {.callout-warning}\n## No threshold run yet\n",
      "Run `Rscript collection-01/scripts/objects_threshold.R` (needs an `oof_grid_5.csv` from",
      "`06-object_model.R cv grid 5`), then re-render.\n:::\n")
```

### 7.1 What is deployed

`config/object_model_thresholds.csv` — the four bands that actually run in `predict`. Deliberately
coarser than the six display classes of §1: a band only earns its own cut if it has enough labels to
place one.

```{r th-deployed}
kable(threshold_table()[, .(stratum, n, threshold, sens, spec, J, t_boot_q05, t_boot_q95)],
      digits = c(0, 0, 4, 3, 3, 3, 3, 3), format.args = list(big.mark = ","),
      col.names = c("size band", "labels", "cut on p_mean", "sens", "spec", "Youden J",
                    "boot q05", "boot q95"),
      caption = "The deployed per-band cuts. `boot q05/q95` bootstraps the Youden pick within the band — a wide interval means this many labels do not identify the cut to three decimals.")
```

The cuts **rise with size** because the model is far more confident on large objects. Read the
bootstrap interval as the honest precision of each cut: where it is wide, the *size* cut rather than
the *probability* cut is the right instrument.

### 7.2 Every stratum and every criterion

Five disjoint size bands plus the pooled views, each at its optimum under four criteria — so the
choice of criterion is visible rather than implied.

```{r th-chosen}
#| eval: !expr HAS_TH
tc <- fread(th_chosen)
kable(tc[criterion == "J", .(stratum, n, t, sens, spec, prec, acc, J, F1, J_area,
                             t_boot_q05, t_boot_q95)],
      format.args = list(big.mark = ","),
      col.names = c("size band", "labels", "cut", "sens", "spec", "prec", "acc", "J", "F1",
                    "J (area-wt)", "boot q05", "boot q95"),
      caption = "Youden's J optimum per stratum. `>=1 ha` and `all` are pooled views of the same objects, reported for reference and excluded from the deployed pick.")
```

```{r th-criteria}
#| eval: !expr HAS_TH
kable(dcast(tc, stratum ~ criterion, value.var = "t")[
        , .(stratum, J, F1, acc, J_area)],
      digits = 4,
      col.names = c("size band", "J", "F1", "accuracy", "J (area-weighted)"),
      caption = "The cut each criterion would choose. Where these disagree strongly, the band's cut is weakly identified — F1 and accuracy chase the majority class, which is a property of the sampling, not of the model.")
```

### 7.3 What the 0.5 default would have done

```{r th-05}
#| eval: !expr HAS_TH
kable(fread(th_at05), format.args = list(big.mark = ","),
      caption = "The naive 0.5 cut, for comparison. Under grid-blocked CV the model under-detects at 0.5 — the wrong way round for a burned-area product.")
```

### 7.4 The curves

```{r th-curves}
#| eval: !expr HAS_TH
#| fig-height: 7
sw <- fread(th_sweep)
DEPLOY <- threshold_table()$stratum
c1 <- ggplot(melt(sw[, .(stratum, t, J, F1, acc, J_area)], id.vars = c("stratum", "t")),
             aes(t, value, colour = variable)) +
  geom_line(linewidth = .4) + facet_wrap(~stratum) +
  geom_vline(xintercept = 0.5, linetype = 3) +
  scale_colour_brewer(palette = "Dark2") +
  labs(title = "Criterion vs threshold", subtitle = "dotted = the 0.5 default",
       x = "cut on p_mean", y = NULL, colour = NULL)
c2 <- ggplot(sw, aes(1 - spec, sens, colour = stratum)) +
  geom_line(linewidth = .5) + geom_abline(linetype = 3) +
  geom_point(data = merge(fread(th_chosen)[criterion == "J"], data.table(stratum = unique(sw$stratum)),
                          by = "stratum"), size = 2) +
  labs(title = "ROC by size band", subtitle = "dot = the Youden pick",
       x = "1 - specificity", y = "sensitivity", colour = NULL)
c1 / c2
```

## 8 · Per-year sanity check — the leak diagnostic

**This is the section to read first after any refit.** The labelled set's per-year prevalence is an
artifact of where collaborators drew, ranging 0.00 to 1.00 across years and absent entirely in seven
fire-years. If any predictor lets the model identify the year, it learns that sampling pattern and
reports it back as a fire rate. That is exactly what happened when `fire_year` and `year_calendar`
were predictors (docs/06 §4): Spearman(label prevalence,
predicted fire %) reached **0.83**, fire-year 1998 came out **100.0 % fire**, and grid-blocked CV
could not see any of it because every fold contains 17–20 of the 21 labelled years.

The current predictor set contains no year identifier, so **this correlation should now be weak**.
It will not be zero — years genuinely differ, and collaborators drew more in big fire years — but a
strong monotone relationship is the alarm.

```{r year-qc}
#| eval: !expr HAS_PREDS
lab_y <- tag[, .(labels = .N, lab_prev = round(mean(class == 1L), 3)), by = fire_year]
yq <- merge(
  d[!is.na(p_mean), .(objects = .N,
                      pred_fire_pct = round(100 * mean(fire == 1L), 1),
                      med_p = round(median(p_mean), 3),
                      area_kha = round(sum(area_ha) / 1e3, 1)), by = fire_year],
  lab_y, by = "fire_year", all.x = TRUE)[order(fire_year)]
kable(copy(yq)[, fire_year := as.character(fire_year)], format.args = list(big.mark = ","),
      col.names = c("fire-year", "objects", "% called fire", "median p", "area (kha)",
                    "labels", "label prevalence"),
      caption = "Per-year model output against per-year label prevalence. Blank label columns are the seven fire-years with no labels at all.")
```

```{r year-qc-cor}
#| eval: !expr HAS_PREDS
#| results: asis
ok <- yq[!is.na(lab_prev)]
rho <- cor(ok$lab_prev, ok$pred_fire_pct, method = "spearman")
cat(sprintf(
  "\nSpearman(label prevalence, %% called fire) over the %d labelled fire-years: **%.3f**",
  nrow(ok), rho))
cat(sprintf(" (Pearson %.3f).\n\n", cor(ok$lab_prev, ok$pred_fire_pct)))
cat(if (abs(rho) > 0.7)
      sprintf("::: {.callout-important}\n## Check this\nAt %.2f this is as strong as the leak that removed `fire_year` (0.83). Either a predictor is identifying the year again, or the labelled years really are the extreme ones — check which before trusting the per-year rates.\n:::\n", rho)
    else
      sprintf("::: {.callout-note}\nAt %.2f this is far below the **0.83** the leaking 22-predictor model showed on the same statistic, and the spread of per-year rates collapsed with it (the leaking model ran 20.1-100.0 %% across years, sd 20.7). Some residual association either way is expected and benign.\n:::\n", rho))
```

### 8.1 Does the fire rate still trend with time?

A leak-free predictor set is not the same as a trend-free product. Nothing left in the model names a
year, but a predictor that merely **correlates** with the observation record would still let an
improving satellite archive masquerade as a rising fire regime — and trend analysis is the point of a
28-year collection. So this is checked directly, on the output.

Both coefficients are reported: Spearman is the honest headline (monotone, outlier-robust, assumes no
functional form), Pearson is what a linear trend test would use. A Spearman clearly above Pearson
says the relationship is rank-monotone but not a constant per-year increment.

```{r time-trend}
#| eval: !expr HAS_PREDS
tt <- d[!is.na(p_mean), .(objects = .N, fire_pct = 100 * mean(fire == 1L),
                          area_kha = sum(area_ha) / 1e3), by = fire_year][order(fire_year)]
kable(copy(tt)[, `:=`(fire_year = as.character(fire_year), fire_pct = round(fire_pct, 1),
                      area_kha = round(area_kha, 1))],
      format.args = list(big.mark = ","),
      col.names = c("fire-year", "objects", "% called fire", "area (kha)"),
      caption = "Per-year fire rate and burned area from the deployed model.")
```

```{r time-trend-cor}
#| eval: !expr HAS_PREDS
kable(data.table(
  relationship = c("fire-year vs % called fire", "fire-year vs burned area"),
  spearman = round(c(cor(tt$fire_year, tt$fire_pct, method = "spearman"),
                     cor(tt$fire_year, tt$area_kha, method = "spearman")), 3),
  pearson  = round(c(cor(tt$fire_year, tt$fire_pct),
                     cor(tt$fire_year, tt$area_kha)), 3)),
  col.names = c("relationship", "Spearman", "Pearson"),
  caption = "Residual time trend in the deployed product.")
```

```{r time-trend-verdict}
#| eval: !expr HAS_PREDS
#| results: asis
rs <- cor(tt$fire_year, tt$fire_pct, method = "spearman")
rp <- cor(tt$fire_year, tt$fire_pct)
cat(sprintf(
  "::: {.callout-note}\nResidual trend in the fire rate: **Spearman %.3f / Pearson %.3f**, over a range of %.1f-%.1f %% (sd %.1f). Some interannual structure is expected and real. What was removed to get here was a predictor that made this trend an artifact — see docs/06 \"Why no predictor may identify the year\" and the note on observation density. Treat any remaining trend as **unattributed** until it is checked against an independent record; do not publish it as a fire-regime finding on this evidence alone.\n:::\n",
  rs, rp, min(tt$fire_pct), max(tt$fire_pct), sd(tt$fire_pct)))
```

## 9 · What the model leans on — importance and ALE

Two questions share this evidence: which metrics an expert needs on the uploaded FeatureCollection to
be able to re-judge a call, and whether any predictor is doing something it should not (this is how
the `fire_year` leak was caught).

**Four measures, because none of them is trustworthy alone here.** The predictors are strongly
correlated — `area_ha` / `n_pixels` / `perimeter_m` are near-collinear, `burned_around_{1,2,3}` are
nested windows — and every importance measure mishandles correlation in its own way:

| measure | what it is | how it misleads |
|---|---|---|
| `split_share` | fraction of all forest splits taken by the column (BART's native measure, read from the saved forest) | biased toward high-cardinality continuous columns; splits credit arbitrarily between correlated columns |
| `perm_dp` | mean \|change in predicted probability\| when the column is shuffled, over a sample of **all** objects | the measure that matches the upload question — but a correlated pair can both look small |
| `perm_auc_drop` | AUC lost on the **labelled** set when the column is shuffled | in-sample (the fit saw these labels), so it is "what separates the classes it was shown", not validation |
| `ale_range` | max − min of the 1-D ALE curve, i.e. effect size in probability units | ALE not PDP **because** of the correlation: a PDP averages over combinations that do not exist (a 1-pixel object with a 10 km perimeter) and invents effects there |

```{r imp-load}
imp_f <- file.path(ANA_DIR, "importance_objects.csv")
ale_f <- file.path(ANA_DIR, "ale_curves_objects.csv")
HAS_IMP <- file.exists(imp_f) && file.exists(ale_f)
```

```{r imp-status}
#| results: asis
if (!HAS_IMP)
  cat("::: {.callout-warning}\n## Not computed yet\n",
      "Run `Rscript collection-01/scripts/objects_importance_ale.R` (in tmux), then re-render.\n:::\n")
```

```{r imp-table}
#| eval: !expr HAS_IMP
imp <- fread(imp_f)
setorder(imp, -perm_dp)
kable(imp[, .(predictor, split_share = round(split_share, 4), root_share = round(root_share, 4),
              perm_dp = round(perm_dp, 4), perm_auc_drop = round(perm_auc_drop, 4),
              ale_range = round(ale_range, 4))],
      col.names = c("predictor", "split share", "root share", "perm |dp|", "perm AUC drop",
                    "ALE range"),
      caption = "The 20 predictors, ordered by how much shuffling them moves the deployed output.")
```

```{r imp-plot}
#| eval: !expr HAS_IMP
#| fig-height: 5.5
m <- melt(imp[, .(predictor, `split share` = split_share, `perm |dp|` = perm_dp,
                  `perm AUC drop` = perm_auc_drop, `ALE range` = ale_range)],
          id.vars = "predictor")
m[, predictor := factor(predictor, levels = imp$predictor)]
ggplot(m, aes(value, predictor)) +
  geom_col(fill = "grey35") +
  facet_wrap(~variable, scales = "free_x", nrow = 1) +
  labs(x = NULL, y = NULL,
       title = "Four views of predictor importance",
       subtitle = "ordered by perm |dp|; disagreement between panels is the correlation structure, not noise")
```

```{r ale-plot}
#| eval: !expr HAS_IMP
#| fig-height: 9
ale <- fread(ale_f)
ale[, predictor := factor(predictor, levels = imp$predictor)]
ggplot(ale, aes(x, ale)) +
  geom_hline(yintercept = 0, colour = "grey70", linewidth = .3) +
  geom_step(colour = "firebrick", linewidth = .5) +
  facet_wrap(~predictor, scales = "free_x", ncol = 4) +
  labs(x = "predictor value", y = "ALE (probability units)",
       title = "1-D accumulated local effects",
       subtitle = "how P(fire) moves with each predictor, holding the rest at their observed joint distribution")
```

Reading the ALE panel: a **flat** curve means the model does not use the variable over most of its
range regardless of what the importance columns say; a **monotone** curve is a legible effect an
expert can reason with; a **jagged** one is the model fitting local structure that may not generalise.
For the upload decision, a variable earns its place by being either legible to a human adjudicating a
call, or required by the product — not by ranking high on `split_share` alone.

## 10 · Notes

- `% called fire` figures inherit the threshold-level caveat from docs/06: the labels are not a
  random sample of objects, so the *level* of the cuts is uncalibrated and these
  percentages are an upper bound on plausible fire rates, not an estimate of them. Removing the
  year leak did not change that — it removed the model *reading* the sampling, not the sampling.
- Objects with NA predictors (all-dieback, docs/05 §3) carry NA probabilities and are excluded
  from §6-9 counts but present in §1-4.
- The predictor set carries no absolute time coordinate. `doy_sin`/`doy_cos` encode season
  circularly because the fire season straddles Dec/Jan; `date_span` is a duration. See docs/06.
