---
title: "Predictor correlations — raw focal + previous-year covariates"
subtitle: "Diagnosing collinearity in the LR design (step 02). Run interactively, chunk by chunk."
author: "MapBiomas Argentina Fire"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: true
    df-print: kable
execute:
  warning: false
  message: false
---

## Why this notebook

`02-model_fitting.R`'s elastic-net search has been slow to converge. The leading
suspect is a **highly collinear design matrix**: the canonical-team design expands
21 focal features + 40 previous-year mosaic bands into 427 terms (pairwise products,
same-band and cross interactions — see `BLOCKS` in the fitting script). If the *raw*
inputs are already strongly correlated, every interaction block inherits and amplifies
that redundancy, and `glmnet`'s coordinate descent crawls.

This notebook is the first, cheapest cut: **pairwise correlations among the raw
variables**, computed *within* the two natural groups —

1. **Focal-date features** (21): the same-date Landsat reflectances + indices.
2. **Previous-year mosaic** (40): `mb_mos_<var>_<summary>`, 10 vars × 4 summaries.

The aim is to spot redundant covariates we can drop *before* the design is built, so a
later pass on the full design matrix starts from a leaner, less collinear raw set.

> **Structure.** Shared setup + correlation helpers first, then **everything for the focal
> set** (matrix → heatmap → high pairs → separability → drop decision), then **everything
> for the previous-year set**. Each group computes its own correlation matrix and analyses
> it in one place.

> **Run this interactively** (RStudio/Positron: run chunks with Ctrl/Cmd+Shift+Enter),
> not via `quarto render`. The data load is isolated in one chunk so you run it once and
> then add/iterate analyses in the same session without reloading.

> **Big data note.** ~5.7M observations across 5 regions. We read only the columns we
> need, and compute correlations with a scaled `crossprod` (BLAS-level) instead of base
> `cor(use = "pairwise.complete.obs")`, which recomputes each pair separately and is far
> slower at this scale. `SAMPLE_N` lets you estimate on a random subsample (correlations
> are stable from ~1M rows) and confirm on the full set.

## Setup

```{r setup}
suppressPackageStartupMessages({
  library(data.table)
  library(corrplot)   # clustered correlation heatmaps
})

# ── paths ────────────────────────────────────────────────────────────────────
# Resolve repo root whether the working dir is repo root or collection-01/notebooks.
find_root <- function() {
  for (p in c(".", "..", "../..", "../../..")) {
    if (file.exists(file.path(p, "collection-01", "config", "veg_fire_remap.csv")))
      return(normalizePath(p))
  }
  stop("Could not locate repo root (collection-01/config/veg_fire_remap.csv not found).")
}
ROOT     <- find_root()
VERSION  <- "1"
REGIONS  <- c("BA", "CHACO", "PAMPA", "CUYO", "PAT")
obs_csv  <- function(reg) file.path(ROOT, "collection-01", "data",
              sprintf("training_observations_%s_v%s.csv", reg, VERSION))

# ── the two raw variable groups (mirror 02-model_fitting.R) ──────────────────
FOCAL <- c("BLUE","GREEN","RED","NIR","SWIR1","SWIR2","NBR","NBR2","MIRBI","NDVI",
           "TCB","TCG","TCW","NDMI","NDSI","SAVI","NDWI","AFRI","kNDVI","EVI2","NIRv") # 21
PREV_VARS <- c("blue","green","red","nir","swir1","swir2","ndvi","ndwi","npv","ndfi") # 10
SUMM      <- c("median", "median_wet", "median_dry", "stdDev")                        # 4
PREV <- as.vector(t(outer(PREV_VARS, SUMM, function(v, s) sprintf("mb_mos_%s_%s", v, s)))) # 40
stopifnot(length(FOCAL) == 21, length(PREV) == 40)

# ── knobs ────────────────────────────────────────────────────────────────────
SAMPLE_N  <- NULL        # NULL = all rows; or e.g. 1e6 for a fast interactive estimate
COR_HI    <- 0.90        # |r| threshold for "redundant" pair reporting / dropping
SEED      <- 1
```

## Load data (run once)

Reads only the 61 raw columns (+ `region`) from all five region tables, then row-binds.
At full size this is the only slow step; keep the result (`dt`) in your session.

```{r load-data}
keep_cols <- c("region", FOCAL, PREV)
dt <- rbindlist(lapply(REGIONS, function(reg) {
  message("  reading ", basename(obs_csv(reg)), " ...")
  fread(obs_csv(reg), select = keep_cols)
}), use.names = TRUE)

if (!is.null(SAMPLE_N) && SAMPLE_N < nrow(dt)) {
  set.seed(SEED)
  dt <- dt[sample(.N, SAMPLE_N)]
  message(sprintf("subsampled to %s rows", format(SAMPLE_N, big.mark = ",")))
}

cat(sprintf("Loaded %s obs x %d raw vars (%d focal + %d prev), regions: %s\n",
            format(nrow(dt), big.mark = ","), length(FOCAL) + length(PREV),
            length(FOCAL), length(PREV), paste(REGIONS, collapse = ", ")))
```

```{r na-report}
# NA / non-finite audit per group — high-NA columns distort any correlation.
na_summary <- function(cols) {
  s <- dt[, lapply(.SD, function(x) sum(!is.finite(x))), .SDcols = cols]
  data.table(var = cols, n_nonfinite = as.integer(unlist(s)),
             pct = round(100 * as.integer(unlist(s)) / nrow(dt), 3))[order(-n_nonfinite)]
}
cat("Focal group — non-finite counts (top):\n");      print(head(na_summary(FOCAL), 10))
cat("\nPrevious-year group — non-finite counts (top):\n"); print(head(na_summary(PREV), 10))
```

## Correlation helpers (big-data friendly)

`cor_fast` standardizes columns once and forms `crossprod / (n-1)` — a single BLAS matmul,
much faster than per-pair `cor()` at millions of rows. Rows with any non-finite value in
the group are dropped (complete-case *within the group*, reported below). `cor_spearman`
ranks each column with `data.table::frank` (fast, C-level) then reuses `cor_fast`.

```{r cor-helpers}
# Pearson via scaled crossprod. M: numeric matrix. Returns p x p correlation matrix.
cor_fast <- function(M) {
  ok <- is.finite(rowSums(M))            # complete cases within this group
  M  <- M[ok, , drop = FALSE]
  attr_n <- nrow(M)
  M  <- scale(M)                         # center + unit sd per column (BLAS-light)
  C  <- crossprod(M) / (attr_n - 1)
  C[!is.finite(C)] <- NA                 # guard zero-variance columns
  attr(C, "n_used") <- attr_n
  C
}

# Spearman = Pearson on ranks. frank handles ties (average) and is fast on big columns.
cor_spearman <- function(M) {
  R <- apply(M, 2, function(x) frank(x, ties.method = "average", na.last = "keep"))
  cor_fast(R)
}

# Long table of pairs sorted by |r|, optionally only those above a threshold.
high_pairs <- function(C, thresh = 0) {
  ut <- which(upper.tri(C), arr.ind = TRUE)
  d  <- data.table(var1 = rownames(C)[ut[, 1]], var2 = colnames(C)[ut[, 2]],
                   r = C[ut])
  d  <- d[is.finite(r)][abs(r) >= thresh][order(-abs(r))]
  d[, r := round(r, 3)]
  d[]
}

# Long table of ALL variable pairs, for interactive View()/sorting.
# both_dirs = TRUE  -> each pair listed twice (i->j and j->i):
# filtering/sorting by var1 shows every partner of a variable together (best for View()).
# both_dirs = FALSE -> upper triangle only (each unordered pair once), like high_pairs.
# Adds abs_r so you can sort by strength regardless of sign.
cor_pairs_long <- function(C, both_dirs = TRUE, digits = 3) {
    idx <- if (both_dirs) which(row(C) != col(C), arr.ind = TRUE)
           else           which(upper.tri(C),     arr.ind = TRUE)
    d <- data.table(var1  = rownames(C)[idx[, 1]],
                    var2  = colnames(C)[idx[, 2]],
                    r     = C[idx],
                    abs_r = abs(C[idx]))
    d <- d[is.finite(r)]
    d[, `:=`(r = round(r, digits), abs_r = round(abs_r, digits))]
    setorder(d, -abs_r)
    d[]
}

# Greedy redundancy pruning (caret::findCorrelation logic, reimplemented, no dependency).
#   protect : variables that are never dropped (must-keep).
#   score   : optional named numeric, higher = keep; if given, the lower-scoring member of
#             an offending pair is dropped. If NULL, fall back to dropping the variable with
#             the higher mean |r| to all others (the more globally-redundant one).
# Returns the variables flagged for removal, in drop order. (See the algorithm note below.)
find_redundant <- function(C, thresh, protect = character(0), score = NULL) {
  A <- abs(C); diag(A) <- NA
  drop <- character(0)
  repeat {
    keep <- setdiff(rownames(A), drop)
    sub  <- A[keep, keep, drop = FALSE]
    if (!any(sub > thresh, na.rm = TRUE)) break
    mx   <- which(sub == max(sub, na.rm = TRUE), arr.ind = TRUE)[1, ]
    cand <- keep[c(mx[1], mx[2])]
    free <- setdiff(cand, protect)
    if (length(free) == 0L) {                       # both protected: retire this pair
      A[cand[1], cand[2]] <- A[cand[2], cand[1]] <- NA; next
    }
    if (length(free) == 1L) { drop <- c(drop, free); next }   # one protected: drop the other
    loser <- if (!is.null(score)) free[which.min(score[free])]      # drop less-separable
             else free[which.max(c(mean(sub[free[1], ], na.rm = TRUE),
                                   mean(sub[free[2], ], na.rm = TRUE)))]  # drop more-redundant
    drop <- c(drop, loser)
  }
  drop
}
```

### How `find_redundant` chooses drop candidates

The pruner is the iterative, pairwise rule from `caret::findCorrelation`. Each step:

1. Take the absolute correlation matrix `|R|` — sign is irrelevant for redundancy.
2. Find the single pair `(i, j)` with the **largest** `|r|` still above the threshold.
3. Decide which of the two to remove:
   - if one is **protected** (a must-keep variable), drop the other;
   - else if both are protected, leave the pair (retire it) and move on;
   - otherwise drop the **less useful** member. With a value score (focal group → the
     univariate separability) it drops the **lower-separability** index. With no score
     (prev-year group) it falls back to dropping the variable with the **higher mean `|r|`
     to all the others** — i.e. the one that is, on average, redundant with the most other
     variables. *(This is exactly the intuition: among the highly-correlated ones, remove
     first the variable that correlates highly with many others.)*
4. Remove that one variable, then repeat on what's left until no pair exceeds the threshold.

It is iterative and **pairwise**, not "rank every variable by how many others it
correlates with and cut the top N". The one-shot version can remove two members of the
same cluster when removing one would already break all the offending pairs; going
pair-by-pair removes the *minimum* set. The mean-`|r|` comparison is only the **tiebreak
within the current worst pair**, not a global ranking.

# Focal-date features (21)

```{r focal-matrix}
Mf <- as.matrix(dt[, ..FOCAL])
Cf_p <- cor_fast(Mf)          # Pearson
Cf_s <- cor_spearman(Mf)      # Spearman
cat(sprintf("Focal Pearson computed on %s complete-case rows.\n",
            format(attr(Cf_p, "n_used"), big.mark = ",")))
```

## Focal — correlation heatmap & high pairs

```{r focal-heatmap, fig.width=9, fig.height=9}
corrplot(Cf_p, method = "color", type = "upper", order = "hclust",
         addCoef.col = "grey25", number.cex = 0.55, tl.cex = 0.8, tl.col = "black",
         title = "Focal features — Pearson (hclust order)", mar = c(0, 0, 1.5, 0))
```

```{r focal-high-pairs}
cat("Focal — Pearson |r| >= ", COR_HI, ":\n", sep = "")
print(high_pairs(Cf_p, COR_HI))
cat("\nFocal — Spearman |r| >= ", COR_HI, " (rank-based, robust to non-linearity):\n", sep = "")
print(high_pairs(Cf_s, COR_HI))
```

## Focal — univariate separability (burned vs unburned, whole dataset)

How well does each index *alone* separate burned from unburned observations? The headline
is the **AUC** (Mann–Whitney / threshold-free): 0.5 = no separation, 1 = perfect; `sep` is
the rescaled `|AUC − 0.5|·2` (0–1). Also reported: Cohen's *d* (standardized mean
difference) and the burned/unburned means. This needs the `burned` label (absent from
`dt`), so it does a **fresh lightweight read** (only `burned` + the 21 focal cols) on the
**full** data — independent of `SAMPLE_N`, and it does not touch `dt`.

```{r separability}
sep_cols <- c("burned", FOCAL)
sepdt <- rbindlist(lapply(REGIONS, function(reg) {
  message("  reading burned+focal from ", basename(obs_csv(reg)), " ...")
  fread(obs_csv(reg), select = sep_cols)
}), use.names = TRUE)

auc1 <- function(x, y) {                       # Mann-Whitney AUC, NA-safe, big-data ok
  ok <- is.finite(x); x <- x[ok]; y <- y[ok]
  n1 <- as.numeric(sum(y == 1L)); n0 <- as.numeric(sum(y == 0L))  # numeric: n1*n0 overflows int
  if (n1 == 0 || n0 == 0) return(NA_real_)
  r <- frank(x, ties.method = "average")
  (sum(r[y == 1L]) - n1 * (n1 + 1) / 2) / (n1 * n0)
}
yb <- as.integer(sepdt$burned)
sep <- rbindlist(lapply(FOCAL, function(f) {
  x  <- sepdt[[f]]; ok <- is.finite(x)
  a  <- auc1(x, yb)
  mb <- mean(x[ok & yb == 1L]); mu <- mean(x[ok & yb == 0L])
  vb <- var(x[ok & yb == 1L]);  vu <- var(x[ok & yb == 0L])
  n1 <- sum(ok & yb == 1L);     n0 <- sum(ok & yb == 0L)
  sp <- sqrt(((n1 - 1) * vb + (n0 - 1) * vu) / (n1 + n0 - 2))   # pooled sd
  data.table(var = f, auc = a, sep = abs(a - 0.5) * 2,
             cohens_d = (mb - mu) / sp, mean_burned = mb, mean_unburned = mu)
}))
sep[, c("auc","sep","cohens_d","mean_burned","mean_unburned") :=
      .(round(auc,4), round(sep,4), round(cohens_d,3), round(mean_burned,4), round(mean_unburned,4))]
setorder(sep, -sep)
print(sep)
```

## Focal — drop candidates (protect NBR/NBR2 + raw optical; tiebreak by separability)

```{r focal-decision}
KEEP_FOCAL <- c("NBR", "NBR2", "BLUE", "GREEN", "RED", "NIR", "SWIR1", "SWIR2")  # always keep
sep_score  <- setNames(sep$sep, sep$var)                       # higher = more separable = keep
drop_focal <- find_redundant(Cf_p, COR_HI, protect = KEEP_FOCAL, score = sep_score)

A <- abs(Cf_p); diag(A) <- NA
focal_summary <- data.table(
  var          = FOCAL,
  separability = round(sep_score[FOCAL], 4),
  mean_abscor  = round(rowMeans(A, na.rm = TRUE), 3),
  max_abscor   = round(apply(A, 1, max, na.rm = TRUE), 3))
focal_summary[, status := fifelse(var %in% KEEP_FOCAL, "protected",
                          fifelse(var %in% drop_focal, "DROP", "keep"))]
setorder(focal_summary, status, -separability)
cat("Focal — drop candidates at |r| >", COR_HI, "(protected vars never dropped):\n")
print(focal_summary)
cat("\nDROP:", paste(drop_focal, collapse = ", "),
    "\nKEEP:", paste(setdiff(FOCAL, drop_focal), collapse = ", "), "\n")
```

## Focal — decision: reduced set (chosen 2026-06-20)

The automatic pruner above is a starting point; the **final focal set** was decided by
combining it with two manual rules — protect one greenness + one moisture index, and
prefer cheap normalized-difference indices over the (expensive) tasseled-cap trio.

**KEEP (12):** `BLUE GREEN RED NIR SWIR1 SWIR2` (raw optical) · `NBR NBR2` (burn) ·
`MIRBI` (independent burn index, max |r| ≈ 0.73) · **`NDVI`** (greenness) ·
**`NDMI`** (moisture) · `NDSI` (the one genuinely uncorrelated extra axis, max |r| ≈ 0.54).

**DROP (9):** `AFRI kNDVI SAVI EVI2 NIRv NDWI` (greenness/burn cluster, all redundant with
NDVI/NBR) **+ `TCB TCG TCW`** (tasseled-cap).

Rationale:

- **Greenness = NDVI, not kNDVI**: `r(NDVI,kNDVI)=0.986`, equal separability; NDVI is
  canonical/cheap and kNDVI's nonlinearity is recoverable from the product terms.
- **Moisture = NDMI, not NDWI/TCW**: McFeeters `NDWI` is ≈ −NDVI (`r=-0.977`), a greenness
  axis in disguise, not independent moisture. NDMI is cheaper than TCW and separates better
  (0.65 vs 0.54), and `r(NDMI,TCW)=0.86` so it covers TCW. The cost is one strongly
  collinear pair, `r(NDMI,NBR)=0.966` — accepted, because the useful moisture/ash signal is
  NDMI's component *orthogonal* to NBR (ash test: NDMI sep 0.73 vs NBR 0.45), and there is
  no cheap NIR–SWIR moisture index decorrelated from NBR (they share NIR).
- **Drop all three tasseled-cap indices**: with the 6 raw optical bands protected, `TCB`,
  `TCG`, `TCW` are *exact linear combinations* of those bands (Baig 2014 coefficients in
  `functions.py`), so a linear model — and the band×· product terms — already **spans**
  them (regress any TC index on the 6 bands → R²≈1). The pairwise pruner kept them only
  because it cannot see *multivariate* redundancy; `max_abscor` understates it.
  - **On the ash/brightness worry:** TCB (brightness) helped flag ash univariately, but
    TCB is just a weighted sum of the 6 protected bands, so the model retains full access
    to that brightness contrast — no information is lost by dropping the index itself.

This TC linear-redundancy point is exactly what the planned **design-matrix collinearity
pass** (VIF / regress-on-the-rest, not pairwise) should formalize for the full 427-term
design — see Next steps.

# Previous-year mosaic (40)

```{r prev-matrix}
Mp <- as.matrix(dt[, ..PREV])
Cp_p <- cor_fast(Mp)
Cp_s <- cor_spearman(Mp)
cat(sprintf("Prev-year Pearson computed on %s complete-case rows.\n",
            format(attr(Cp_p, "n_used"), big.mark = ",")))
```

## Prev-year — correlation heatmap & high pairs

```{r prev-heatmap, fig.width=12, fig.height=12}
corrplot(Cp_p, method = "color", type = "upper", order = "hclust",
         tl.cex = 0.6, tl.col = "black",
         title = "Previous-year mosaic — Pearson (hclust order)", mar = c(0, 0, 1.5, 0))
```

```{r prev-high-pairs}
cat("Prev-year — Pearson |r| >= ", COR_HI, ":\n", sep = "")
print(high_pairs(Cp_p, COR_HI))
cat("\nPrev-year — Spearman |r| >= ", COR_HI, ":\n", sep = "")
print(high_pairs(Cp_s, COR_HI))
```

## Prev-year — drop candidates (mean-correlation tiebreak)

No separability score is applied here (the request was about the focal indices), so the
prev-year group uses the mean-`|r|` tiebreak. Add a `protect=`/`score=` argument if you
want to force-keep specific mosaic bands too.

```{r prev-decision}
drop_prev <- find_redundant(Cp_p, COR_HI)
cat("Prev-year — drop candidates at |r| >", COR_HI, ":\n")
print(drop_prev)
cat("\nKEEP:", paste(setdiff(PREV, drop_prev), collapse = ", "), "\n")
```

## Prev-year — decision: reduced main-term set (chosen 2026-06-20)

All `>0.9` redundancy in the 40-band set is one structure: the **visible trio
(blue ≈ green ≈ red)** are near-duplicates within *every* summary (median / wet / dry /
stdDev, r ≈ 0.93–0.99). Nothing else exceeds 0.9 except `swir1_median_dry ~
swir2_median_dry` (0.917), which we **keep both** (not high enough to bother).

**Decision — keep 32 prev bands as main terms; drop 8:**

- **Keep `green` as the single visible representative** (all 4 summaries) and **drop `blue`
  and `red` entirely** (8 columns) — they're the same axis (r ≈ 0.95–0.99 with green),
  which also leaves a single optical SD in the visible (`green_stdDev`).
- Keep all 4 summaries of the other 7 base vars: `nir, swir1, swir2, ndvi, ndwi, npv, ndfi`.
- → 8 base vars × 4 summaries = **32 main terms**.

Notes:

- Prev-year is far **less collinear than focal** — only the visible trio is redundant; the
  median/wet/dry/stdDev summaries of every other var are distinct enough to keep. `stdDev`
  is its own variability axis, and for the indices `stdDev ≈ the wet−dry seasonal spread`
  (cor 0.95 for NDVI, 0.78 for NDFI) — but we keep both as cheap main effects.
- This is the **main-effect** set only. The expensive part — which bands enter the
  *interaction* blocks (B4/B5/B6, prev × focal) — is decided separately; main effects are
  cheap and only mildly collinear, so we keep them generously here.

```{r prev-decision-final}
PREV_DROP_VARS <- c("blue", "red")                       # visible duplicates of green
PREV_MAIN <- PREV[!grepl(sprintf("mb_mos_(%s)_", paste(PREV_DROP_VARS, collapse = "|")), PREV)]
stopifnot(length(PREV_MAIN) == 32)
cat("Prev main-term set:", length(PREV_MAIN), "bands (dropped blue_* and red_*):\n")
print(PREV_MAIN)
```

# Focal interaction terms (pairwise)

With the 12-feature focal main set fixed, the next term-count lever is the **pairwise
products** (the design's B3 block — the old 21-feature design had C(21,2) = 210; with 12
features it is C(12,2) = **66**). We keep an interaction only if it adds structure beyond
the mains and the other kept interactions — i.e. prune by correlation with the **12 mains
protected** (never dropped).

**Products are built from CENTERED factors**, `(X−mean)(Y−mean)`, not raw `X·Y` — the only
honest way to judge interaction redundancy, since a raw product is *mechanically* correlated
with its parents. Centering the *factors* (not re-centering the finished product) genuinely
changes the column: `(X−a)(Y−b) = XY − aY − bX + ab`.

> **Is this just reparameterization?** Only without a penalty. In an unpenalized model that
> also holds `X`, `Y` and an intercept, raw and centered products span the same space →
> identical predictions (the `−aY − bX + ab` shuffles into the intercept/mains). **Under
> glmnet's penalty it is a different model**: the penalty is computed on the coefficients,
> which differ between the two parameterizations for the same surface, so the penalized
> optimum (selection + shrinkage) differs — in the better-conditioned direction. glmnet's
> per-column standardization does *not* erase this (it scales each column but does not
> orthogonalize the product against its parents, which factor-centering partially does). For
> **deployment** the distinction vanishes: once fitted, the chosen function re-expresses
> exactly in raw products (fold the constants into the intercept/mains), so GEE keeps doing
> raw band multiplies — no means vector required.

```{r focal-interactions}
FOCAL_KEEP <- c("BLUE","GREEN","RED","NIR","SWIR1","SWIR2","NBR","NBR2","MIRBI","NDVI","NDMI","NDSI")
OPT_K <- c("BLUE","GREEN","RED","NIR","SWIR1","SWIR2")
IDX_K <- c("NBR","NBR2","MIRBI","NDVI","NDMI","NDSI")

Mk <- as.matrix(dt[, ..FOCAL_KEEP]); Mk <- Mk[is.finite(rowSums(Mk)), ]
Mc <- sweep(Mk, 2, colMeans(Mk), "-")                    # CENTER factors before multiplying

pr <- combn(FOCAL_KEEP, 2)                               # 66 distinct pairs
P  <- vapply(seq_len(ncol(pr)), function(k) Mc[, pr[1, k]] * Mc[, pr[2, k]], numeric(nrow(Mc)))
colnames(P) <- apply(pr, 2, paste, collapse = ":")

Cint     <- cor_fast(cbind(Mk[, FOCAL_KEEP], P))         # mains (raw; corr center-invariant) + centered products
drop_int <- find_redundant(Cint, COR_HI, protect = FOCAL_KEEP)   # only interactions are droppable
keep_int <- setdiff(colnames(P), drop_int)

blk <- function(t) { ab <- strsplit(t, ":")[[1]]
  if (all(ab %in% OPT_K)) "opt:opt" else if (all(ab %in% IDX_K)) "idx:idx" else "opt:idx" }
itab <- data.table(term = colnames(P), block = vapply(colnames(P), blk, ""),
                   status = fifelse(colnames(P) %in% keep_int, "keep", "DROP"))
cat(sprintf("Interactions kept: %d / 66  (|r| > %.2f, 12 mains protected, centered factors)\n",
            length(keep_int), COR_HI))
print(itab[, .(keep = sum(status == "keep"), total = .N), by = block][order(block)])
cat("\nKEEP:\n"); print(sort(keep_int))
cat("\nDROP:\n"); print(sort(drop_int))
```

## Focal interactions — decision (chosen 2026-06-20)

**Keep 34 of 66** pairwise interactions (|r| > 0.9 pruning, 12 mains protected, centered
factors) — **confirmed on the full 5.72M-obs data**. By block:

- **opt × opt: 5 / 15** — visible×visible products are as collinear as the visible bands
  themselves; only NIR/SWIR-bearing products survive (`BLUE:SWIR1, GREEN:NIR, NIR:SWIR1,
  NIR:SWIR2, SWIR1:SWIR2`).
- **idx × idx: 11 / 15** — indices carry the most complementary conditional signal.
- **opt × idx: 18 / 36**.

Every kept interaction has |r| < 0.8 to every main term. This is **34 vs the old 210**
focal pairwise terms — the single biggest term-count cut so far. (The chunk above
regenerates the kept/dropped lists from the loaded data, in sync with `COR_HI`.)

**Deployment:** fit with centered factors (better-conditioned), then fold the centering
constants into the intercept + main-effect coefficients at export → GEE deploys raw band
products, no means vector. See the note box above.

# Previous-year × focal interactions (curated blocks, median + sd)

For prev × focal we keep the **curated block structure** of the original design
(`02-model_fitting.R`, blocks B4–B6) rather than crossing everything: the blocks encode a
*relevance* prior — a band interacts with its own previous-year level (same-band), and
previous-year vegetation/fraction state modulates the fire signal — which a pure
correlation prune would discard (it removes *redundant* terms, not *useless* ones).

Two reductions vs the original: every block is restricted to the **median + sd** summaries
only (dropping `wet ≈ median` and the partly-redundant `dry`), and block membership is
aligned to the **reduced main sets** (12 focal, 32 prev).

| block | prev side (× median + sd) | × focal side | count |
|---|---|---|---|
| **B4 same-band** | green, nir, swir1, swir2, ndvi | each × its own focal band | 5 × 2 = 10 |
| **B5 idx × fire-idx** | ndvi, ndwi, npv, ndfi | × {NBR, NBR2, MIRBI, NDVI}; skip ndvi×NDVI | 4×4×2 − 2 = 30 |
| **B6 idx × fire-band** | ndvi, ndwi, npv, ndfi | × {RED, NIR, SWIR1, SWIR2} | 4×4×2 = 32 |
| **total** | | | **72** |

(was 156 with all 4 summaries.)

Aligning to the reduced sets drops three same-band pairings present in the old B4: `blue`
and `red` (removed from the prev mains — redundant with green) and `ndwi` (removed from the
focal mains). Their same-band terms would reintroduce variables we just dropped (e.g.
`blue_prev × BLUE` ≈ `green_prev × GREEN`), so they're excluded. The literal old membership
would be 78 (add those 3 pairs × 2 summaries).

The blocks are **not** correlation-pruned among themselves — they're already a curated,
sparse set. The multivariate (cross-block) collinearity of the *whole* assembled design is
checked next, with VIF.

# Multivariate collinearity (VIF) of the assembled design

Pairwise correlation can't see when a term is a linear combination of *several* others.
VIF can: `VIF_j = 1/(1 − R²_j)`, with `R²_j` from regressing term *j* on all others
(>5 moderate, >10 serious). It's cheap — `VIF = diag(solve(cor(X)))`, one matrix inverse,
not 150 regressions — and **n-invariant** (a function of the correlation matrix only, so a
subsample gives the same answer).

**Caveat that dominates here:** we deliberately kept the 6 raw optical bands *and* the
indices that are (near-)functions of them (NBR, NBR2, NDMI… are ratios of NIR/SWIR), plus
prev band-summaries *and* their derived index-summaries. So the **protected main set is
intentionally near-rank-deficient** — `cor(X)` is numerically singular and naïve
`diag(solve())` returns garbage (VIFs in the millions). That collinearity is **not** removed
by deleting terms; it's what the **elastic-net ridge component (α<1) is for**. Plain VIF is
the wrong tool for the *mains*.

So we use VIF only for the one question it can answer robustly: **does an interaction add
anything beyond the main-effect span?** — i.e. R² of each interaction regressed on the mains,
via a **pseudo-inverse** (well-defined even when the mains are collinear). Mains are never
dropped; only interactions are screened.

```{r design-vif}
PREV_SUMM2 <- c("median","stdDev")
SAMEBAND_K <- c("green","nir","swir1","swir2","ndvi")
CONTEXT_K  <- c("ndvi","ndwi","npv","ndfi")
FIRE_IDX   <- c("NBR","NBR2","MIRBI","NDVI"); FIRE_BAND <- c("RED","NIR","SWIR1","SWIR2")

need <- unique(c(FOCAL_KEEP, PREV_MAIN))
D  <- as.matrix(dt[, ..need]); D <- D[is.finite(rowSums(D)), ]   # VIF is n-invariant; subsample dt if memory-tight
Dc <- sweep(D, 2, colMeans(D), "-")                              # centered factors

fi  <- strsplit(keep_int, ":")                                   # kept focal interactions (centered products)
Xfi <- vapply(fi, function(ab) Dc[, ab[1]] * Dc[, ab[2]], numeric(nrow(Dc))); colnames(Xfi) <- keep_int

specs <- list()                                                  # prev x focal blocks (median+sd)
for (v in SAMEBAND_K) for (s in PREV_SUMM2) specs <- c(specs, list(c(sprintf("mb_mos_%s_%s", v, s), toupper(v))))
for (v in CONTEXT_K)  for (s in PREV_SUMM2) for (f in FIRE_IDX)  if (!(v=="ndvi" && f=="NDVI")) specs <- c(specs, list(c(sprintf("mb_mos_%s_%s", v, s), f)))
for (v in CONTEXT_K)  for (s in PREV_SUMM2) for (f in FIRE_BAND) specs <- c(specs, list(c(sprintf("mb_mos_%s_%s", v, s), f)))
Xpf <- vapply(specs, function(z) Dc[, z[1]] * Dc[, z[2]], numeric(nrow(Dc)))
colnames(Xpf) <- vapply(specs, function(z) paste(z[1], z[2], sep = ":"), "")

X <- cbind(D[, c(FOCAL_KEEP, PREV_MAIN)], Xfi, Xpf)
mains <- c(FOCAL_KEEP, PREV_MAIN); ints <- setdiff(colnames(X), mains)
C  <- cor(X); eA <- eigen(C, symmetric = TRUE)
rk <- function(M){ e <- eigen(M, symmetric = TRUE)$values; sum(e > max(e) * 1e-8) }
cat(sprintf("Design: %d terms (%d main + %d interaction)\n", ncol(X), length(mains), length(ints)))
cat(sprintf("MAIN block rank: %d/%d  |  FULL rank: %d/%d  |  recip-cond %.1e\n",
            rk(C[mains, mains]), length(mains), sum(eA$values > max(eA$values)*1e-8), ncol(X),
            min(eA$values)/max(eA$values)))

# robust 'interaction VIF' = R^2 of each interaction on the MAIN span (pseudo-inverse)
ev   <- eigen(C[mains, mains], symmetric = TRUE); tol <- max(ev$values) * 1e-8
pinv <- ev$vectors %*% ((ifelse(ev$values > tol, 1/ev$values, 0)) * t(ev$vectors))
R2   <- vapply(ints, function(j){ cc <- C[j, mains]; min(max(as.numeric(t(cc) %*% pinv %*% cc), 0), 1) }, 0.0)
vift <- data.table(term = ints, R2_on_mains = round(R2, 3),
                   vif = round(1/(1 - pmin(R2, 1 - 1e-12)), 1))[order(-R2_on_mains)]
print(vift[1:12])
drop_vif <- vift[R2_on_mains > 0.9, term]
cat(sprintf("\nVIF drop (R^2_on_mains > 0.9 / VIF > 10): %d interactions: %s\n",
            length(drop_vif), paste(drop_vif, collapse = ", ")))
```

## VIF — decision (chosen 2026-06-20)

- **MAIN block rank 43/44, full design rank 136/150** — the design is near-singular **by
  design** (bands + their derived indices, kept on purpose). This residual multivariate
  collinearity is left to the **elastic-net ridge term (α<1)**, not removed by deleting
  terms. Plain `diag(solve(cor(X)))` is meaningless here (singular) — expected, not a bug.
- **Drop the 6 interactions with R²_on_mains > 0.9** (VIF > 10) — all focal opt×idx pairs
  that survived the pairwise-0.9 prune but are *multivariately* ≥90% reproducible from the
  main effects: `NIR:NDMI, GREEN:NDSI, SWIR1:NBR2, SWIR1:NBR, SWIR2:NBR, NIR:NDVI`. This is
  exactly the redundancy pairwise correlation missed. (None exceed 0.95, so it's a soft call
  at the 0.9 line — raise the threshold to keep more.)
- Net: **focal interactions 34 → 28**, total design **150 → 144**.

# Design size — before vs after

| component | old (427-term design) | reduced |
|---|--:|--:|
| focal mains | 21 | 12 |
| prev-year mains | 40 | 32 |
| focal × focal | 210 | 28 |
| prev × focal (B4–B6) | 156 | 72 |
| **total terms (+ intercept)** | **427** | **144** |

focal × focal: 66 candidates → 34 after pairwise pruning → **28** after the VIF screen
(−6 redundant with the main span). A **~66% cut** (427 → 144), concentrated in the
interaction blocks — exactly where glmnet's coordinate descent was crawling. The remaining
collinearity (full rank 136/144) is the intentional bands+indices structure, handled by the
elastic-net ridge term.

# Next steps

- The lists are *candidates* — sanity-check each dropped index against its cluster in the
  heatmap and its separability (don't drop a high-separability index just because it
  correlates with a protected one; the protect rule already prevents that for the optical
  bands / NBR / NBR2).
- Lower `COR_HI` (0.95 → 0.90 → 0.85) to see how aggressively the raw set shrinks.
- Re-run with `SAMPLE_N <- 1e6` first for speed; confirm the final pick on the full data.
  (Separability always uses the full data regardless.)
- Once a reduced raw set is agreed, the planned follow-up is to rebuild a slimmed design
  matrix from it and inspect *its* collinearity (the interaction blocks are where the
  redundancy compounds). That goes in a new `02-`-suffixed fitting variant, not the
  current script.
```
