---
title: "SNIC seeds & candidates — data-based thresholds"
subtitle: "Burned vs unburned distributions of the period-based bpts metrics; commission/omission curves and a shallow tree to decide the seed/candidate cuts"
author: "Iván Barberá"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    embed-resources: true
    theme: cosmo
    fig-align: center
    code-fold: true
---

```{r setup, message=FALSE, warning=FALSE}
library(data.table)
library(tidyverse)
library(scales)
library(knitr)
library(rpart)
library(rpart.plot)

theme_set(theme_minimal(base_size = 11))

# One row per training point, from scripts/bp_ts_metrics_local_train.R: the
# step-03 burn-probability time-series metrics recomputed locally over each
# point's OBSERVATION PERIOD (not a calendar year), using the deployed P050
# model on the fit-cleaned observations. label: 1 = burned, 0 = unburned.
# Period-based windows dissolve the cross-year + contamination problems (see
# docs/04-snic.md), so there is no negative-set gating here.
d <- fread("../data/annual_data_resolved.csv")
lab <- c(`0` = "unburned", `1` = "burned")

# Seeds (strict, K=3) vs candidates (loose, K=2): magnitude (pmaxK) + change
# (deltaK_peak) + persistence (minforeK_peak). docs/04-snic.md §"Tune seeds…".
seed_metrics <- c("pmax3", "delta3_peak", "minfore3_peak")
cand_metrics <- c("pmax2", "delta2_peak", "minfore2_peak")
```

# What this notebook decides

Whether a simple **empirical threshold / shallow tree on the recomputed `bpts`
metrics** separates burned from unburned points well enough to define the SNIC
**seed** and **candidate** surfaces — or whether a **second (annual) model** is
warranted (docs/04-snic.md §2). Rule of thumb: if a depth-2/3 tree on 3–4
metrics reaches near-ceiling separation (§4), the second model is unnecessary; a
GEE random forest stays a fallback, not the plan.

Seeds and candidates have **opposite objectives**, so read every curve from two
ends (§3):

- **Seeds → precision** (few unburned above the cut). Seeds only need *part* of a
  scar; supervised SNIC grows the rest.
- **Candidates → recall** (few burned below the cut). Candidates define the
  footprint SNIC may grow into; missed burned area is the real cost.

**Validation.** The local metric computation is checked against the exported
`bpts` image on a single mid-year fire (`scripts/test-bp_ts_metrics_local.R`);
run that before trusting the thresholds transfer to production (§7).

# Coverage & class balance

```{r coverage}
d[, .(points = .N, burned = sum(label == 1L), unburned = sum(label == 0L)),
  by = region][order(region)] |>
  kable(caption = "Points by region.")

d[, .(points = .N, burned = sum(label == 1L)), by = .(veg_fire, veg_fire_name)][
  order(-points)][1:15] |>
  kable(caption = "Top veg_fire classes by point count.")
```

# Per-metric distributions, burned vs unburned {#sec-dist}

```{r ecdf, fig.width=9, fig.height=6}
long <- melt(d, id.vars = "label",
             measure.vars = c(seed_metrics, cand_metrics),
             variable.name = "metric", value.name = "value", na.rm = TRUE)

ggplot(long, aes(value, colour = factor(label, labels = lab))) +
  stat_ecdf(linewidth = 0.7) +
  facet_wrap(~ metric, scales = "free_x") +
  labs(colour = NULL, x = "metric value (probability scale)", y = "ECDF") +
  theme(legend.position = "top")
```

```{r violin, fig.width=9, fig.height=4}
ggplot(long[metric %in% seed_metrics],
       aes(factor(label, labels = lab), value, fill = factor(label, labels = lab))) +
  geom_violin(alpha = 0.6, scale = "width") +
  facet_wrap(~ metric, scales = "free_y") +
  labs(x = NULL, y = NULL, fill = NULL) +
  theme(legend.position = "none")
```

# Commission / omission vs threshold {#sec-curves}

For each metric, sweep the threshold: **commission** = fraction of *kept* points
that are unburned (precision failure); **omission** = fraction of *burned* points
*below* the cut (recall failure). Read the seed cut off commission, the candidate
cut off omission.

```{r co-curves, fig.width=9, fig.height=6}
co_curve <- function(dt, metric, n = 200) {
  v <- dt[[metric]]; y <- dt$label
  ok <- !is.na(v); v <- v[ok]; y <- y[ok]
  thr <- unique(quantile(v, probs = seq(0, 1, length.out = n), na.rm = TRUE))
  rbindlist(lapply(thr, function(t) data.table(
    metric = metric, threshold = t,
    commission = if (sum(v >= t)) mean(y[v >= t] == 0L) else NA_real_,
    omission   = mean(v[y == 1L] < t))))
}
curves <- rbindlist(lapply(c(seed_metrics, cand_metrics), co_curve, dt = d))

ggplot(melt(curves, id.vars = c("metric", "threshold"),
            variable.name = "error", value.name = "rate"),
       aes(threshold, rate, colour = error)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~ metric, scales = "free_x") +
  scale_y_continuous(labels = percent) +
  labs(x = "threshold (≥ kept)", y = "error rate", colour = NULL) +
  theme(legend.position = "top")
# TODO: mark chosen seed (low-commission) & candidate (low-omission) cuts, then
# transcribe to config/snic_seed_candidate_thresholds.csv.
```

# Joint separation & a shallow tree {#sec-tree}

```{r scatter, fig.width=7, fig.height=5}
ggplot(d, aes(delta3_peak, pmax3, colour = factor(label, labels = lab))) +
  geom_point(alpha = 0.25, size = 0.6) +
  labs(colour = NULL, x = "delta3_peak", y = "pmax3") +
  theme(legend.position = "top")
```

```{r cart, fig.width=8, fig.height=5}
# Shallow tree as a GUIDE (not a deployable model): which splits does the data
# suggest, and does it plateau short of clean separation?
dt <- d[!is.na(delta3_peak)]
fit <- rpart(label ~ pmax3 + delta3_peak + minfore3_peak + pmax2 +
               delta2_peak + minfore2_peak + n,
             data = dt, method = "class",
             control = rpart.control(maxdepth = 3, cp = 0.005))
rpart.plot(fit, extra = 104, box.palette = "RdYlGn", nn = TRUE)
cat(sprintf("resubstitution accuracy (depth ≤3): %.3f\n",
            mean(predict(fit, type = "class") == dt$label)))
# Near-ceiling => empirical tree suffices, no second model. Clear plateau => model.
```

# Stratification — does one threshold set hold? {#sec-strata}

Separability drifts by vegetation and observation density (grassland recovers
fast, forest slow; sparse series overlap more). If the cut moves a lot across
strata, seeds/candidates may need to be veg/region-stratified.

```{r strata-region, fig.width=9, fig.height=4}
ggplot(d, aes(delta3_peak, colour = factor(label, labels = lab))) +
  stat_ecdf(linewidth = 0.6) +
  facet_wrap(~ region, nrow = 1) +
  labs(colour = NULL, x = "delta3_peak", y = "ECDF") +
  theme(legend.position = "top")
```

```{r strata-veg, fig.width=9, fig.height=6}
top_veg <- d[, .N, by = veg_fire_name][order(-N)][1:9, veg_fire_name]
ggplot(d[veg_fire_name %in% top_veg],
       aes(delta3_peak, colour = factor(label, labels = lab))) +
  stat_ecdf(linewidth = 0.6) +
  facet_wrap(~ veg_fire_name) +
  labs(colour = NULL, x = "delta3_peak", y = "ECDF",
       subtitle = "by veg_fire (top 9 by point count)") +
  theme(legend.position = "top")
```

```{r strata-n, fig.width=9, fig.height=4}
d[, n_bin := cut(n, breaks = c(-Inf, 5, 10, 20, Inf),
                 labels = c("≤5", "6–10", "11–20", ">20"))]
ggplot(d[!is.na(n_bin)], aes(delta3_peak, colour = factor(label, labels = lab))) +
  stat_ecdf(linewidth = 0.6) +
  facet_wrap(~ n_bin, nrow = 1) +
  labs(colour = NULL, x = "delta3_peak", y = "ECDF",
       subtitle = "by annualised observation density n") +
  theme(legend.position = "top")
```

# Transfer to production & proposed thresholds {#sec-thresholds}

These metrics are **period-based**; production applies the cuts to the
**year-based** `bpts`. Confirm the single-fire hard-check
(`scripts/test-bp_ts_metrics_local.R`) passes on the magnitude metrics before
committing thresholds. Then transcribe the seed (precision-side) and candidate
(recall-side) cuts into `config/snic_seed_candidate_thresholds.csv`
(hand-editable; the SNIC step reads them) and record here the values chosen, the
reasoning, and whether a single global set held or stratification was needed.

```{r thresholds}
# TODO: fill in from the curves above, e.g.
# fread("../config/snic_seed_candidate_thresholds.csv") |> kable()
```
